---
title: "Practice drawing vector maps"
date: 2019-03-01

type: book
toc: true
draft: false
aliases: ["/notes/vector-maps-practice/"]
categories: ["dataviz", "geospatial"]

weight: 54
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
library(sf)
library(tidycensus)
library(colorspace)
library(scales)

# useful on MacOS to speed up rendering of geom_sf() objects
if (!identical(getOption("bitmapType"), "cairo") && isTRUE(capabilities()[["cairo"]])) {
  options(bitmapType = "cairo")
}

options(digits = 3)
set.seed(123)
theme_set(theme_minimal())
```

{{% callout note %}}

Run the code below in your console to download this exercise as a set of R scripts.

```r
usethis::use_course("cis-ds/visualize-spatial-ii")
```

{{% /callout %}}

## American Community Survey

The U.S. Census Bureau conducts the [American Community Survey](https://www.census.gov/programs-surveys/acs) which gathers detailed information on topics such as demographics, employment, educational attainment, etc. They make a vast portion of their data available through an [application programming interface (API)](/notes/application-program-interface/), which can be accessed intuitively through R via the [`tidycensus` package](https://walkerke.github.io/tidycensus/index.html). We previously discussed how to use this package to [obtain statistical data from the decennial census](/notes/application-program-interface/#census-data-with-tidycensus). However the Census Bureau also has detailed information on political and geographic boundaries which we can combine with their statistical measures to easily construct geospatial visualizations.

{{% callout note %}}

If you have not already, [obtain an API key](https://api.census.gov/data/key_signup.html) and [store it securely](/notes/application-program-interface/#census-data-with-tidycensus) on your computer.

{{% /callout %}}

## Exercise: Visualize income data

1. Obtain information on median household income in 2020 for Tompkins County, NY at the tract-level using the ACS. To retrieve the geographic features for each tract, set `geometry = TRUE` in your function.

{{% callout note %}}

You can use `load_variables(year = 2020, dataset = "acs5")` to retrieve the list of variables available and search to find the correct variable name.

{{% /callout %}}
    
{{< spoiler text="Click for the solution" >}}
    
```{r income-tompkins, results = "hide"}
tompkins_inc <- get_acs(
  state = "NY",
  county = "Tompkins",
  geography = "tract",
  variables = c(medincome = "B19013_001"),
  year = 2020,
  geometry = TRUE,
  output = "wide"
)
```

```{r income-tompkins-print, dependson = "income-tompkins"}
tompkins_inc
```
    
{{< /spoiler >}}

2. Draw a choropleth using the median household income data. Use a continuous color gradient to identify each tract's median household income.

{{% callout note %}}

Use the object  below to add informative labels to each plot without having to copy-and-paste.

```{r map-labels}
# create reusable labels for each plot
map_labels <- labs(
  title = "Median household income in Tompkins County, NY",
  subtitle = "In 2020",
  color = NULL,
  fill = NULL,
  caption = "Source: American Community Survey"
)
```

{{% /callout %}}

{{< spoiler text="Click for the solution" >}}

```{r income-tompkins-map, dependson = "income-tompkins"}
ggplot(data = tompkins_inc) +
  # use fill and color to avoid gray boundary lines
  geom_sf(aes(fill = medincomeE, color = medincomeE)) +
  # increase interpretability of graph
  scale_color_continuous(labels = label_dollar()) +
  scale_fill_continuous(labels = label_dollar()) +
  map_labels
```

{{< /spoiler >}}

## Exercise: Customize your maps

1. Use the [`viridis` color palette](/notes/optimal-color-palettes/#viridis) for the Tompkins County map drawn using the continuous measure.

    {{< spoiler text="Click for the solution" >}}
    
```{r viridis, dependson = "income-tompkins"}
ggplot(data = tompkins_inc) +
  # use fill and color to avoid gray boundary lines
  geom_sf(aes(fill = medincomeE, color = medincomeE)) +
  # increase interpretability of graph
  scale_fill_continuous_sequential(
    palette = "viridis",
    rev = FALSE,
    aesthetics = c("fill", "color"),
    labels = label_dollar(),
    name = NULL
  ) +
  map_labels
```
            
    {{< /spoiler >}}

2. Draw the same choropleth for Tompkins County, but convert median household income into a discrete variable with 6 levels.

    {{< spoiler text="Click for the solution" >}}

    * Using `cut_interval()`:

```{r cut-interval, dependson = "income-tompkins"}
tompkins_inc %>%
  mutate(inc_cut = cut_interval(medincomeE, n = 6)) %>%
  ggplot() +
  # use fill and color to avoid gray boundary lines
  geom_sf(aes(fill = inc_cut, color = inc_cut)) +
  # increase interpretability of graph
  scale_fill_discrete_sequential(
    palette = "viridis",
    rev = FALSE,
    aesthetics = c("fill", "color"),
    name = NULL
  ) +
  map_labels
```
        
    * Using `cut_number()`:

```{r cut-number, dependson = "income-tompkins"}
tompkins_inc %>%
  mutate(inc_cut = cut_number(medincomeE, n = 6)) %>%
  ggplot() +
  # use fill and color to avoid gray boundary lines
  geom_sf(aes(fill = inc_cut, color = inc_cut)) +
  # increase interpretability of graph
  scale_fill_discrete_sequential(
    palette = "viridis",
    rev = FALSE,
    aesthetics = c("fill", "color"),
    name = NULL
  ) +
  map_labels
```

    - Using `binned_scale()`

```{r}
# default breaks
ggplot(data = tompkins_inc) +
  geom_sf(mapping = aes(fill = medincomeE, color = medincomeE)) +
  scale_fill_binned_sequential(
    palette = "viridis",
    rev = FALSE,
    aesthetics = c("fill", "color"),
    labels = label_dollar()
  ) +
  # increase interpretability of graph
  map_labels

# quintiles
ggplot(data = tompkins_inc) +
  geom_sf(mapping = aes(fill = medincomeE, color = medincomeE)) +
  scale_fill_binned_sequential(
    palette = "viridis",
    rev = FALSE,
    aesthetics = c("fill", "color"),
    n.breaks = 4, nice.breaks = FALSE,
    labels = label_dollar()
  ) +
  # increase interpretability of graph
  map_labels
```

    {{< /spoiler >}}

### Session Info

```{r child = here::here("R", "_session-info.Rmd")}
```
