---
date: "2018-09-09T00:00:00-05:00"
draft: false
weight: 20
title: "Install Git"
toc: true
type: book
aliases: ["/git01.html", "/setup/git/"]
---

```{r}
#| label = "setup",
#| include = FALSE
knitr::opts_chunk$set(cache = TRUE)
```

## Git Installation: Windows

Install [Git for Windows](https://git-for-windows.github.io/), also known as `msysgit` or "Git Bash", to get Git in addition to some other useful tools, such as the Bash shell. Yes, all those names are totally confusing, but you might encounter them elsewhere and I want you to be well-informed.

This method of installing Git for Windows leaves the Git executable in a conventional location, which will help you and other programs, e.g. RStudio, find it and use it. This also supports a transition to more expert use, because the "Git Bash" shell will be useful as you venture outside of R/RStudio.

* When asked about "Adjusting your PATH environment", make sure to select "Git from the command line and also from 3rd-party software". Otherwise, we believe it is good to accept the defaults.
* Note that RStudio for Windows prefers for Git to be installed below `C:/Program Files` and this appears to be the default. This implies, for example, that the Git executable on my Windows system is found at `C:/Program Files/Git/bin/git.exe`. Unless you have specific reasons to otherwise, follow this convention.

{{% callout note %}}

If you follow these instructions and install RStudio version 1.1 or higher, you can use the shell (also called the **terminal**) directly from within RStudio. See [this article for more information on how to use it and setup RStudio to use Git Bash, rather than the Command Prompt](https://support.rstudio.com/hc/en-us/articles/115010737148-Using-the-RStudio-Terminal).

{{% /callout %}}

## Git Installation: Mac OS X

Mac OS X already includes the shell, so all you need to do is install Git.

## Option 1 (highly recommended)

Install the Xcode command line tools (**not all of Xcode**), which includes Git.

Go to the shell and enter one of these commands to elicit an offer to install developer command line tools:

``` bash
git --version
git config
```

Accept the offer! Click on "Install".

Here's another way to request this installation, more directly:

``` bash
xcode-select --install
```

We just happen to find this Git-based trigger apropos.

Note also that, after upgrading macOS, you might need to re-do the above and/or re-agree to the Xcode license agreement. We have seen this cause the RStudio Git pane to disappear on a system where it was previously working. Use commands like those above to tickle Xcode into prompting you for what it needs, then restart RStudio.

## Option 2 (recommended)

Install Git from here: <http://git-scm.com/downloads>.

* This arguably sets you up the best for the future. It will certainly get you the latest version of Git of all approaches described here.
* The GitHub home for the macOS installer is here: <https://github.com/timcharper/git_osx_installer>.
    * At that link, you can find more info if something goes wrong or you are working on an old version of macOS.

## Git Installation: Linux

If Git is not already available on your machine you can try to install it via your distro's package manager.

Debian/Ubuntu

`sudo apt-get install git`

Fedora/Redhat Linux

`sudo yum install git`

## Acknowledgments

```{r}
#| child = here::here("R", "_ack_stat545.Rmd")
```
```{r}
#| child = here::here("R", "_ack_swcgit.Rmd")
```
* ["Chapter 6: Install Git" from Happy Git and GitHub for the useR](https://happygitwithr.com/install-git.html)
