---
# About/Biography widget.
widget: "blank"
headless: true
active: true
date: 2018-12-13T00:00:00

# Order that this section will appear in.
weight: 10

design:
  columns: "1"

output:
  blogdown::html_page:
    toc: false
    number_sections: false
---

# MACS 30500^[aka ENST 20550/MACS 20500/CHDV 30511/MAPS 30500/PLSC 30235/SOCI 20278/SOCI 40176] - Computing for the Social Sciences

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
library(sf)
library(tidycensus)
library(viridis)

# useful on MacOS to speed up rendering of geom_sf() objects
if (!identical(getOption("bitmapType"), "cairo") && isTRUE(capabilities()[["cairo"]])) {
  options(bitmapType = "cairo")
}

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal())
```

<div class="row">
  <div class="col-sm-7 col-xs-12">

  {{% alert note %}}
  
  All class meetings and office hours will be held on Zoom. Links to these meetings can be found on the [Canvas site](https://canvas.uchicago.edu/courses/27637) (UChicago authentication required).
  
  {{% /alert %}}

  * Instructor: [Benjamin Soltoff](http://www.bensoltoff.com)
  * Teaching Assistants
      * Dominique Janvier 
      * Deblina Mukherjee
      * Neeraj Sharma
  * Meeting day/time: MW 1:30-2:50pm
  * Online course discussion: [GitHub discussion repo](https://github.com/uc-cfss/Discussion)
  * Open lab session - W 3-4:20pm
  * Office hours:
      * Benjamin: M 10-12pm
      * Dominique: F 2-4pm
      * Deblina: F 11-1pm
      * Neeraj: Th 11-1pm
  * Prerequisites: None
  * Requirements: Bring your own laptop

  </div>
  
  <div class="col-sm-5 col-xs-12">
  
```{r income-cook, results = "hide"}
cook_inc <- get_acs(
  state = "IL",
  county = "Cook",
  geography = "tract",
  variables = c(medincome = "B19013_001"),
  year = 2017,
  geometry = TRUE
)
```

```{r income-cook-map-viridis, dependson = "income-cook", fig.width = 5, fig.height = 5, fig.cap = "Example of student work using `sf`, `tidycensus`, and `ggplot2`"}
ggplot(data = cook_inc) +
  # use fill and color to avoid gray boundary lines
  geom_sf(aes(fill = estimate, color = estimate)) +
  # increase interpretability of graph
  scale_color_viridis(labels = scales::dollar) +
  scale_fill_viridis(labels = scales::dollar) +
  labs(
    title = "Median household income in Cook County, IL",
    subtitle = "In 2017",
    color = NULL,
    fill = NULL,
    caption = "Source: American Community Survey"
  )
```

  </div>

</div>
  
## Course Description

This is an applied course for social scientists with little-to-no programming experience who wish to harness growing digital and computational resources. The focus of the course is on generating **reproducible research** through the use of programming languages and version control software. Major emphasis is placed on a pragmatic understanding of core principles of programming and packaged implementations of methods. Students will leave the course with basic computational skills implemented through many computational methods and approaches to social science; while students will not become expert programmers, they will gain the knowledge of how to adapt and expand these skills as they are presented with new questions, methods, and data.

## Course Objectives

By the end of the course, students will:

* Construct and execute basic programs in R using elementary programming techniques and [`tidyverse`](http://tidyverse.org/) packages (e.g. loops, conditional statements, user-defined functions)
* Apply stylistic principles of coding to generate reusable, interpretable code
* Debug programs for errors
* Identify and use external libraries to expand on base functions
* Implement best practices for reproducible research
* Implement statistical learning algorithms
* Visualize information and data using appropriate graphical techniques
* Import data from files or the internet
* Munge raw data into a tidy format
* Scrape websites to collect data for analysis
* Parse and analyze text documents
* Construct geospatial visualizations
* Create interactive web pages using [`flexdashboard`](http://rmarkdown.rstudio.com/flexdashboard/) and [Shiny](http://shiny.rstudio.com/)
