---
title: "Data storage types"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Objectives

* Review the major types of vectors
* Demonstrate how to subset vectors
* Demonstrate vector recycling
* Define *lists*

```{r packages}
library(tidyverse)
set.seed(1234)
```

# Vectors in R

So far the only type of data object in R you have encountered is a `data.frame` (or the `tidyverse` variant `tibble`). At its core though, the primary method of data storage in R is the **vector**. [We discussed them briefly a couple weeks ago](block005_parsing_vectors.html) merely to note that a data frame is composed from vectors and that there are a few different types of vectors: logical, numeric, and character. But now we want to understand more precisely how these data objects are structured and related to one another.

## Types of vectors

![Figure 20.1 from [*R for Data Science*](http://r4ds.had.co.nz/vectors.html)](http://r4ds.had.co.nz/diagrams/data-structures-overview.png)

There are two categories of vectors:

1. **Atomic vectors** - these are the types previously covered, including logical, integer, double, and character.
1. **Lists** - there are new and we will cover them later in this block. Lists are distinct from atomic vectors because lists can contain other lists.

Atomic vectors are **homogenous** - that is, all elements of the vector must be the same type. Lists can be **hetergenous** and contain multiple types of elements. `NULL` is the counterpart to `NA`. While `NA` represents the absence of a value, `NULL` represents the absence of a vector.

## Atomic vectors

### Logical vectors

**Logical vectors** take on one of three possible values:

* `TRUE`
* `FALSE`
* `NA` (missing value)

```{r}
parse_logical(c(TRUE, TRUE, FALSE, TRUE, NA))
```

> Whenever you filter a data frame, R is (in the background) creating a vector of `TRUE` and `FALSE` - whenever the condition is `TRUE`, keep the row, otherwise exclude it.

### Numeric vectors

**Numeric vectors** contain numbers (duh!). They can be stored as **integers** (whole numbers) or **doubles** (numbers with decimal points). In practice, you rarely need to concern yourself with this difference, but just know that they are different but related things.

```{r}
parse_integer(c(1, 5, 3, 4, 12423))
parse_double(c(4.2, 4, 6, 53.2))
```

> Doubles can store both whole numbers and numbers with decimal points.

### Character vectors

**Character vectors** contain **strings**, which are typically text but could also be dates or any other combination of characters.

```{r}
parse_character(c("Goodnight Moon", "Runaway Bunny", "Big Red Barn"))
```

## Using atomic vectors

Be sure to read ["Using atomic vectors"](http://r4ds.had.co.nz/vectors.html#using-atomic-vectors) for more detail on how to use and interact with atomic vectors. I have no desire to rehash everything Hadley already wrote, but here are a couple things about atomic vectors I want to reemphasize.

### Scalars

**Scalars** are a single number; **vectors** are a set of multiple values. In R, scalars are merely a vector of length 1. So when you try to perform arithmetic or other types of functions on a vector, it will **recycle** the scalar value.

```{r}
(x <- sample(10))
x + c(100, 100, 100, 100, 100, 100, 100, 100, 100, 100)
x + 100
```

This is why you don't need to write a `for` loop when performing these basic operations - R automatically converts it for you.

Sometimes this isn't so great, because R will also recycle vectors if the lengths are not equal:

```{r}
# create a sequence of numbers between 1 and 10
1:2
1:10

# add together two sequences of numbers
1:10 + 1:2
```

Did you really mean to recycle `1:2` five times, or was this actually an error? `tidyverse` functions will only allow you to implicitly recycle scalars, otherwise it will throw an error and you'll have to manually recycle shorter vectors.

### Subsetting

To filter a vector, we cannot use `filter()` because that only works for filtering rows in a `tibble`. `[` is the subsetting function for vectors. It is used like `x[a]`.

#### Subset with a numeric vector containing integers

```{r}
x <- c("one", "two", "three", "four", "five")
```

Subset with positive integers keeps the corresponding elements:

```{r}
x[c(3, 2, 5)]
```

Negative values drop the corresponding elements:

```{r}
x[c(-1, -3, -5)]
```

You cannot mix positive and negative values:

```{r, error = TRUE}
x[c(-1, 1)]
```

#### Subset with a logical vector

Subsetting with a logical vector keeps all values corresponding to a `TRUE` value.

```{r}
x <- c(10, 3, NA, 5, 8, 1, NA)

# All non-missing values of x
x[!is.na(x)]

# All even (or missing!) values of x
x[x %% 2 == 0]
```

### Exercises: subset the vector

```{r}
(x <- 1:10)
```

Create the sequence above in your R session. Write commands to subset the vector in the following ways:

1. Keep the first through fourth elements, plus the seventh element.

    <details> 
      <summary>Click for the solution</summary>
      <p>

    ```{r}
    x[c(1, 2, 3, 4, 7)]
    
    # use a sequence shortcut
    x[c(1:4, 7)]
    ```
    
      </p>
    </details>

1. Keep the first through eighth elements, plus the tenth element.

    <details> 
      <summary>Click for the solution</summary>
      <p>

    ```{r}
    # long way
    x[c(1, 2, 3, 4, 5, 6, 7, 8, 10)]
    
    # sequence shortcut
    x[c(1:8, 10)]
    
    # negative indexing
    x[c(-9)]
    ```
    
      </p>
    </details>

1. Keep all elements with values greater than five.

    <details> 
      <summary>Click for the solution</summary>
      <p>

    ```{r}
    x[x > 5]
    ```
    
      </p>
    </details>

1. Keep all elements evenly divisible by three.

    <details> 
      <summary>Click for the solution</summary>
      <p>

    ```{r}
    x[x %% 3 == 0]
    ```
    
      </p>
    </details>

# Lists

**Lists** are an entirely different type of vector.

```{r}
x <- list(1, 2, 3)
x
```

Use `str()` to view the **structure** of the list.

```{r}
str(x)

x_named <- list(a = 1, b = 2, c = 3)
str(x_named)
```

Unlike the other atomic vectors, lists are recursive. This means they can:

1. Store a mix of objects.
    ```{r}
    y <- list("a", 1L, 1.5, TRUE)
    str(y)
    ```
    
1. Contain other lists.
    ```{r}
    z <- list(list(1, 2), list(3, 4))
    str(z)
    ```
    
    It isn't immediately apparent why you would want to do this, but in later units we will discover the value of lists as many packages for R store non-tidy data as lists.

You've already worked with lists without even knowing it. **Data frames and `tibble`s are a type of a list.** Notice that you can store a data frame with a mix of column types.

```{r}
str(diamonds)
```

## How to subset lists

Sometimes lists (especially deeply-nested lists) can be confusing to view and manipulate. Take the example from [R for Data Science](http://r4ds.had.co.nz/vectors.html#subsetting-1):

```{r}
a <- list(a = 1:3, b = "a string", c = pi, d = list(-1, -5))
str(a)
```

* `[` extracts a sub-list. The result will always be a list.

    ```{r}
    str(a[1:2])
    str(a[4])
    ```
* `[[`` extracts a single component from a list and removes a level of hierarchy.

    ```{r}
    str(a[[1]])
    str(a[[4]])
    ```

* `$` can be used to extract named elements of a list.

    ```{r}
    a$a
    a[["a"]]
    ```

![Figure 20.2 from [R for Data Science](http://r4ds.had.co.nz/vectors.html#fig:lists-subsetting)](http://r4ds.had.co.nz/diagrams/lists-subsetting.png)

> Still confused about list subsetting? [Review the pepper shaker.](http://r4ds.had.co.nz/vectors.html#lists-of-condiments)

## Exercises: subset a list

```{r}
a <- list(a = 1:3, b = "a string", c = pi, d = list(-1, -5))
str(a)
```

Create the list above in your R session. Write commands to subset the list in the following ways:

1. Subset `a`. The result should be an atomic vector.

    <details> 
      <summary>Click for the solution</summary>
      <p>

    ```{r}
    # use the index value
    a[[1]]
    
    # use the element name
    a$a
    a[["a"]]
    ```
    
      </p>
    </details>

1. Subset `pi`. The results should still be a new list.

    <details> 
      <summary>Click for the solution</summary>
      <p>

    ```{r}
    # correct method
    str(a["c"])
    
    # incorrect method to produce another list
    # the result is a scalar
    str(a$c)
    ```
    
      </p>
    </details>

1. Subset the first and third elements from `a`.

    <details> 
      <summary>Click for the solution</summary>
      <p>

    ```{r}
    a[c(1, 3)]
    a[c("a", "c")]
    ```
    
      </p>
    </details>

# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```




