---
title: "Iteration"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Objectives

* Demonstrate how to write an iterative operation using a `for` loop
* Practice writing a `for` loop
* Demonstrate how to write an iterative operation using a `map` function
* Practice writing a `map` function

## Prerequisites

* [Chapter 21: "Iteration" in *R for Data Science*](http://r4ds.had.co.nz/iteration.html)

```{r packages, cache = FALSE}
library(tidyverse)
set.seed(1234)
```

# Writing `for` loops

Functions are one method of reducing duplication in your code. Another tool for reducing duplication is **iteration**, which lets you do the same thing to multiple inputs.

## Example for loop

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

Let's say we want to compute the median for each column.

```{r}
median(df$a)
median(df$b)
median(df$c)
median(df$d)
```

Boo! We've copied-and-pasted `median()` three times. We don't want to do that. Instead, we can use a `for` loop:

```{r}
output <- vector(mode = "double", length = ncol(df))  # 1. output
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}
output
```

Let's review the three components of every `for` loop:

1. **Output**

    `output <- vector("double", length(x))`
    
    Before you start a loop, you need to create an empty vector to store the output of the loop. Preallocating space for your output is much more efficient than creating space as you move through the loop. The `vector()` function allows you to create an empty vector of any type. The first argument `mode` defines the type of vector ("logical", "integer", "double", "character", etc.) and the second argument `length` defines the length of the vector.
    
    Numeric vectors are initialized to `0`, logical vectors are initialized to `FALSE`, character vectors are initialized to `""`, and list elements to `NULL`.
    
    ```{r}
    vector(mode = "double", length = ncol(df))
    vector(mode = "logical", length = ncol(df))
    vector(mode = "character", length = ncol(df))
    vector(mode = "list", length = ncol(df))
    ```

1. **Sequence**

    `i in seq_along(df)`
    
    This component determines what to loop over. Each iteration through the `for` loop, a new value will be assigned to `i` based on the the defined sequence. Here, the sequence is `seq_along(df)` which creates a numeric vector for a sequence of numbers beginning at 1 and continuing until it reaches the length of `df` (the length here is the number of columns in `df`).
    
    ```{r}
    seq_along(df)
    ```
    
1. **Body**

    `output[[i]] <- median(df[[i]])`
    
    This is the code that actually performs the desired calculations. It runs multiple times for every value of `i`. We use `[[` notation to reference each column of `df` and store it in the appropriate element in `output`.

## Why we preallocate space for the output

If you don't preallocate space for the output, each time the `for` loop iterates, it makes a copy of the output and appends the new value at the end. Copying data takes time and memory. If the output is preallocated space, the loop simply fills in the slots with the correct values.

```{r}
x <- rnorm(1000, mean = 0, sd = 1)
str(x)

# load microbenchmark library to time code
library(microbenchmark)

microbenchmark(
  # don't preallocate
  nopre = {
    output <- vector("numeric", 0)
    
    for (i in seq_along(x)) {
      output <- c(output, x[[i]] + 1)
    }
  },
  # preallocate
  pre = {
    output <- vector("numeric", length(x))
    
    for (i in seq_along(x)) {
      output[[i]] <- x[[i]] + 1
    }
  })
```

Here, preallocating space for the vector cuts the computation time by a factor of 3.

## Exercise: write a `for` loop

### Mean of columns in `mtcars`

Write a `for` loop that calculates the arithmetic mean for every column in `mtcars`.

```{r}
as_tibble(mtcars)
```

Before you write the `for` loop, identify the three components you need:

* Output
* Sequence
* Body

<details> 
  <summary>Click for the solution</summary>
  <p>

* Output - a numeric vector of length `r ncol(mtcars)`
* Sequence - `i in seq_along(mtcars)`
* Body - calculate the `mean()` of the $i$th column, store the new value as the $i$th element of the vector `output`

```{r}
output <- vector("integer", ncol(mtcars))

for(i in seq_along(mtcars)){
  output[[i]] <- mean(mtcars[[i]], na.rm = TRUE)
}

output
```
    
  </p>
</details>

### Maximum value in each column of `diamonds`

Write a `for` loop that calculates the maximum value in each column of `diamonds`.

```{r}
diamonds
```

Before you write the `for` loop, identify the three components you need:

* Output
* Sequence
* Body

<details> 
  <summary>Click for the solution</summary>
  <p>

* Output - a vector of length `r ncol(ggplot2::diamonds)`
* Sequence - `i in seq_along(diamonds)`
* Body - get the maximum value of the $i$th column of the data frame `diamonds`, store the new value as the $i$th element of the list `output`

```{r}
output <- vector("numeric", ncol(diamonds))

for(i in seq_along(diamonds)){
  output[i] <- max(diamonds[[i]])
}

output
```

To preserve the name attributes from `diamonds`, use the `names()` function to extract the names of each column in `diamonds` and apply them as the names to the `output` vector:

```{r}
# get the names of the columns in diamonds
names(diamonds)

# assign the names of the diamonds columns to output
names(output) <- names(diamonds)
output
```

  </p>
</details>

# Map functions

You will frequently need to iterate over vectors or data frames, perform an operation on each element, and save the results somewhere. `for` loops are not the devil. At first, they may seem more intuitive to use because you are explicitly identifying each component of the process. However the downside is that they focus on a lot of non-essential stuff. You have to track the value on which you are iterating, you need to explicitly create a vector to store the output, you have to assign the output of each iteration to the appropriate element in the output vector, etc. `tidyverse` is all about focusing on **verbs**, not **nouns**. That is, focus on the operation being performed (e.g. `mean()`, `median()`, `max()`), not all the extra code needed to make the operation work.

The `purrr` library provides a family of functions that mirrors `for` loops. They:

* Loop over a vector
* Do something to each element
* Save the results

There is one function for each type of output:

* `map()` makes a list.
* `map_lgl()` makes a logical vector.
* `map_int()` makes an integer vector.
* `map_dbl()` makes a double vector.
* `map_chr()` makes a character vector.

```{r}
map_dbl(df, mean)
map_dbl(df, median)
map_dbl(df, sd)
```

Just like all of our functions in the `tidyverse`, the first argument is always the data object, and the second argument is the function to be applied. Additional arguments for the function to be applied can be specified like this:

```{r}
map_dbl(df, mean, na.rm = TRUE)
```

Or data can be piped:

```{r}
df %>%
  map_dbl(mean, na.rm = TRUE)
```

## Exercise: rewrite our `for` loops using a `map()` function

### Mean of columns in `mtcars`

Write a `map()` function that calculates the arithmetic mean for every column in `mtcars`.

```{r}
as_tibble(mtcars)
```

<details> 
  <summary>Click for the solution</summary>
  <p>

```{r}
map_dbl(mtcars, mean)
```
    
  </p>
</details>

### Maximum value in each column of `diamonds`

Write a `map()` function that calculates the maximum value in each column of `diamonds`.

```{r}
diamonds
```

<details> 
  <summary>Click for the solution</summary>
  <p>

```{r}
map_dbl(diamonds, max)
```

  </p>
</details>


# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```




