---
title: "Visualizations and the grammar of graphics"
output:
  html_document:
    toc: true
    toc_float: true
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Objectives

* Identify the importance of graphics in communicating information
* Define the layered grammar of graphics
* Practice generating layered graphics using [`ggplot2`](https://github.com/hadley/ggplot2)

```{r packages, message = FALSE}
library(tidyverse)
library(knitr)
library(broom)
library(stringr)
```

# Identify the importance of graphics in communicating information

Research methods classes in graduate school generally teach important skills such as probability and statistical theory, regression, analysis of variance (ANOVA), maximum likelihood estimation (MLE), etc. While these are important methods for analyzing data and assessing research questions, sometimes drawing a picture (aka **visualization**) can be more precise than conventional statistical computations.^[Example drawn from [*The Visual Display of Quantitative Information* by Edward Tufte](https://www.edwardtufte.com/tufte/books_vdqi)]

Consider the following data sets:

```{r misleading, echo = FALSE}
# input data
dat1 <- tibble(x = c(10, 8, 13, 9, 11, 14, 6, 4, 12, 7, 5),
               y = c(8.04, 6.95, 7.58, 8.81, 8.33, 9.96, 7.24, 4.26, 10.84, 4.82, 5.68))
dat2 <- tibble(x = c(10, 8, 13, 9, 11, 14, 6, 4, 12, 7, 5),
               y = c(9.14, 8.14, 8.74, 8.77, 9.26, 8.10, 6.13, 3.1, 9.13, 7.26, 4.74))
dat3 <- tibble(x = c(10, 8, 13, 9, 11, 14, 6, 4, 12, 7, 5),
               y = c(7.46, 6.77, 12.74, 7.11, 7.81, 8.84, 6.08, 5.39, 8.15, 6.42, 5.73))
dat4 <- tibble(x = c(8, 8, 8, 8, 8, 8, 8, 19, 8, 8, 8),
               y = c(6.58, 5.76, 7.71, 8.84, 8.47, 7.04, 5.25, 12.5, 5.56, 7.91, 6.89))

# combine data
dat <- bind_rows(dat1, dat2, dat3, dat4, .id = "id")
```

```{r misleading_print, dependson = 'misleading', echo = FALSE, results='asis'}
dat_print <- dat %>%
  group_by(id) %>%
  nest(x, y)

for(i in seq_len(nrow(dat_print))){
  print(kable(dat_print[i,]$data,
              caption = str_c("Dataset", i, sep = " ")))
}
```

What are the corresponding relationships between $X$ and $Y$? Using traditional metrics, the relationships appear identical across the samples:

```{r mislead_stats, dependson = 'misleading', echo = FALSE}
# get stats
dat %>%
  group_by(id) %>%
  summarize(n = n(),
           mean_x = mean(x),
           mean_y = mean(y),
           cor = cor(x, y)) %>%
  kable(col.names = c("Dataset ID", "$N$", "$\\bar{X}$", "$\\bar{Y}$", "$R^2$"))
```

If we estimated linear regression models for each dataset, we would obtain virtually identical coefficients (again suggesting the relationships are identical):

```{r mislead_model, dependson = 'misleading', results = 'asis', echo = FALSE}
# estimate linear model
dat_model <- dat %>%
  split(.$id) %>%
  map(~ lm(y ~ x, data = .x)) %>%
  map(tidy)

for(i in seq_along(dat_model)){
  print(kable(dat_model[i],
              caption = str_c("Dataset", i, sep = " "),
              col.names = c("Term", "Estimate", "Standard Error", "$T$-statistic", "p-value")))
}
```

But what happens if we draw a picture?

```{r mislead_plot, dependson = 'misleading', echo = FALSE}
# now draw a graph
dat %>%
  mutate(id = str_c("Dataset", id, sep = " ")) %>%
  ggplot(aes(x, y)) +
  facet_wrap(~id) +
  geom_point() +
  theme_bw()
```

A good picture tells the reader much more than any table or text can provide.

# The layered grammar of graphics

Google defines a **grammar** as "the whole system and structure of a language or of languages in general, usually taken as consisting of syntax and morphology (including inflections) and sometimes also phonology and semantics".^[[Google](https://www.google.com/search?q=grammar)] Others consider a grammar to be "the fundamental principles or rules of an art or science".[^layer] Applied to visualizations, a **grammar of graphics** is a grammar used to describe and create a wide range of statistical graphics.^[[Wilkinson, Leland. (2005). *The Grammar of Graphics*. (UChicago authentication required)](http://link.springer.com.proxy.uchicago.edu/book/10.1007%2F0-387-28695-0)]

The **layered grammar of graphics** approach is implemented in [`ggplot2`](https://cran.r-project.org/web/packages/ggplot2/index.html), a widely used graphics library for R. All graphics in this library are built using a layered approach, building layers up to create the final graphic.

# Components of the layered grammar of graphics

* Layer
    * Data
    * Mapping
    * Statistical transformation (stat)
    * Geometric object (geom)
    * Position adjustment (position)
* Scale
* Coordinate system (coord)
* Faceting (facet)
* Defaults
    * Data
    * Mapping

## Layer

**Layers** are used to create the objects on a plot. They are defined by five basic parts:

1. Data
1. Mapping
1. Statistical transformation (stat)
1. Geometric object (geom)
1. Position adjustment (position)

Layers are typically related to one another and share many common features. For instance, multiple layers can be built using the same underlying data. An example would be a scattterplot overlayed with a smoothed regression line to summarize the relationship between the variables:

```{r layers, echo = FALSE}
set.seed(1234)

data_frame(x = runif(50),
           y = x + rnorm(50, 0, .2)) %>%
  ggplot(aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

### Data and mapping

**Data** defines the source of the information to be visualized, but is independent from the other elements. So a layered graphic can be built which can utilize different data sources while keeping the other components the same. Here is a portion of a dataset contained in the `ggplot2` package: `mpg`.

```{r mpg}
head(mpg) %>%
  kable(caption = "Dataset of automobiles")
```

**Mapping** defines how the variables are applied to the graphic. So if we were graphing information from `mpg`, we might map a car's engine displacement to the $x$ position and highway mileage to the $y$ position.

```{r mapping}
mpg %>%
  select(displ, hwy) %>%
  rename(x = displ,
         y = hwy)
```

### Statistical transformation

A **statistical transformation** (*stat*) transforms the data, generally by summarizing the information. For instance, in a bar graph you typically are not trying to graph the raw data because this doesn't make any inherent sense. Instead, you might summarize the data by graphing the total number of observations within a set of categories. Or if you have a dataset with many observations, you might transform the data into a smoothing line which summarizes the overall pattern of the relationship between variables.

A stat takes a dataset as input and returns a dataset as output, and so a stat can add new variables to the original dataset. So instead of graphing this data in its raw form:

```{r stat_raw}
mpg %>%
  select(cyl)
```

You would transform it to:

```{r stat_transform}
mpg %>%
  count(cyl)
```

> Sometimes you don't need to make a statistical transformation. For example, in a scatterplot you use the raw values for the $x$ and $y$ variables to map onto the graph. In these situations, the statistical transformation is an *identity* transformation - the stat simply passes in the original dataset and exports the exact same dataset.

### Geometric objects

**Geometric objects** (*geoms*) control the type of plot you create. Geoms are classified by their dimensionality:

* 0 dimensions - point, text
* 1 dimension - path, line
* 2 dimensions - polygon, interval

Each geom can only display certain aesthetics. For example, a point geom has position, color, shape, and size aesthetics.

```{r geom_point}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point() +
  ggtitle("A point geom with position and color aesthetics")
```

A bar geom has position, height, width, and fill color.

```{r geom_bar}
ggplot(mpg, aes(cyl)) +
  geom_bar() +
  ggtitle("A bar geom with position and height aesthetics")
```

### Position adjustment

Sometimes with dense data we need to adjust the position of elements on the plot, otherwise data points might obscure one another. Bar plots frequently stack or dodge the bars to avoid overlap:

```{r position_dodge}
count(mpg, class, cyl) %>%
  ggplot(aes(cyl, n, fill = class)) +
  geom_bar(stat = "identity") +
  ggtitle("A stacked bar chart")

count(mpg, class, cyl) %>%
  ggplot(aes(cyl, n, fill = class)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("A dodged bar chart")
```

Sometimes scatterplots with few unique $x$ and $y$ values are **jittered** (random noise is added) to reduce overplotting.

```{r position}
ggplot(mpg, aes(cyl, hwy)) +
  geom_point() +
  ggtitle("A point geom with obscured data points")

ggplot(mpg, aes(cyl, hwy)) +
  geom_jitter() +
  ggtitle("A point geom with jittered data points")
```

## Scale

A **scale** controls how data is mapped to aesthetic attributes, so we need one scale for every aesthetic property employed in a layer. For example, this graph defines a scale for color:

```{r scale_color}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point() +
  guides(color = guide_legend(override.aes = list(size = 4)))
```

Note that the scale is consistent - every point for a compact car is drawn in tan, whereas SUVs are drawn in pink. The scale can be changed to use a different color palette:

```{r scale_color_palette}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  guides(color = guide_legend(override.aes = list(size = 4)))
```

Now we are using a different palette, but the scale is still consistent: all compact cars utilize the same color, whereas SUVs use a different color **but each SUV uses the same color**.

## Coordinate system

A **coordinate system** (*coord*) maps the position of objects onto the plane of the plot, and control how the axes and grid lines are drawn. Plots typically use two coordinates ($x, y$), but could use any number of coordinates. Most plots are drawn using the [**Cartesian coordinate system**](https://en.wikipedia.org/wiki/Cartesian_coordinate_system):

```{r coord_cart}
x1 <- c(1, 10)
y1 <- c(1, 5)
p <- qplot(x1, y1, geom = "blank", xlab = NULL, ylab = NULL) +
  theme_bw()
p 
```

This system requires a fixed and equal spacing between values on the axes. That is, the graph draws the same distance between 1 and 2 as it does between 5 and 6. The graph could be drawn using a [**semi-log coordinate system**](https://en.wikipedia.org/wiki/Semi-log_plot) which logarithmically compresses the distance on an axis:

```{r coord_semi_log}
p + coord_trans(y = "log10")
```

Or could even be drawn using [**polar coordinates**](https://en.wikipedia.org/wiki/Polar_coordinate_system):

```{r coord_polar}
p + coord_polar()
```

## Faceting

**Faceting** can be used to split the data up into subsets of the entire dataset. This is a powerful tool when investigating whether patterns are the same or different across conditions, and allows the subsets to be visualized on the same plot (known as **conditioned** or **trellis** plots). The faceting specification describes which variables should be used to split up the data, and how they should be arranged.

```{r facet}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~class)
```

## Defaults

Rather than explicitly declaring each component of a layered graphic (which will use more code and introduces opportunities for errors), we can establish intelligent defaults for specific geoms and scales. For instance, whenever we want to use a bar geom, we can default to using a stat that counts the number of observations in each group of our variable in the $x$ position.

```{r default}
ggplot() +
  layer(
    data = mpg, mapping = aes(x = displ, y = hwy),
    geom = "point", stat = "identity", position = "identity"
  ) +
  scale_x_continuous() +
  scale_y_continuous() +
  coord_cartesian()
```

The above code:

* Creates a new plot object (`ggplot`)
* Adds a layer (`layer`)
    * Specifies the data (`mpg`)
    * Maps engine displacement to the $x$ position and highway mileage to the $y$ position (`mapping`)
    * Uses the point geometric transformation (`geom = "point"`)
    * Implements an identity transformation and position (`stat = "identity"` and `position = "identity"`)
* Establishes two continuous position scales (`scale_x_continuous` and `scale_y_continuous`)
* Declares a cartesian coordinate system (`coord_cartesian`)

How can we simplify this using intelligent defaults?

1. We only need to specify one geom and stat, since each geom has a default stat.
1. Cartesian coordinate systems are most commonly used, so it is the default.
1. Default scales can be added based on the aesthetic and type of variables.
    * Continuous values are transformed with a linear scaling.
    * Discrete values are mapped to integers.
    * Scales for aesthetics such as color, fill, and size can also be intelligently defaulted.

Using these defaults, we can rewrite the above code as:

```{r default2}
ggplot() +
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy))
```

Because multiple layers can use the same components (data, mapping, etc.), we can specify that information in the `ggplot` function rather than in the `layer` function:

```{r default3}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

And as we will learn, function arguments in R use specific ordering, so we can omit the explicit call to `data` and `mapping`:

```{r default4}
ggplot(mpg, aes(displ, hwy)) +
  geom_point()
```

With this specification, it is easy to build the graphic up with additional layers, without modifying the original code:

```{r default5}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth()
```

Because we called `aes(x = displ, y = hwy)` within the `ggplot` function, it is automatically passed along to both `geom_point()` and `geom_smooth()`.

# How to build a complicated, layered graphic

## Charles Minard's map of Napoleon's disastrous Russian campaign of 1812

![["Carte figurative des pertes successives en hommes de l'Armee Français dans la campagne de Russe 1812–1813" by Charles Joseph Minard](https://en.wikipedia.org/wiki/Charles_Joseph_Minard)](https://upload.wikimedia.org/wikipedia/commons/2/29/Minard.png)

The graphic is notable for its representation in two dimensions of six types of data: the number of Napoleon's troops; distance; temperature; the latitude and longitude; direction of travel; and location relative to specific dates.^[This exercise is drawn from [Wickham, Hadley. (2010) "A Layered Grammar of Graphics". *Journal of Computational and Graphical Statistics*, 19(1).](http://www.jstor.org.proxy.uchicago.edu/stable/25651297)]

## Building Minard's map in R

```{r minard_data}
# get data on troop movements and city names
troops <- read_table("data/minard-troops.txt")
cities <- read_table("data/minard-cities.txt")
troops
cities
```

### Grammar of Minard's graphic

**Exercise: Write out what the grammar of graphics will look for this graph.**

<details> 
  <summary>**Click here for solution**</summary>
  <p>
* Layer
    * Data - `troops`
    * Mapping
        * $x$ and $y$ - troop position (`lat` and `long`)
        * Size - `survivors`
        * Color - `direction`
    * Statistical transformation (stat) - `identity`
    * Geometric object (geom) - `path`
    * Position adjustment (position) - none
* Layer
    * Data - `cities`
    * Mapping
        * $x$ and $y$ - city position (`lat` and `long`)
        * Label - `city`
    * Statistical transformation (stat) - `identity`
    * Geometric object (geom) - `text`
    * Position adjustment (position) - none
* Scale
    * Size - range of widths for troop `path`
    * Color - colors to indicate advancing or retreating troops
* Coordinate system - map projection (Mercator or something else)
* Faceting - none
  </p>
</details>

### Write the R code

First we want to build the layer for the troop movement:

```{r plot_troops}
plot_troops <- ggplot(troops, aes(long, lat)) +
  geom_path(aes(size = survivors,
                color = direction,
                group = group))
plot_troops
```

Next let's add the cities layer:

```{r plot_cities, dependson = 'plot_troops'}
plot_both <- plot_troops + 
  geom_text(data = cities, aes(label = city), size = 4)
plot_both
```

Now that the basic information is on there, we want to clean up the graph and polish the visualization by:

* Adjusting the size scale aesthetics for troop movement to better highlight the loss of troops over the campaign.
* Change the default colors to mimic Minard's original grey and tan palette.
* Change the coordinate system to a map-based system that draws the $x$ and $y$ axes at equal intervals.
* Give the map a title and remove the axis labels.

```{r plot_clean, dependson = 'plot_cities'}
plot_polished <- plot_both +
  scale_size(range = c(0, 12),
             breaks = c(10000, 20000, 30000),
             labels = c("10,000", "20,000", "30,000")) + 
  scale_color_manual(values = c("tan", "grey50")) +
  coord_map() +
  labs(title = "Map of Napoleon's Russian campaign of 1812",
       x = NULL,
       y = NULL)
plot_polished
```

Finally we can change the default `ggplot` theme to remove the background and grid lines, as well as the legend:

```{r plot_final, dependson = 'plot_clean'}
plot_polished +
  theme_void() +
  theme(legend.position = "none")
```

# Practice generating layered graphics using `ggplot2`

Now let's practice generating layered graphics in R using the data from [Gapminder World](https://www.gapminder.org/data/), which compiles country-level data on quality-of-life measures.

## Load the `gapminder` dataset

If you have not already installed the `gapminder` package and you try to load it using the following code, you will get an error:

```{r, eval = FALSE}
library(gapminder)
```

```
Error in library(gapminder) : there is no package called ‘gapminder’
```

If this happens, install the gapminder package by running `install.packages("gapminder")` in your console.

Once you've done this, run the following code to load the gapminder dataset, the `ggplot2` library, and a helper library for printing the contents of `gapminder`:

```{r gapminder}
library(ggplot2)
library(tibble)
library(gapminder)

str(gapminder)
gapminder
```

> Run `?gapminder` in the console to open the help file for the data and definitions for each of the columns.

Using the grammar of graphics and your knowledge of the `ggplot2` library, generate a series of graphs that explore the relationships between specific variables.

## Exercises

**What is the average life expectancy, per continent?**

<details> 
  <summary>Click for the solution</summary>
  <p>
```{r lifeexp}
ggplot(gapminder, aes(continent, lifeExp)) +
  geom_boxplot()
```
  </p>
</details>

**What is the relationship between GDP and life expectancy?**

<details> 
  <summary>Click for the solution</summary>
  <p>
```{r gdp_lifeexp}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point() +
  geom_smooth()
```
  </p>
</details>

**What is causing the outlier in `gdpPercap`?**

<details> 
  <summary>Click for the solution</summary>
  <p>
```{r lifeexp_outlier}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point() +
  geom_smooth() +
  geom_text(aes(label = country))
```
  </p>
</details>


# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```

[^layer]: [Wickham, Hadley. (2010) "A Layered Grammar of Graphics". *Journal of Computational and Graphical Statistics*, 19(1).](http://www.jstor.org.proxy.uchicago.edu/stable/25651297)



