---
title: "Practice transforming college education (data)"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
```

The Department of Education collects [annual statistics on colleges and universities in the United States](https://collegescorecard.ed.gov/). I have included a subset of this data from 2013 in the [`rcfss`](https://github.com/uc-cfss/rcfss) library from GitHub. To install the package, run the command `devtools::install_github("uc-cfss/rcfss")` in the console.

> If you don't already have the `devtools` library installed, you will get an error. Go back and install this first using `install.packages("devtools")`, then run `devtools::install_github("uc-cfss/rcfss")`.

```{r scorecard}
library(rcfss)
data("scorecard")
str(scorecard)
```

Type `?scorecard` in the console to open up the help file for this data set. This includes the documentation for all the variables. Use your knowledge of the `dplyr` functions to perform the following tasks.

# Generate a data frame of schools with a greater than 40% share of first-generation students

<details> 
  <summary>Click for the solution</summary>
  <p>
```{r filter}
filter(scorecard, firstgen > .40)
```
  </p>
</details>

# Generate a data frame with the 10 most expensive colleges in 2013

<details> 
  <summary>Click for the solution</summary>
  <p>
```{r top-10}
# arrange the data frame by cost from highest to lowest and slice the first 10 rows
arrange(scorecard, desc(cost)) %>%
  slice(1:10)

# use the top_n function
top_n(scorecard, 10, wt = cost)
```
  </p>
</details>

# Generate a data frame with the average SAT score for each type of college

<details> 
  <summary>Click for the solution</summary>
  <p>
```{r highest-sat}
scorecard %>%
  group_by(type) %>%
  summarize(mean_sat = mean(satavg, na.rm = TRUE))
```
  </p>
</details>

# Calculate for each school how many students it takes to pay the average faculty member's salary and generate a data frame with the school's name and the calculated value

<details> 
  <summary>Click for the solution</summary>
  <p>
```{r avg-sal}
scorecard %>%
  mutate(ratio = avgfacsal / cost) %>%
  select(name, ratio)
```
  </p>
</details>

# Calculate how many private, nonprofit schools have a smaller cost than the University of Chicago

Hint: the result should be a data frame with one row for the University of Chicago, and a column containing the requested value.

## Report the number as the total number of schools

<details> 
  <summary>Click for the solution</summary>
  <p>
```{r uchicago-raw}
scorecard %>%
  filter(type == "Private, nonprofit") %>%
  arrange(cost) %>%
  mutate(school_cheaper = row_number()) %>%
  filter(name == "University of Chicago")
```
  </p>
</details>

## Report the number as the percentage of schools

<details> 
  <summary>Click for the solution</summary>
  <p>
```{r uchicago-pct}
scorecard %>%
  filter(type == "Private, nonprofit") %>%
  mutate(cost_rank = percent_rank(cost)) %>%
  filter(name == "University of Chicago")
```
  </p>
</details>

# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```

