---
title: "Geospatial visualization"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
library(knitr)
library(broom)
library(stringr)
library(modelr)
library(forcats)
library(ggmap)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal())
```

# Geospatial visualization

Geospatial visualizations are some of the oldest data visualization methods in human existence. Data maps were first popularized in the seventeenth century and have grown in complexity and detail since then. Consider [Google Maps](https://www.google.com/maps), the sheer volume of data depicted, and the analytical pathways available to its users.

Geometric visualizations are used to depict spatial features, and with the incorporation of data reveal additional attributes and information. The main features of a map are defined by its **scale** (the proportion between distances and sizes on the map), its **projection** (how the three-dimensional Earth is represented on a two-dimensional surface), and its **symbols** (how data is depicted and visualized on the map).

# Map boundaries

Drawing maps in R is a layer-like process. Typically you start by drawing the boundaries of the geographic regions you wish to visualize, then you add additional layers defined by other variables:

* Points
* Symbols
* Fills (choropleths)

Let's start by first drawing a map's boundaries. Later on we will fill these in with data and turn them into data visualizations.

## Storing map boundaries

A **geographic information system** (GIS) is software that is "designed to capture, store, manipulate, analyze, manage, and present spatial or geographic data".^[Source: [Wikipedia](https://en.wikipedia.org/wiki/Geographic_information_system).] There are many specialized software packages for spatial data analysis, many of which are commercial or proprietary software. For serious spatial data analysis tasks, you probably want to learn how to use these products. However for casual usage, R has a number of tools for drawing maps, most importantly `ggplot2`.

## Using `maps` boundaries

The `maps` package includes the `map()` function for drawing maps based on bundled geodatabases using the `graphics` package:

```{r maps}
library(maps)

# map of the world
map()

# usa boundaries
map("usa")
map("state")

# county map of illinois
map("county", "illinois")
```

These are fine, but we'd rather use them with our friendly `ggplot2` library. We can do this by converting the geodatabases into data frames for plotting with `ggplot2`.

```{r map_data}
# map of the world
map_data("world") %>%
  as_tibble

# usa boundaries
map_data("usa") %>%
  as_tibble
map_data("state") %>%
  as_tibble

# county map of illinois
map_data("county", "illinois") %>%
  as_tibble
```

`map_data()` converts the geodatabases into data frames where each row is a single point on the map. The resulting data frame contains the following variables:

* `long` - longitude. Things to the west of the prime meridian are negative
* `lat` - latitude
* `order` - this identifies the order `ggplot()` should follow to "connect the dots" and draw the borders
* `region` and `subregion` identify what region or subregion a set of points surrounds
* `group` - this is perhaps the most important variable in the data frame. `ggplot()` uses the `group` aesthetic to determine whether adjacent points should be connected by a line. If they are in the same group, then the points are connected. If not, then they are not connected. This is the same basic principle for standard `geom_line()` plots:

    ```{r group}
    library(gapminder)
    
    # no group aesthetic
    ggplot(gapminder, aes(year, lifeExp)) +
      geom_line()
    
    # with grouping by country
    ggplot(gapminder, aes(year, lifeExp, group = country)) +
      geom_line()
    ```

    Note that `group` is not the same thing as `region` or `subregion`. If a region contains landmasses that are discontiguous, there should be multiple groups to properly draw the region:
    
    ```{r michigan}
    map("state", "michigan")
    ```
    
### Drawing the United States

Let's draw a map of the United States. First we need to store the USA map boundaries in a data frame:

```{r usa-bound}
usa <- as_tibble(map_data("usa"))
usa
```

#### Simple black map

* We can use `geom_polygon()` to draw lines between points and close them up (connect the last point with the first point)
* `x = long` and `y = lat`
* We map the `group` aesthetic to the `group` column

```{r usa-map}
ggplot() +
  geom_polygon(data = usa, aes(x = long, y = lat, group = group))
```

Ta-da! A few things we want to immediately start thinking about. First, because latitude and longitude have absolute relations to one another, we need to fix the aspect ratio so that we don't accidentially compress the graph in one dimension. Fixing the aspect ratio also means that even if you change the outer dimensions of the graph (i.e. adjust the window size), then the aspect ratio of the graph itself remains unchanged. We can do this using the `coord_fixed()` function:

```{r coord-fixed}
ggplot() +
  geom_polygon(data = usa, aes(x = long, y = lat, group = group)) +
  coord_fixed()
```

Now it looks a little squished. You can play around with the aspect ratio to find a better projection:^[Or as we'll see shortly, apply a map projection to the graph.]

```{r coord-fixed-13}
ggplot() +
  geom_polygon(data = usa, aes(x = long, y = lat, group = group)) +
  coord_fixed(1.3)
```

#### Change the colors

Since this is a `ggplot()` object, we can change the fill and color aesthetics for the map:

```{r fill-color}
ggplot() +
  geom_polygon(data = usa, aes(x = long, y = lat, group = group),
               fill = NA, color = "red") + 
  coord_fixed(1.3)

gg1 <- ggplot() + 
  geom_polygon(data = usa, aes(x = long, y = lat, group = group),
               fill = "violet", color = "blue") + 
  coord_fixed(1.3)
gg1
```

#### Always remember to use the group aesthetic

What happens if we plot the map without the group aesthetic?

```{r no-group}
ggplot() + 
  geom_polygon(data = usa, aes(x = long, y = lat),
               fill = "violet", color = "blue") + 
  coord_fixed(1.3)
```

Oops. The map doesn't connect the dots in the correct order.

### State maps

`maps` also comes with a state boundaries geodatabase:

```{r states}
states <- as_tibble(map_data("state"))
states
```

By default, each state is filled with the same color:

```{r states-no-fill}
ggplot(data = states) + 
  geom_polygon(aes(x = long, y = lat, group = group), color = "white") + 
  coord_fixed(1.3)
```

We can adjust this by using the `fill` aesthetic. Here, let's map `region` to fill:

```{r states-fill}
ggplot(data = states) + 
  geom_polygon(aes(x = long, y = lat, fill = region, group = group), color = "white") + 
  coord_fixed(1.3) +
  # turn off color legend
  theme(legend.position = "none")
```

Here, each state is assigned a different color at random. You can start to imagine how we might build a choropleth by mapping a different variable to `fill`, but we'll return to that in a little bit.

#### Plot a subset of states

We can use `filter()` to subset the `states` data frame and draw a map with only a subset of the states. For example, if we want to only graph states in the Midwest:

```{r midwest}
midwest <- filter(states,
                  region %in% c("illinois", "indiana", "iowa",
                                "kansas", "michigan", "minnesota",
                                "missouri", "nebraska", "north dakota",
                                "ohio", "south dakota", "wisconsin"))

ggplot(data = midwest) + 
  geom_polygon(aes(x = long, y = lat, group = group),
               fill = "palegreen", color = "black") + 
  coord_fixed(1.3)
```

#### Zoom in on Illinois and its counties

First let's get the Illinois boundaries:

```{r il-state}
il_df <- filter(states, region == "illinois")
```

Now let's get the accompanying counties:

```{r il-counties}
counties <- as_tibble(map_data("county"))
il_county <- filter(counties, region == "illinois")
il_county
```

Plot the state first. This time, lets' remove all the axes gridlines and background junk using `theme_void()`:

```{r il-map}
il_base <- ggplot(data = il_df,
                  mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "black", fill = "gray")

il_base +
  theme_void()
```

Now let's plot the county boundaries in white:

```{r il-county-plot}
il_base +
  theme_void() + 
  geom_polygon(data = il_county, fill = NA, color = "white") +
  # get the state border back on top
  geom_polygon(color = "black", fill = NA)
```

### But what about Alaska and Hawaii?

If you were observant, you noticed `map_data("states")` only includes the 48 contiguous states in the United States. This is because Alaska and Hawaii exist far off from the rest of the states. What happens if you try to draw a map with them in it?

```{r all-50, echo = FALSE}
library(maptools)
library(rgeos)
library(rgdal)

#state shape file from
# http://www.arcgis.com/home/item.html?id=f7f805eb65eb4ab787a0a3e1116ca7e5
loc <- file.path(tempdir(), "stats_dat")
unzip(system.file("extdata", "states_21basic.zip", package = "fiftystater"),
      exdir = loc)
fifty_states_sp <- readOGR(dsn = loc, layer = "states", verbose = FALSE)

fifty_states_sp %>%
  fortify(region = "STATE_NAME") %>%
  mutate(id = tolower(id)) %>%
  ggplot(mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "black", fill = "gray")
```

Yup, that doesn't look right. Most maps of the United States place Alaska and Hawaii as **insets** to the south of California. Until recently, in R this was an extremely tedious task that required manually changing the latitude and longitude coordinates for these states to place them in the correct location. Fortunately a new package is available that has already done the work for you. [`fiftystater`](https://cran.r-project.org/web/packages/fiftystater/vignettes/fiftystater.html) includes the `fifty_states` data frame which contains adjusted coordinates for Alaska and Hawaii to plot them with the mainland:

```{r fiftystater}
library(fiftystater)

data("fifty_states")
as_tibble(fifty_states)

ggplot(data = fifty_states, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "black", fill = "gray")
```

## Using shapefiles

`maps` contains a very limited number of geodatabases. If you want to import a different country's borders or some other geographic information, you will likely find the data in a **shapefile**. This is a special file format that encodes points, lines, and polygons in geographic space. Files appear with a `.shp` extension, sometimes with accompanying files ending in `.dbf` and `.prj`.

* `.shp` stores the geographic coordinates of the geographic features (e.g. country, state, county)
* `.dbf` stores data associated with the geographic features (e.g. unemployment rate, crime rates, percentage of votes cast for Donald Trump)
* `.prj` stores information about the projection of the coordinates in the shapefile (we'll handle this shortly)

Let's start with a shapefile for state boundaries in the United States.^[Originally obtained from the Census Bureau.] We'll use `readOGR()` from the `rgdal` package to read in the data file:

```{r readshapepoly}
library(rgdal)

usa <- readOGR("data/census_bureau/cb_2013_us_state_20m/cb_2013_us_state_20m.shp")
str(usa, max.level = 2) # view structure of object
```

This is decidedly not a tidy data frame. Once you import the shapefile, it's best to convert it to a data frame for `ggplot()`. We can do this using `fortify()`:

```{r states-convert}
fortify(usa) %>%
  head()
```

Under this approach, the `id` variable is just a number assigned to each region (in this case, each state/territory). However the shapefile contains linked data with attributes for each region. We can access this using the `@data` accessor:

```{r at-data}
as_tibble(usa@data)
```

We can keep these variables in the new data frame through parameters to `fortify(region = "")`:

```{r fortify-id}
# state name
usa %>%
  fortify(region = "NAME") %>%
  head()

# FIPS code
usa %>%
  fortify(region = "STATEFP") %>%
  head()

# keep it all
(usa2 <- usa %>%
  fortify(region = "NAME") %>%
  as_tibble() %>%
  left_join(usa@data, by = c("id" = "NAME")))
```

> [FIPS codes for each state and outlying territory.](https://www.census.gov/geo/reference/ansi_statetables.html)

Now we can plot it like normal using `ggplot()`:

```{r usa2-all}
ggplot(data = usa2, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "black", fill = "gray")
```

Because the data file comes from the Census Bureau, we also get boundaries for Alaska, Hawaii, and Puerto Rico. To remove them from the data, just use `filter()`:

```{r usa2-filter}
usa2 <- usa2 %>%
  filter(id != "Alaska", id != "Hawaii", id != "Puerto Rico")

ggplot(data = usa2, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "black", fill = "gray")
```

## `ggmap`

Rather than relying on geodatabases or shapefiles which store boundaries as numeric data, we can use `ggmap` to retrieve raster map tiles from online mapping services. The most popular provider of these map tiles is [Google Maps](https://www.google.com/maps):

```{r ggmap}
library(ggmap)
```

```{r google-maps}
get_googlemap("university of chicago", zoom = 12) %>%
  ggmap()
```

## Changing map projections

<iframe width="560" height="315" src="https://www.youtube.com/embed/vVX-PrBRtTY?rel=0" frameborder="0" allowfullscreen></iframe>

Representing portions of the globe on a flat surface can be challenging. Depending on how you project the map, you can distort or emphasize certain features of the map. Fortunately, `ggplot()` includes the `coord_map()` function which allows us to easily implement different projection methods.^[This function replaces `coord_fixed()`.] Depending on the projection method, you may need to pass additional arguments to `coord_map()` to define the standard parallel lines used in projecting the map:

```{r projections}
map_proj_base <- ggplot(data = usa2,
                        mapping = aes(x = long, y = lat, group = group)) + 
  geom_polygon(color = "black", fill = "gray")

map_proj_base +
  coord_map() +
  ggtitle("Mercator projection (default)")

map_proj_base +
  coord_map(projection = "albers", lat0 = 25, lat1 = 50) +
  ggtitle("Albers equal-area projection")
```

# Adding data to the map

Region boundaries serve as the background in geospatial data visualization - so now we need to add data. Some types of geometric objects (points and symbols) are overlaid on top of the boundaries, whereas other geoms (fill) are incorporated into the region layer itself. Let's look at the first type of object.

## Points

Let's use our `usa2` map data to add some points. The `airports` data frame in the `nycflights13` package includes geographic info on airports in the United States.

```{r nycflights}
library(nycflights13)
airports
```

Each airport has it's geographic location encoded through `lat` and `lon`. To draw these points on the map, basically we draw a scatterplot with `x = lon` and `y = lat`. In fact we could simply do that:

```{r scatter}
ggplot(airports, aes(lon, lat)) +
  geom_point()
```

Let's overlay it with the mapped state borders:

```{r flights-usa}
ggplot() + 
  coord_map() + 
  geom_polygon(data = usa2, mapping = aes(x = long, y = lat, group = group),
               color = "black", fill = "gray") +
  geom_point(data = airports, aes(x = lon, y = lat), shape = 1)
```

Slight problem. We have airports listed outside of the continental United States. There are a couple ways to rectify this. Unfortunately `airports` does not include a variable identifying state so the `filter()` operation is not that simple. The easiest solution is to crop the limits of the graph to only show the mainland:

```{r crop}
ggplot() + 
  coord_map(xlim = c(-130, -60),
            ylim = c(20, 50)) + 
  geom_polygon(data = usa2, mapping = aes(x = long, y = lat, group = group),
               color = "black", fill = "gray") +
  geom_point(data = airports, aes(x = lon, y = lat), shape = 1)
```

If we want to change the projection method, the points will automatically adjust too:

```{r crop-proj}
ggplot() + 
  coord_map(projection = "albers", lat0 = 25, lat1 = 50,
            xlim = c(-130, -60),
            ylim = c(20, 50)) + 
  geom_polygon(data = usa2, mapping = aes(x = long, y = lat, group = group),
               color = "black", fill = "gray") +
  geom_point(data = airports, aes(x = lon, y = lat), shape = 1)
```

## Symbols

We can change the size or type of symbols on the map. For instance, we can draw a **bubble plot** (also known as a **proportional symbol map**) and encode the altitude of the airport through the size channel:

```{r airport-alt}
ggplot() + 
  coord_map(xlim = c(-130, -60),
            ylim = c(20, 50)) + 
  geom_polygon(data = usa2, mapping = aes(x = long, y = lat, group = group),
               color = "black", fill = "white") +
  geom_point(data = airports, aes(x = lon, y = lat, size = alt),
             fill = "grey", color = "black", alpha = .2) +
  # strip out background junk and remove legend
  theme_void() +
  theme(legend.position = "none")
```

Circle area is proportional to the airport's altitude (in feet). Or we could scale it based on the number of arriving flights in `flights`:

```{r airport-dest}
(airports_n <- flights %>%
  count(dest) %>%
  left_join(airports, by = c("dest" = "faa")))

ggplot() + 
  coord_map(xlim = c(-130, -60),
            ylim = c(20, 50)) + 
  geom_polygon(data = usa2, mapping = aes(x = long, y = lat, group = group),
               color = "black", fill = "white") +
  geom_point(data = airports_n, aes(x = lon, y = lat, size = n),
             fill = "grey", color = "black", alpha = .2) +
  theme_void() +
  theme(legend.position = "none")
```

> `airports` contains a list of virtually all commercial airports in the United States. However `flights` only contains data on flights departing from New York City airports (JFK, LaGuardia, or Newark) and only services a few airports around the country.

# Drawing choropleth maps

**Choropleth maps** encode information by assigning shades of colors to defined areas on a map (e.g. countries, states, counties, zip codes). There are lots of ways to tweak and customize these graphs, which is generally a good idea because decoding and interpreting information conveyed via colorshades is notoriously tricky (if the color palette is a bad choice).

## Loading the data

We'll continue to use the `usa2` shapefile. Let's reload it and also load and tidy the county shapefile:

```{r county-shp}
# load USA shapefile for map boundaries
usa <- readOGR("data/census_bureau/cb_2013_us_state_20m/cb_2013_us_state_20m.shp")

# fortify and filter to keep contiguous USA
usa2 <- usa %>%
  fortify(region = "GEOID") %>%
  as_tibble() %>%
  left_join(usa@data, by = c("id" = "GEOID")) %>%
  # filter out Alaska, Hawaii, Puerto Rico via FIPS codes
  filter(!(STATEFP %in% c("02", "15", "72")))

# do the same thing for the county shapefile
counties <- readOGR("data/census_bureau/cb_2013_us_county_20m/cb_2013_us_county_20m.shp")

counties2 <- counties %>%
  fortify(region = "GEOID") %>%
  as_tibble() %>%
  left_join(counties@data, by = c("id" = "GEOID")) %>%
  # filter out Alaska, Hawaii, Puerto Rico via FIPS codes
  filter(!(STATEFP %in% c("02", "15", "72")))

# plot county map boundaries
ggplot(counties2, mapping = aes(x = long, y = lat, group = group)) + 
  geom_polygon(color = "black", fill = "gray") +
  coord_map()
```

We'll draw choropleths for the number of foreign-born individuals in each region (state or county). We can get those files from the `census_bureau` folder. Let's also normalize our measure by the total population to get the rate of foreign-born individuals in the population:

```{r import-foreign}
(fb_state <- read_csv("data/census_bureau/ACS_13_5YR_B05012_state/ACS_13_5YR_B05012.csv") %>%
  mutate(rate = HD01_VD03 / HD01_VD01))

(fb_county <- read_csv("data/census_bureau/ACS_13_5YR_B05012_county/ACS_13_5YR_B05012.csv") %>%
  mutate(rate = HD01_VD03 / HD01_VD01))
```

## Joining the data to regions

Now that we have our data, we want to draw it on the map. To do that, we have to join together our data sources - the shapefiles and the CSVs. Normally joining data files requires a `_join()` operation of some sort. However when using `ggplot2`, we don't have to do this. Remember that we can pass different data frames into different layers of a `ggplot()` object. Rather than using `geom_polygon()` to draw our maps, now we switch to `geom_map()`:

```{r geom-map-state}
ggplot(fb_state, aes(map_id = GEO.id2)) +
  geom_map(aes(fill = rate), map = usa2) +
  expand_limits(x = usa2$long, y = usa2$lat)
```

Let's break down what just happened:

* `fb_state` is the data frame with the variables we want to visualize
* `map_id = GEO.id2` identifies the column in `fb_state` that uniquely matches each observation to a region in `usa2`
* `geom_map(aes(fill = rate), map = usa2) `
    * `fill = rate` identifies the column in `fb_state` we will use to determine the color of each region
    * `map = usa2` is the data frame containing the boundary coordinates
* `expand_limits(x = usa2$long, y = usa2$lat)` ensures the graph is drawn to the proper window. Because the default data frame for this `ggplot()` object is `fb_state`, it won't contain the necessary information to size the window

We can then tweak this up by adding a title, removing the background (but retaining the legend), and projecting the map using a different method:

```{r geom-map-state-neat}
ggplot(fb_state, aes(map_id = GEO.id2)) +
  geom_map(aes(fill = rate), map = usa2) +
  expand_limits(x = usa2$long, y = usa2$lat) +
  scale_fill_continuous(labels = scales::percent) +
  labs(title = "Rate of foreign-born individuals in the population",
       fill = NULL) +
  ggthemes::theme_map() +
  coord_map(projection = "albers", lat0 = 25, lat1 = 50)
```

We could do the same thing for the county-level data:

```{r geom-map-county}
ggplot(fb_county, aes(map_id = GEO.id2)) +
  geom_map(aes(fill = rate), map = counties2) +
  expand_limits(x = counties2$long, y = counties2$lat) +
  scale_fill_continuous(labels = scales::percent) +
  labs(title = "Rate of foreign-born individuals in the population",
       fill = NULL) +
  ggthemes::theme_map() +
  coord_map(projection = "albers", lat0 = 25, lat1 = 50)
```

### Binning data

> Remember `cut()`? There are some `ggplot2` variants on this function that more explicitly identify how we want to bin the numeric vector (column).

* `cut_interval()` - makes `n` groups with equal range

```{r cut-interval}
fb_county %>%
  mutate(rate_cut = cut_interval(rate, 6)) %>%
  ggplot(aes(map_id = GEO.id2)) +
  geom_map(aes(fill = rate_cut), map = counties2) +
  expand_limits(x = counties2$long, y = counties2$lat) +
  labs(title = "Rate of foreign-born individuals in the population",
       fill = NULL) +
  ggthemes::theme_map() +
  coord_map(projection = "albers", lat0 = 25, lat1 = 50)
```

* `cut_number()` - makes `n` groups with (approximately) equal numbers of observations

```{r cut-number}
fb_county %>%
  mutate(rate_cut = cut_number(rate, 6)) %>%
  ggplot(aes(map_id = GEO.id2)) +
  geom_map(aes(fill = rate_cut), map = counties2) +
  expand_limits(x = counties2$long, y = counties2$lat) +
  labs(title = "Rate of foreign-born individuals in the population",
       fill = NULL) +
  ggthemes::theme_map() +
  coord_map(projection = "albers", lat0 = 25, lat1 = 50)
```

## Defining colors

Selection of your **color palette** is perhaps the most important decision to make when drawing a choropleth.

By default, `ggplot2` picks evenly spaced hues around the [Hue-Chroma-Luminance (HCL) color space](https://en.wikipedia.org/wiki/HCL_color_space):^[Check out chapter 6.6.2 in *`ggplot2`: Elegant Graphics for Data Analysis* for a much more thorough explanation of the theory behind this selection process]

```{r color-wheel, echo = FALSE}
# generate simulated data points
sim_points <- data_frame(x = factor(1:6))

plots <- purrr::map(1:6, ~ ggplot(sim_points[1:.x, ], aes(x, x, color = x)) +
  geom_point(size = 5) +
    ggtitle(paste(.x, "color")) +
  theme(legend.position = "none"))

gridExtra::marrangeGrob(plots, nrow = 2, ncol = 3, top = NULL)
```

`ggplot2` gives you many different ways of defining and customizing your `scale_color_` and `scale_fill_` palettes, but will not tell you if they are optimal for your specific usage in the graph.

### RColorBrewer

[Color Brewer](http://colorbrewer2.org/) is a diagnostic tool for selecting optimal color palettes for maps with discrete variables. The authors have generated different color palettes designed to make differentiating between categories easy depending on the scaling of your variable. All you need to do is define the number of categories in the variable, the nature of your data (sequential, diverging, or qualitative), and a color scheme. There are also options to select palettes that are colorblind safe, print friendly, and photocopy safe. Depending on the combination of options, you may not find any color palette that matches your criteria. In such a case, consider reducing the number of data classes.

#### Sequential

```{r cb-seq}
# define ggplot object
seq_plot <- fb_county %>%
  mutate(rate_cut = cut_number(rate, 6)) %>%
  ggplot(aes(map_id = GEO.id2)) +
  geom_map(aes(fill = rate_cut), map = counties2) +
  expand_limits(x = counties2$long, y = counties2$lat) +
  labs(title = "Rate of foreign-born individuals in the population",
       fill = NULL) +
  ggthemes::theme_map() +
  coord_map(projection = "albers", lat0 = 25, lat1 = 50)

seq_plot +
  scale_fill_brewer(palette = "BuGn")

seq_plot +
  scale_fill_brewer(palette = "YlGn")

seq_plot +
  scale_fill_brewer(palette = "Blues")
```

### Qualitative

```{r qual}
state_data <- data_frame(name = state.name,
                         region = state.region,
                         subregion = state.division,
                         abb = state.abb) %>%
  bind_cols(as_tibble(state.x77)) %>%
  # get id variable into data frame
  left_join(usa2 %>%
              select(id, NAME) %>%
              distinct(),
            by = c("name" = "NAME")) %>%
  # remove Alaska and Hawaii
  na.omit()

# set region base plot
region_p <- ggplot(state_data, aes(map_id = id)) +
  geom_map(aes(fill = region), map = usa2) +
  expand_limits(x = usa2$long, y = usa2$lat) +
  labs(fill = NULL) +
  ggthemes::theme_map() +
  coord_map(projection = "albers", lat0 = 25, lat1 = 50)
region_p

# try different color brewers
region_p +
  scale_fill_brewer(palette = "Paired")

region_p +
  scale_fill_brewer(palette = "Dark2")

region_p +
  scale_fill_brewer(palette = "Pastel2")

# set subregion base plot
subregion_p <- ggplot(state_data, aes(map_id = id)) +
  geom_map(aes(fill = subregion), map = usa2) +
  expand_limits(x = usa2$long, y = usa2$lat) +
  labs(fill = NULL) +
  ggthemes::theme_map() +
  coord_map(projection = "albers", lat0 = 25, lat1 = 50)
subregion_p

subregion_p +
  scale_fill_brewer(palette = "Paired")

subregion_p +
  scale_fill_brewer(palette = "Set1")

subregion_p +
  scale_fill_brewer(palette = "Pastel1")
```

# Faceting maps

If you have geospatial data with a temporal component (i.e. measuring a variable in a country over a period of time) and want to incorporate the temporal data into your visualization, we have a few different options available. Let's demonstrate using data on world fertility rates and a world map.

## Get world fertility rate data

```{r world-shape}
# import and fortify world shapefile
world <- readOGR("data/nautral_earth/ne_110m_admin_0_countries/ne_110m_admin_0_countries.shp")
world2 <- fortify(world, region = "iso_a3")
 
# Country-level data
(fertility <- read_csv("data/API_SP.DYN.TFRT.IN_DS2_en_csv_v2/API_SP.DYN.TFRT.IN_DS2_en_csv_v2.csv") %>%
  select(-X62) %>%
  # tidy the data frame
  gather(year, fertility, `1960`:`2016`, convert = TRUE) %>%
  filter(year < 2015) %>%
  mutate(fertility = as.numeric(fertility),
         # cut into six intervals
         fertility_rate = cut_interval(fertility, 6)))
```

## Separate images

If we want each map stored in a separate image, we could write a standard iterative operation using techniques we already know.

### `For` loop

```{r world-plot-for}
for(year in seq(from = 1970, to = 2010, by = 5)){
  p <- fertility %>%
    filter(year == year) %>%
    ggplot(aes(map_id = `Country Code`)) +
    geom_map(aes(fill = fertility_rate), map = world2) +
    expand_limits(x = world2$long, y = world2$lat) +
    scale_fill_brewer(palette = "BuGn") +
    labs(title = "Fertility rate",
         subtitle = year,
         fill = NULL) +
    ggthemes::theme_map() +
    coord_map(projection = "mollweide", xlim = c(-180, 180))
  print(p)
}
```

## `purrr::map()`

```{r world-plot-map}
seq(from = 1970, to = 2010, by = 5) %>%
  purrr::map(~ fertility %>%
    filter(year == .x) %>%
    ggplot(aes(map_id = `Country Code`)) +
    geom_map(aes(fill = fertility_rate), map = world2) +
    expand_limits(x = world2$long, y = world2$lat) +
    scale_fill_brewer(palette = "BuGn") +
    labs(title = "Fertility rate",
         subtitle = .x,
         fill = NULL) +
    ggthemes::theme_map() +
    coord_map(projection = "mollweide", xlim = c(-180, 180)))
```

## `facet_wrap()`

Or if we want them in a single image, we can use `facet_wrap()` or `facet_grid()` to draw the small multiples plot:

```{r world-facet-grid-all}
ggplot(fertility, aes(map_id = `Country Code`)) +
  facet_wrap(~ year) +
  geom_map(aes(fill = fertility_rate), map = world2) +
  expand_limits(x = world2$long, y = world2$lat) +
  scale_fill_brewer(palette = "BuGn") +
  labs(title = "Fertility rate",
       fill = NULL) +
  ggthemes::theme_map() +
  coord_map(projection = "mollweide", xlim = c(-180, 180)) +
  theme(legend.position = "none")
```

```{r world-facet-grid-south-africa}
south_africa <- world2 %>%
  filter(id == "ZAF")

fertility %>%
  filter(`Country Name` == "South Africa") %>%
  ggplot(aes(map_id = `Country Code`)) +
  facet_wrap(~ year) +
  geom_map(aes(fill = fertility_rate), map = south_africa) +
  expand_limits(x = south_africa$long, y = south_africa$lat) +
  scale_fill_brewer(palette = "BuGn") +
  labs(title = "Fertility rate",
       subtitle = "South Africa",
       fill = NULL) +
  ggthemes::theme_map() +
  coord_map() +
  theme(legend.position = "none")
```

## `gganimate`

Alternatively, we can make an animated visualization using the [`gganimate`](https://github.com/dgrtwo/gganimate) package. The major difference is that we now add a `frame` aesthetic which defines the variable we are using to animate the graph.

```{r gganimate, fig.show = "animate"}
library(gganimate)

p <- ggplot(fertility, aes(map_id = `Country Code`, frame = year)) +
  geom_map(aes(fill = fertility_rate), map = world2) +
  expand_limits(x = world2$long, y = world2$lat) +
  scale_fill_brewer(palette = "BuGn") +
  labs(title = "Fertility rate",
       fill = NULL) +
  ggthemes::theme_map() +
  coord_map(projection = "mollweide", xlim = c(-180, 180)) +
  theme(legend.position = "none")
gganimate(p)
```

# Acknowledgments {.toc-ignore}

* [Making Maps with R](http://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html)
* [Mapping Geographic Data in R](https://flowingdata.com/geographic-r/)

# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```
