---
title: "Drawing interactive maps with `leaflet`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
library(leaflet)
library(RColorBrewer)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal())
```

# Intro to Leaflet

[Leaflet](https://leafletjs.com/) is an open-source JavaScript library for creating interactive maps. Unlike static visualization packages such as `ggplot2` or `ggmap`, Leaflet maps are fully interactive and can include features such as:

* Interactive panning/zooming
* Pop-up tooltips and labels
* Highlighting/selecting regions

It is used by many news organizations and tech websites to visualize geographic data. The `leaflet` package for R enables the creation of interactive maps within R without learning how to write JavaScript code. The [`leaflet` documentation](https://rstudio.github.io/leaflet/) is a handy walkthrough for the basics of creating Leaflet maps in R. Let's explore here how to create Leaflet maps using the same data we used to create [raster maps with `ggmap`](geoviz_ggmap.html), [crime data from the city of Chicago in 2017](https://data.cityofchicago.org/Public-Safety/Crimes-2017/d62x-nvdr).^[[Full documentation of the data from the larger 2001-present crime dataset.](https://data.cityofchicago.org/Public-Safety/Crimes-2001-to-present/ijzp-q8t2).]

```{r import-crimes}
crimes <- read_csv("data/Crimes_-_2017.csv")
glimpse(crimes)
```

# Basic usage

Leaflet maps are built using layers, similar to `ggplot2`.

1. Create a map widget by calling `leaflet()`
1. Add **layers** to the map using one or more of the layer functions (e.g. `addTiles()`, `addMarkers()`, `addPolygons()`)
1. Repeat step 2 as many times as necessary to incorporate the necessary information
1. Display the map widget

A basic example is:

```{r leaflet-basic}
m <- leaflet() %>%
  addTiles() %>%
  addMarkers(lng = -87.597241, lat = 41.789829,
             popup = "Saieh Hall of Economics")
m
```

# Basemaps

Like `ggmap`, `leaflet` supports basemaps using map tiles. By default, OpenStreetMap tiles are used.

```{r basemap}
m <- leaflet() %>%
  setView(lng = -87.597241, lat = 41.789829, zoom = 14)
m %>%
  addTiles()
```

To use a different basemap provider, use `addProviderTiles()` with the name of the provider you wish to implement. You can view the complete list of providers using `names(providers)`.

```{r basemap-third-party}
m %>% addProviderTiles(providers$Stamen.Toner)
m %>% addProviderTiles(providers$CartoDB.Positron)
m %>% addProviderTiles(providers$Esri.NatGeoWorldMap)
m %>% addProviderTiles(providers$Wikimedia)
```

# Add markers


# Add labels


# Add lines and shapes


# Add legend


# Choropleth of crime rates by neighborhood

# Additional resources {.toc-ignore}


# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```
