---
title: "Personalize Git"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

*You only have to do this once per machine.*

In order to track changes and attribute them to the correct user, we need to tell Git your name and email address. Open the shell on your computer. From there, type the following commands (replace the relevant parts with your own information):

* `git config --global user.name 'Benjamin Soltoff'`
    * This can be your full name, your username on GitHub, whatever you want. Each of your commits will be logged with this name, so make sure it is informative *for others*.
* `git config --global user.email 'ben@bensoltoff.com'`
    * This must be the email address you used to register on GitHub.

You will not see any output from these commands. To ensure the changes were made, run `git config --global --list`.

### Acknowledgments

```{r child='_ack_stat545.Rmd'}
```
