---
title: "Homework 06: Collecting data from the web"
output: html_document
---

# Overview

Due before class Wednesday November 9th.

We learned three ways of collecting data from the internet:

* Accessing data using packages that wrap APIs
* Running (basic) API queries "by hand"
* Web scraping

For the homework, you will

* Combine two existing datasets in a novel (and reproducible!) way
* Create a new dataset by writing your own API interface or web scraping

# Fork the `hw06` repository

Go [here](https://github.com/uc-cfss/hw06) to fork the repo for homework 06.

# Part 1: Exploring the `gapminder` data (even more)

We've examined the `gapminder` data quite a bit. One relationship we've looked at (or are about to) is the relationship between population and life expectancy.

```{r, message = FALSE}
library(tidyverse)
library(gapminder)

gapminder %>%
  # convert to population in millions
  mutate(pop = pop / 1000000) %>%
  ggplot(aes(pop, lifeExp)) +
  geom_point(alpha = .2) +
  geom_smooth() +
  scale_x_log10() +
  labs(title = "As population increases, life expectancy increases",
       x = "Population (in millions)",
       y = "Average life expectancy")
```

For the assignment, I want you to replace *population* with *population density*. To do this:

1. Get the country information using [geonames](http://www.geonames.org/) - remember there is a [R package for this](https://cran.r-project.org/web/packages/geonames/index.html) (see the lecture notes)
1. Merge `gapminder` and the country information from `geonames`
    * Use `left_join` from `dplyr` to [merge the tables](http://r4ds.had.co.nz/relational-data.html)
    * Note that you cannot directly do this - `gapminder` writes the name of countries differently from `geonames`. To complete the merge, you need a unique *key* to match observations between the data frames
    * There is neat little package for R called [`countrycode`](https://github.com/vincentarelbundock/countrycode) that helps solve this problem. `countrycode()` takes as an input the column of country names in a specific format and outputs it using whatever format you specify.
        * `gapminder` stores them using the `country.name` format
        * `geonames` stores them under the `countryCode` column using the `iso2c` format
        * I leave it to you to make the joining operation work
1. Calculate the population density for each observation
1. Produce an updated graph using population density
    * If you want to be real fancy, estimate a statistical model or compare the relationship across continents

# Part 2: More complicated data collection

For the second part of the assignment, I want you to create a new dataset. Note that I am not asking you to analyze it, simply collect it. Don't make it harder than it needs to be.

Unlike the first part, you cannot use a pre-written package in R or Python to obtain your data. If you go the API route, you need to write your own code or function to query the server and obtain the results. If you decide to skip the API entirely, you will need to use `rvest` or `BeautifulSoup` to scrape the content of a web page and extract the relevant information.

The end result must be a tidy data frame stored in the repository ready to be analyzed. I should be able to run your code and reproduce your data.^[Obviously if you are scraping from a web page that frequently updates its content, I may not perfectly reproduce your results. That's fine - just make sure you've saved a copy of the data frame in the repo.]

# Submit the assignment

Your assignment should be submitted as a set of R scripts, R Markdown documents, Jupyter Notebooks, data files, etc. Whatever is necessary to show your code and present your results. Follow instructions on [homework workflow](hw00_homework_guidelines.html#homework_workflow). As part of the pull request, you're encouraged to reflect on what was hard/easy, problems you solved, helpful tutorials you read, etc.

# Rubric

Check minus:

Check: Solid effort. Hits all the elements. No clear mistakes. Easy to follow (both the code and the output). Nothing spectacular, either bad or good.

Check plus:

### Acknowledgments {.toc-ignore}

```{r child='_ack_stat545.Rmd'}
```

