---
title: "Homework 07: Analyzing text data"
output: html_document
---

# Overview

Due before class Wednesday November 16th.

# Fork the `hw07` repository

Go [here](https://github.com/uc-cfss/hw07) to fork the repo for homework 07.

# Your mission

Perform text analysis.

## Okay, I need more information

Perform sentiment analysis or topic modeling using text analysis methods as demonstrated in class and in the readings.

## Okay, I need even more information.

Do the above. Can't think of a data source?

* `gutenbergr`
* `AssociatedPress` from the `topicmodels` package
* `NYTimes` or `USCongress` from the `RTextTools` package
* [Reuters-21578](http://www.daviddlewis.com/resources/testcollections/reuters21578/) - a standard set of text documents (articles published by Reuters in 1987). To access the document-term matrix for this data set, run the following code:
    ```{r, eval = FALSE}
    install.packages("tm.corpus.Reuters21578", repos = "http://datacube.wu.ac.at")
    library(tm.corpus.Reuters21578)
    data("Reuters21578_DTM")
    ```
* [State of the Union speeches](http://www.presidency.ucsb.edu/sou.php)
* Scrape tweets using `twitteR` (you know how to use the API now, right?)

Analyze the text for sentiment OR topic. **You do not need to do both**. The lecture notes and [Tidy Text Mining with R](http://tidytextmining.com/) are good starting points for templates to perform this type of analysis.

# Submit the assignment

Your assignment should be submitted as a set of R scripts, R Markdown documents, Jupyter Notebooks, data files, etc. Whatever is necessary to show your code and present your results. Follow instructions on [homework workflow](hw00_homework_guidelines.html#homework_workflow). As part of the pull request, you're encouraged to reflect on what was hard/easy, problems you solved, helpful tutorials you read, etc.

# Rubric

Check minus: TBD

Check: Solid effort. Hits all the elements. No clear mistakes. Easy to follow (both the code and the output). Nothing spectacular, either bad or good.

Check plus: TBD

### Acknowledgments {.toc-ignore}

```{r child='_ack_stat545.Rmd'}
```

