---
title: "Presenting academic research"
author: "MACS 30200 - Perspectives on Computational Research"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Objectives

* Identify major components of a research poster
* Discuss differences between a research paper and a research poster
* Present examples of research posters
* Identify templates for creating research posters in Powerpoint, $\LaTeX$, and R Markdown
* Identify locations to print research posters at UChicago

```{r packages, cache = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(broom)
library(forcats)
library(modelr)
library(stringr)
library(car)
library(rcfss)
library(coefplot)
library(RColorBrewer)
library(lme4)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal())
```

# Research posters


# Templates for creating research posters

## Powerpoint

* [UChicago template](https://github.com/UC-MACSS/persp-research/blob/master/demos/E-Poster%20Template%20Academic%20conference.ppt)
* [Poster Presentations templates](https://www.posterpresentations.com/html/free_poster_templates.html) - [PosterPresentations.com](http://www.posterpresentations.com) provides poster printing services (shipped to you, so turnaround time is longer), but also offers free Powerpoint templates

## $\LaTeX$

* [$\LaTeX$ poster template](http://www.brian-amberg.de/uni/poster/) - older and clunkier, but what I used for the first example
* [`beamerposter`](https://github.com/deselaers/latex-beamerposter) - used for the second example poster

## R Markdown

* [`flexdashboard` poster](https://github.com/odeleongt/flexdashboard-poster) - uses `flexdashboard` to design the poster. Rendering as a PDF seems to be broken, but rendering as a `.png` file works fine and quality looks good. Need to adjust the parameters in the rendering script to get an appropriate size and aspect ratio
* [`drposter`](https://github.com/bbucior/drposter) - package for generating posters in HTML/CSS using `reveal.js` to generate the pdf document. Still in alpha testing, not sure how effective this will be

# Acknowledgments {.toc-ignore}


# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```




