---
title: "Presenting academic research"
author: "MACS 30200 - Perspectives on Computational Research"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Objectives

* Identify major components of a research poster
* Discuss differences between a research paper and a research poster
* Present examples of research posters
* Identify templates for creating research posters in Powerpoint, $\LaTeX$, and R Markdown
* Identify locations to print research posters at UChicago

```{r packages, cache = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(broom)
library(forcats)
library(modelr)
library(stringr)
library(car)
library(rcfss)
library(coefplot)
library(RColorBrewer)
library(lme4)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal())
```

# Designing research posters

## Potential modes of presenting results at conferences

Conferences are a common location to present research and share results with the larger academic community. Depending on the conference, you may have opportunities to present work in the following ways:

### Oral presentation

Typically you are a part of a panel of related papers (between 4-6 papers per panel). You submit a proposal months in advance, typically just based on the abstract.^[Some highly selective conferences require you to submit the entire paper when you apply for the conference.] You write the paper and send it to the other members of the panel. The panel consists of the other presenters, as well as a **chair** and/or **discussants**. Discussants are assigned to read the papers in advance of the conference and prepare a critique to share during the panel session. The chair may also serve as a discussant, but more generally is responsible for cajoling the panel members into submitting their papers in a timely fashion, ensuring the discussants read the papers and show up for the panel, and keep track of time during the panel to make sure panelists stick to the time requirements.

During the panel session, each panelist is allotted between 8-15 minutes to present their paper. Panelists typically prepare a slideshow to support their oral presentation. There may be an additional 5-10 minutes of critique from the discussants, and then Q&A with the audience.

#### Lightning talk

**Lightning talks** are a specialized format that condenses oral presentations into rapid fire forms. Individuals typically get five minutes for their oral presentation, with talks arranged back-to-back. Many lightning talks also standardize the slides format, so that the speaker can use precisely 10 slides that auto-advance every 30 seconds. This format forces speakers to condense their points and be very clear and precise to convey only the critical information.

### Poster presentation

While graduate students can be accepted at conferences as part of panels, many graduate students participate in **poster sessions**. Again I am generalizing, but typically conferences group research projects by common themes or subdisciplines with all speakers scheduled for a 2 hour session. During this time the posters are on display and the authors stand with their posters as interested audience members walk around and view the posters. Unlike with oral presentations, all interaction is one-on-one with the audience. The poster serves as the visual medium for drawing in the audience and communicating the key findings of the project, and the audience can then ask follow-up questions of the author.

## Goals of a research poster

A research poster can be thought of as an extended abstract for your research paper or project. It is an opportunity to present research results without requiring the audience to read a full paper or even attend a guest lecture or oral presentation. It is a **snapshot** of your research - more in-depth than an abstract, but less in-depth than the full paper. Designing a good-quality research poster is tricky, as you want to provide enough information so that readers can engage in a detailed conversation about the work, but remain short enough so that a reader can consume it in just a few minutes and walk away understanding the importance of the research.

## Major components


## Tips for designing poster content and layouts



# Templates for creating research posters

## Powerpoint

* [UChicago template](https://github.com/UC-MACSS/persp-research/blob/master/demos/E-Poster%20Template%20Academic%20conference.ppt)
* [Poster Presentations templates](https://www.posterpresentations.com/html/free_poster_templates.html) - [PosterPresentations.com](http://www.posterpresentations.com) provides poster printing services (shipped to you, so turnaround time is longer), but also offers free Powerpoint templates

## $\LaTeX$

* [$\LaTeX$ poster template](http://www.brian-amberg.de/uni/poster/) - older and clunkier, but what I used for the first example
* [`beamerposter`](https://github.com/deselaers/latex-beamerposter) - used for the second example poster

## R Markdown

* [`flexdashboard` poster](https://github.com/odeleongt/flexdashboard-poster) - uses `flexdashboard` to design the poster. Rendering as a PDF seems to be broken, but rendering as a `.png` file works fine and quality looks good. Need to adjust the parameters in the rendering script to get an appropriate size and aspect ratio
* [`drposter`](https://github.com/bbucior/drposter) - package for generating posters in HTML/CSS using `reveal.js` to generate the pdf document. Still in alpha testing, not sure how effective this will be

# Acknowledgments {.toc-ignore}


# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```




