---
title: "Syllabus"
---

```{r include = FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
library(stringr)
```

```{r include = FALSE}
# import excel table with topics
rawdat <- read_excel(file.path("course_admin", "schedule.xlsx"))

# clean up table and extract date info
cdat <- rawdat %>%
  mutate(cm = str_replace_na(cm),
         link_it = as.logical(link_it),
         topic = ifelse(is.na(topic), "", topic),
         notes = sprintf("%s: %s", cm, topic))

# identify links to course modules
cmdat <- data_frame(html = list.files(pattern = '^(cm|setup|git|lab).*html$'))
cmdat <- cmdat %>%
  mutate(cm = str_extract(html, "(cm|setup|git|lab)[0-9]+")) %>%
  filter(!str_detect(html, "slides"))
cdat <- cdat %>%
  left_join(cmdat)

# add links to notes column
cdat$notes[cdat$link_it] <-
  str_replace(cdat$notes[cdat$link_it],
              "((cm[0-9]{3})|((setup|git|lab)[0-9]{2}))", '<a href="LINK">\\1</a>')
cdat$notes[cdat$link_it] <-
  str_replace(cdat$notes[cdat$link_it], "LINK", cdat$html[cdat$link_it])
```

```{r results = 'asis', echo = FALSE, message = FALSE}
cdat %>%
  select(cm, date, notes) %>%
  mutate(cm = str_extract(cm, "[:alpha:]+")) %>%
  spread(cm, notes) %>%
  mutate(lab = ifelse(is.na(lab), "", lab),
         date = sprintf("%s %s %s",
                        wday(date, label = TRUE),
                        month(date, label = TRUE),
                        day(date))) %>%
  knitr::kable(col.names = c("Date", "Topic", "Lab"))
```

## Homework

* [Homework 01: Edit README.md](hw01-edit-README.html)
* [Homework 02: Exploring and visualizing data](hw02-explore-data.html)
* [Homework 03: Wrangling and visualizing data](hw03-wrangle-data.html)
* [Homework 04: Programming in R](hw04-programming.html)
* [Homework 05: Generating reproducible research](hw05-reproducible-research.html)
* [Homework 06: Statistical learning](hw06-stat-learn.html)
* [Homework 07: Resampling methods and distributed learning](hw07-resamp-distrib-learn.html)
* [Homework 08: Collecting and analyzing data from the web](hw08-webdata.html)
* [Homework 09: Analyzing text data](hw09-text_analysis.html)
* [Homework 10: Build a Shiny application](hw10-shiny.html)
