---
title: "Practicing `tidytext` with song titles"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r packages, cache = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(acs)
library(tidytext)

set.seed(1234)
theme_set(theme_minimal())
```

Today let's practice our `tidytext` skills with a basic analysis of song titles. That is, how often is each U.S. state mentioned in a popular song? We'll define popular songs as those in *Billboard*'s Year-End Hot 100 from 1958 to the present.

# Download population data for U.S. states

> To import the data in-class, just use `read_csv("http://cfss.uchicago.edu/data/pop2014.csv")`

First let's use the `acs` package to access the U.S. Census Bureau API and obtain population numbers for each state in 2014. This will help us later to normalize state mentions based on relative population size.^[For instance, California has a lot more people than Rhode Island so it makes sense that California would be mentioned more often in popular songs. But per capita, are these mentions different?]

```{r acs-pop}
# retrieve state populations in 2014 from Census Bureau
stategeo <- geo.make(state = "*")
popfetch <- acs.fetch(geography = stategeo, 
                      endyear = 2014,
                      span = 5, 
                      table.number = "B01003",
                      col.names = "pretty",
                      key = getOption("census_key"))

# convert to data frame and keep only relevant columns/rows
pop_df <- popfetch %>%
  estimate %>%
  reshape2::melt() %>%
  as_tibble %>%
  mutate(state_name = tolower(Var1),
         pop2014 = value) %>%
  select(state_name, pop2014) %>%
  filter(state_name != "puerto rico") %>%
  write_csv("data/pop2014.csv")

# do these results make sense?
pop_df %>% 
  arrange(desc(pop2014)) %>%
  top_n(10)
```

# Retrieve song lyrics

> To import the data in-class, just use `read_csv("http://cfss.uchicago.edu/data/billboard_lyrics_1964-2015.csv")`

Next we need to retrieve the song lyrics for all our songs. [Kaylin Walker](http://kaylinwalker.com/50-years-of-pop-music/) provides a [GitHub repo with the necessary files.](https://github.com/walkerkq/musiclyrics)

```{r lyrics-download, include = FALSE, eval = FALSE}
# download file from GitHub repo
library(downloader)
download("https://github.com/walkerkq/musiclyrics/raw/master/billboard_lyrics_1964-2015.csv",
         destfile = "data/billboard_lyrics_1964-2015.csv")
```

```{r lyrics-import}
# import into R
song_lyrics <- read_csv("data/billboard_lyrics_1964-2015.csv")
names(song_lyrics)
```

# Find and visualize the state names in the song lyrics

Now your work begins!

### Use `tidytext` to create a data frame with one row for each token in each song

Hint: To search for matching state names, this data frame should include both **unigrams** and **bi-grams**.

<details> 
  <summary>Click for the solution</summary>
  <p>
  
```{r lyrics-tidy}
tidy_lyrics <- bind_rows(song_lyrics %>% 
                           unnest_tokens(output = state_name,
                                         input = Lyrics),
                         song_lyrics %>% 
                           unnest_tokens(output = state_name,
                                         input = Lyrics, 
                                         token = "ngrams", n = 2))
tidy_lyrics
```

The variable `state_name` in this data frame contains all the possible words and bigrams that might be state names in all the lyrics.

  </p>
</details>

### Find all the state names occurring in the song lyrics

First create a data frame that meets this criteria, then save a new data frame that only includes one observation for each matching song. That is, if the song is ["New York, New York"](https://www.youtube.com/watch?v=btFfXgUdIzY), there should only be one row in the resulting table for that song.

<details> 
  <summary>Click for the solution</summary>
  <p>
  
```{r lyrics-state-names}
inner_join(tidy_lyrics, pop_df)
```

Let's only count each state once per song that it is mentioned in.

```{r lyrics-state-distinct}
tidy_lyrics <- inner_join(tidy_lyrics, pop_df) %>%
  distinct(Rank, Song, Artist, Year, state_name, .keep_all = TRUE)
tidy_lyrics
```

  </p>
</details>

### Calculate the frequency for each state's mention in a song and create a new column for the frequency adjusted by the state's population

<details> 
  <summary>Click for the solution</summary>
  <p>
  
```{r state-counts}
(state_counts <- tidy_lyrics %>% 
   count(state_name) %>% 
   arrange(desc(n)))
```

```{r state-counts-poprel}
pop_df <- pop_df %>% 
  left_join(state_counts) %>% 
  mutate(rate = n / pop2014 * 1e6)

pop_df %>%
  arrange(desc(rate)) %>%
  top_n(10)
```

  </p>
</details>

### Make a choropleth map for both the raw frequency counts and relative frequency counts

The [`statebins` package](https://github.com/hrbrmstr/statebins) is a nifty shortcut for making basic U.S. cartogram maps.

```{r state-map, warning = FALSE}
library(statebins)

pop_df %>%
  mutate(state_name = stringr::str_to_title(state_name),
         state_name = if_else(state_name == "District Of Columbia",
                              "District of Columbia", state_name)) %>%
  statebins_continuous(state_col = "state_name", value_col = "n") +
  labs(title = "Frequency of states mentioned in song lyrics",
       subtitle = "Number of mentions") +
  theme(legend.position = "bottom")

pop_df %>%
  mutate(state_name = stringr::str_to_title(state_name),
         state_name = if_else(state_name == "District Of Columbia",
                              "District of Columbia", state_name)) %>%
  statebins_continuous(state_col = "state_name", value_col = "rate") +
  labs(title = "Frequency of states mentioned in song lyrics",
       subtitle = "Number of mentions per capita") +
  theme(legend.position = "bottom")
```

# Acknowledgments {.toc-ignore}

* This page is derived in part from [SONG LYRICS ACROSS THE UNITED STATES](https://juliasilge.com/blog/song-lyrics-across/) and licensed under a [Creative Commons Attribution-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-sa/4.0/).

# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```
