---
title: "Text analysis: fundamentals and sentiment analysis"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(stringr)
library(tidytext)
library(broom)
library(scales)

theme_set(theme_bw())
```

# Objectives

* Identify the basic workflow for conducting text analysis
* Define *sentiment analysis* and review how sentiment analysis was used in *Conservatives report, but liberals display, greater happiness*
* Define the *tidy text format*
* Demonstrate how to conduct sentiment analysis using David Robinson's [Text analysis of Trump's tweets confirms he writes only the (angrier) Android half](http://varianceexplained.org/r/trump-tweets/)

# Basic workflow for text analysis

Draws on [Common Text Mining Workflow](https://dzone.com/articles/common-text-mining-workflow)

* Obtain your text sources
* Extract documents and move into a corpus
* Transformation
* Extract features
* Perform analysis

## Obtain your text sources

Can come from lots of areas:

* Web sites
    * Twitter
* Databases
* PDF documents
* Digital scans of printed materials

The easier to convert your text data into digitally stored text, the cleaner your results and fewer transcription errors

## Extract documents and move into a corpus

* *Text corpus* - large and structured set of texts
* Typically stores the text as a raw character string with metadata and details stored with the text

## Transformation

Examples of typical transformations include:

* Tag segments of speech for part-of-speech (nouns, verbs, adjectives, etc.) or entity recognition (person, place, company, etc.)
* Standard text processing
    * Convert to lower case
    * Remove punctuation
    * Remove numbers
    * Remove *stopwords* - common parts of speech that are not informative such as *a*, *an*, *be*, *of*, etc.
    * Remove domain-specific stopwords
    * Stemming - reduce words to their word stem
        * "Fishing", "fished", and "fisher" -> "fish"

## Extract features

* Convert the text string into some sort of quantifiable measures
* Bag-of-words model
    * Each document is represented as a vector which counts the frequency of each term's appearance in the document
    * Combine all the vectors for each document together and you create a *term-document matrix*
        * Each row is a document
        * Each column is a term
        * Each cell represents the frequency of the term appearing in the document
    * Ignores context
        * Could randomly scramble the order of terms appearing in the document
        * Would still get the same term-document matrix

## Perform analysis

* Lots of approaches to take
* Basic
    * Word frequency
    * Collocation - words commonly appearing near each other
    * Dictionary tagging - locating a specific set of words in the texts
* Advanced
    * Document classification
        * Supervised
        * Unsupervised
    * Corpora comparison - comparing the content of different groups of text
    * Detecting clusters of document features - *topic modeling*

# *Conservatives report, but liberals display, greater happiness*

Based on a theory of psychology that conservatives are generally happier than liberals. Authors argue that previous findings are flawed because they rely on self-reported happiness, which is an inaccurate measure of individual happiness. Instead, they measure happiness (in part) based on the text of Congressional speeches and Twitter posts.

## Sentiment analysis

*Sentiment analysis* uses text analysis to estimate the attitude of a speaker or writer with respect to some topic or the overall polarity of the document. For example, the sentence

> I am happy

contains words and language typically associated with positive feelings and emotions. Therefore if some one tweeted "I am happy", we could make an educated guess that the person is expressing positive feelings.

Obviously it would be difficult for us to create a complete dictionary that classifies words based on their emotional affect; fortunately other scholars have already done this for us. Some simply classify words and terms as positive or negative:

```{r}
get_sentiments("bing")
```

Others rate them on a numeric scale:

```{r}
get_sentiments("afinn")
```

Still others rate words based on specific sentiments

```{r}
get_sentiments("nrc")

get_sentiments("nrc") %>%
  count(sentiment)
```

In order to assess the document or speaker's overall sentiment, you simply count up the number of words associated with each sentiment. For instance, [how positive or negative are Jane Austen's novels](http://tidytextmining.com/sentiment.html#sentiment-analysis-with-inner-join)? We can determine this by counting up the number of positive and negative words in each chapter, like so:

```{r}
library(janeaustenr)

tidy_books <- austen_books() %>%
  group_by(book) %>%
  mutate(linenumber = row_number(),
         chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]", 
                                                 ignore_case = TRUE)))) %>%
  ungroup() %>%
  unnest_tokens(word, text)

janeaustensentiment <- tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(book, index = linenumber %/% 80, sentiment) %>%
  spread(sentiment, n, fill = 0) %>%
  mutate(sentiment = positive - negative)

ggplot(janeaustensentiment, aes(index, sentiment, fill = book)) +
        geom_bar(alpha = 0.8, stat = "identity", show.legend = FALSE) +
        facet_wrap(~book, ncol = 2, scales = "free_x")
```

Ignoring the specific code, this is a relatively simple operation. Once you have the text converted into a format suitable for analysis, tabulating and counting term frequency is not a complicated operation.

You can also conduct further analysis once you have this information. In this article, the researchers used speeches from the Congressional record as their corpus. They read this into a software package, converted the text into individual terms within each politician's speeches, and counted up the number of positive and negative terms spoken by each member of Congress. Once they had this value for each politician, they could build a statistical model to predict how positive or negative an individual would sound based on known characteristics (e.g. political conservatism/liberalism, party membership, chamber majority, age, ethnicity, sex)

Likewise, the authors performed a similar analysis on a corpus of tweets from followers of the Democratic and Republican parties. Using the Twitter API, they harvested the tweets, imported them into their software, then conducted the same type of analysis - counting the number of positive and negative words in each tweet, then aggregating for each political party.

# `tidytext`

A lot of these actions can be performed using tidy methods, as long as the text is stored in a tidy format. The [`tidytext`](https://github.com/juliasilge/tidytext) package is a recent addition to the `tidyverse` that attempts to implement a tidy framework for text analysis in R. To my knowledge, there is no comparable tidy method for Python. Python is arguably stronger than R in some aspects of text analysis, however if we want to perform this work primarily using a tidy approach R is the way to go.

## Tidy text format

* Defined as one-term-per-row
* Differs from the document-term matrix (one-document-per-row and one-term-per-column)

# Exploring content of Donald Trump's Twitter timeline

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Every non-hyperbolic tweet is from iPhone (his staff). <br><br>Every hyperbolic tweet is from Android (from him). <a href="https://t.co/GWr6D8h5ed">pic.twitter.com/GWr6D8h5ed</a></p>&mdash; Todd Vaziri (\@tvaziri) <a href="https://twitter.com/tvaziri/status/762005541388378112">August 6, 2016</a></blockquote>
<script async src="http://platform.twitter.com/widgets.js" charset="utf-8"></script>

If you want to know what Donald Trump personally tweets from his account versus his handlers, it looks like we might have a way of detecting this difference. Tweets from an iPhone are his staff; tweets from an Android are from him. Can we quantify this behavior or use text analysis to lend evidence to this argument? **Yes**.

## Obtaining documents

```{r}
library(twitteR)
```

```{r, eval = FALSE}
# You'd need to set global options with an authenticated app
setup_twitter_oauth(getOption("twitter_api_key"),
                    getOption("twitter_api_token"))

# We can request only 3200 tweets at a time; it will return fewer
# depending on the API
trump_tweets <- userTimeline("realDonaldTrump", n = 3200)
trump_tweets_df <- trump_tweets %>%
  map_df(as.data.frame) %>%
  tbl_df()
```

```{r trump_tweets_df}
# if you want to follow along without setting up Twitter authentication,
# just use this dataset:
load(url("http://varianceexplained.org/files/trump_tweets_df.rda"))
```

## Clean up the data

```{r tweets, dependson = "trump_tweets_df"}
tweets <- trump_tweets_df %>%
  select(id, statusSource, text, created) %>%
  extract(statusSource, "source", "Twitter for (.*?)<") %>%
  filter(source %in% c("iPhone", "Android"))

tweets %>%
  head() %>%
  knitr::kable(caption = "Example of Donald Trump tweets")
```

## Comparison using metadata

One consideration is what time of day the tweets occur, which we'd expect to be a "signature" of their user. Here we can certainly spot a difference:

```{r dependson = "tweets"}
tweets %>%
  count(source, hour = hour(with_tz(created, "EST"))) %>%
  mutate(percent = n / sum(n)) %>%
  ggplot(aes(hour, percent, color = source)) +
  geom_line() +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Hour of day (EST)",
       y = "% of tweets",
       color = "")
```

Trump on the Android does a lot more tweeting in the morning, while the campaign posts from the iPhone more in the afternoon and early evening.

Another place we can spot a difference is in Trump's anachronistic behavior of "manually retweeting" people by copy-pasting their tweets, then surrounding them with quotation marks:

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">&quot;<a href="https://twitter.com/trumplican2016">\@trumplican2016</a>: <a href="https://twitter.com/realDonaldTrump">\@realDonaldTrump</a> <a href="https://twitter.com/DavidWohl">\@DavidWohl</a> stay the course mr trump your message is resonating with the PEOPLE&quot;</p>&mdash; Donald J. Trump (\@realDonaldTrump) <a href="https://twitter.com/realDonaldTrump/status/758512401629192192">July 28, 2016</a></blockquote>
<script async src="http://platform.twitter.com/widgets.js" charset="utf-8"></script>

Almost all of these quoted tweets are posted from the Android:

```{r dependson = "tweets"}
tweets %>%
  count(source,
        quoted = ifelse(str_detect(text, '^"'), "Quoted", "Not quoted")) %>%
  ggplot(aes(source, n, fill = quoted)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "", y = "Number of tweets", fill = "") +
  ggtitle('Whether tweets start with a quotation mark (")')
```

## Comparison of words

Now that we're sure there's a difference between these two accounts, what can we say about the difference in the *content*? We can use the [`tidytext`](https://cran.r-project.org/web/packages/tidytext) package to analyze this.

We start by dividing into individual words using the `unnest_tokens` function, and removing some common "stopwords". This is a common aspect to preparing text for analysis. Typically, *tokens* are single words from a document. However they can also be *bi-grams* (pairs of words), *tri-grams* (three-word sequences), *n-grams* ($n$-length sequences of words), or in this case, individual words, hashtags, or references to other Twitter users. Because tweets are a special form of text (they can include words, urls, references to other users, hashtags, etc.) we need to use a custom tokenizing function to convert the text into tokens.

```{r tweet_words, dependson = "tweets"}
library(tidytext)

reg <- "([^A-Za-z\\d#@']|'(?![A-Za-z\\d#@]))"   # custom regular expression to tokenize tweets

# function to neatly print the first 10 rows using kable
print_neat <- function(df){
  df %>%
    head() %>%
    knitr::kable()
}

# tweets data frame
tweets %>%
  print_neat()

# remove manual retweets
tweets %>%
  filter(!str_detect(text, '^"')) %>%
  print_neat()

# remove urls
tweets %>%
  filter(!str_detect(text, '^"')) %>%
  mutate(text = str_replace_all(text, "https://t.co/[A-Za-z\\d]+|&amp;", "")) %>%
  print_neat()

# unnest into tokens - tidytext format
tweets %>%
  filter(!str_detect(text, '^"')) %>%
  mutate(text = str_replace_all(text, "https://t.co/[A-Za-z\\d]+|&amp;", "")) %>%
  unnest_tokens(word, text, token = "regex", pattern = reg) %>%
  print_neat()

# remove stop words
tweets %>%
  filter(!str_detect(text, '^"')) %>%
  mutate(text = str_replace_all(text, "https://t.co/[A-Za-z\\d]+|&amp;", "")) %>%
  unnest_tokens(word, text, token = "regex", pattern = reg) %>%
  filter(!word %in% stop_words$word,
         str_detect(word, "[a-z]")) %>%
  print_neat()

# store for future use
tweet_words <- tweets %>%
  filter(!str_detect(text, '^"')) %>%
  mutate(text = str_replace_all(text, "https://t.co/[A-Za-z\\d]+|&amp;", "")) %>%
  unnest_tokens(word, text, token = "regex", pattern = reg) %>%
  filter(!word %in% stop_words$word,
         str_detect(word, "[a-z]"))
```

What were the most common words in Trump's tweets overall?

```{r tweet_words_plot, dependson = "tweet_words", fig.height = 6, fig.width = 8, echo = FALSE}
tweet_words %>%
  count(word, sort = TRUE) %>%
  head(20) %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(word, n)) +
  geom_bar(stat = "identity") +
  labs(title = "Frequency of tokens in @realDonaldTrump timeline",
         x = "Word",
         y = "Occurrences") +
  coord_flip()
```

Yeah, sounds about right.

### Assessing word and document frequency

One measure of how important a word may be is its *term frequency* (tf), how frequently a word occurs within a document. The problem with this approach is that some words occur many times in a document, yet are probably not important (e.g. “the”, “is”, “of”). Instead, we want a way of downweighting words that are common across all documents, and upweighting words that are frequent within a small set of documents.

Another approach is to look at a term's *inverse document frequency* (idf), which decreases the weight for commonly used words and increases the weight for words that are not used very much in a collection of documents. This can be combined with term frequency to calculate a term's *tf-idf*, the frequency of a term adjusted for how rarely it is used. It is intended to measure how important a word is to a document in a collection (or corpus) of documents. It is a rule-of-thumb or heuristic quantity, not a theoretically proven method. The inverse document frequency for any given term is defined as

$$idf(\text{term}) = \ln{\left(\frac{n_{\text{documents}}}{n_{\text{documents containing term}}}\right)}$$

To calculate *tf-idf* for this set of documents, we will pool all the tweets from iPhone and Android together and treat them as if they are two total documents. Then we can calculate the frequency of terms in each group, and standardize that relative to the the term's frequency across the entire corpus.

```{r}
tweet_words_count <- tweet_words %>%
  count(source, word, sort = TRUE) %>%
  ungroup()
tweet_words_count

total_words <- tweet_words_count %>%
  group_by(source) %>%
  summarize(total = sum(n))
total_words

tweet_words_count <- left_join(tweet_words_count, total_words)
tweet_words_count

tweet_words_count <- tweet_words_count %>%
  bind_tf_idf(word, source, n)
tweet_words_count
```

Which terms have a high tf-idf?

```{r}
tweet_words_count %>%
  select(-total) %>%
  arrange(desc(tf_idf))
```

```{r}
tweet_important <- tweet_words_count %>%
  arrange(desc(tf_idf)) %>%
  mutate(word = factor(word, levels = rev(unique(word))))

tweet_important %>%
  group_by(source) %>%
  slice(1:15) %>%
  ggplot(aes(word, tf_idf, fill = source)) +
  geom_bar(alpha = 0.8, stat = "identity") +
  labs(title = "Highest tf-idf words in @realDonaldTrump",
       subtitle = "Top 15 for Android and iPhone",
       x = NULL, y = "tf-idf") +
  coord_flip()
```

* **Most hashtags come from the iPhone**. Indeed, almost no tweets from Trump's Android contained hashtags, with some rare exceptions like [this one](https://twitter.com/realDonaldTrump/status/753960134422900736). (This is true only because we filtered out the quoted "retweets", as Trump does sometimes quote tweets [like this](https://twitter.com/realDonaldTrump/status/731805331425218560) that contain hashtags).

* **Words like "join", and times like "7pm", also came only from the iPhone**. The iPhone is clearly responsible for event announcements like [this one](https://twitter.com/realDonaldTrump/status/743522630230228993) ("Join me in Houston, Texas tomorrow night at 7pm!")

* **A lot of "emotionally charged" words, like "badly" and "dumb", were overwhelmingly more common on Android.** This supports the original hypothesis that this is the "angrier" or more hyperbolic account.

### Sentiment analysis

Since we've observed a difference in sentiment between the Android and iPhone tweets, let's try quantifying it. We'll work with the [NRC Word-Emotion Association](http://saifmohammad.com/WebPages/NRC-Emotion-Lexicon.htm) lexicon, available from the `tidytext` package, which associates words with 10 sentiments: **positive**, **negative**, **anger**, **anticipation**, **disgust**, **fear**, **joy**, **sadness**, **surprise**, and **trust**.

```{r nrc}
nrc <- sentiments %>%
  filter(lexicon == "nrc") %>%
  select(word, sentiment)
nrc
```

To measure the sentiment of the Android and iPhone tweets, we can count the number of words in each category:

```{r by_source_sentiment}
sources <- tweet_words %>%
  group_by(source) %>%
  mutate(total_words = n()) %>%
  ungroup() %>%
  distinct(id, source, total_words)
sources

by_source_sentiment <- tweet_words %>%
  inner_join(nrc, by = "word") %>%
  count(sentiment, id) %>%
  ungroup() %>%
  complete(sentiment, id, fill = list(n = 0)) %>%
  inner_join(sources) %>%
  group_by(source, sentiment, total_words) %>%
  summarize(words = sum(n)) %>%
  ungroup()

head(by_source_sentiment)
```

(For example, we see that `r by_source_sentiment$words[1]` of the `r by_source_sentiment$total_words[1]` words in the Android tweets were associated with "anger"). We then want to measure how much more likely the Android account is to use an emotionally-charged term relative to the iPhone account. Since this is count data, we can use a [Poisson test](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/poisson.test.html) to measure the difference:

```{r}
# function to calculate the poisson.test for a given sentiment
poisson_test <- function(df){
  poisson.test(df$words, df$total_words)
}

# use the nest() and map() functions to apply poisson_test to each sentiment and 
# extract results using broom::tidy()
sentiment_differences <- by_source_sentiment %>%
  group_by(sentiment) %>%
  nest() %>%
  mutate(poisson = map(data, poisson_test),
         poisson_tidy = map(poisson, tidy)) %>%
  unnest(poisson_tidy, .drop = TRUE)
sentiment_differences
```

And we can visualize it with a 95% confidence interval:

```{r}
sentiment_differences %>%
  ungroup() %>%
  mutate(sentiment = reorder(sentiment, estimate)) %>%
  mutate_each(funs(. - 1), estimate, conf.low, conf.high) %>%
  ggplot(aes(estimate, sentiment)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  scale_x_continuous(labels = percent_format()) +
  labs(x = "% increase in Android relative to iPhone",
       y = "Sentiment")
```

Thus, Trump's Android account uses about 40-80% more words related to **disgust**, **sadness**, **fear**, **anger**, and other "negative" sentiments than the iPhone account does. (The positive emotions weren't different to a statistically significant extent).

We're especially interested in which words drove this different in sentiment. Let's consider the words with the largest changes within each category:

```{r, fig.height = 10}
tweet_important %>%
  inner_join(nrc, by = "word") %>%
  filter(!sentiment %in% c("positive", "negative")) %>%
  mutate(sentiment = reorder(sentiment, -tf_idf),
         word = reorder(word, -tf_idf)) %>%
  group_by(sentiment) %>%
  top_n(10, tf_idf) %>%
  ungroup() %>%
  ggplot(aes(word, tf_idf, fill = source)) +
  facet_wrap(~ sentiment, scales = "free", nrow = 4) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "",
       y = "tf-idf") +
  scale_fill_manual(name = "", labels = c("Android", "iPhone"),
                    values = c("red", "lightblue"))
```

This confirms that lots of words annotated as negative sentiments are more common in Trump's Android tweets than the campaign's iPhone tweets. It's no wonder [Trump's staff took away his tweeting privileges for the remainder of the campaign](http://www.nytimes.com/2016/11/07/us/politics/donald-trump-presidential-race.html).

# Acknowledgments {.toc-ignore}

```{r child='_ack_stat545.Rmd'}
```
* This page is derived in part from ["Text analysis of Trump's tweets confirms he writes only the (angrier) Android half"](http://varianceexplained.org/r/trump-tweets/) and licensed under a [Creative Commons Attribution-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-sa/4.0/).
* This page is derived in part from ["Tidy Text Mining with R"](http://tidytextmining.com/) and licensed under a [Creative Commons Attribution-NonCommercial-ShareAlike 3.0 United States License](https://creativecommons.org/licenses/by-nc-sa/3.0/us/).

# Session Info {.toc-ignore}

```{r sessioninfo}
devtools::session_info()
```
