---
title: "Practice getting data from the Twitter API"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
library(broom)

set.seed(1234)
theme_set(theme_minimal())
```

There are several packages for R for accessing and searching Twitter. Twitter actually has two separate APIs:

1. The **REST** API - this allows you programmatic access to read and write Twitter data. For research purposes, this allows you to search the recent history of tweets and look up specific users.
1. The **Streaming** API - this allows you to access the public data flowing through Twitter in real-time. It requires your R session to be running continuously, but allows you to capture a much larger sample of tweets while avoiding rate limits for the REST API.

Popular packages for the Twitter API in R include:

* [`twitteR`](https://cran.rstudio.com/web/packages/twitteR/index.html) is the most popular package for R, but it only allows you to access the REST API. It is also deprecated (not actively updated), in lieu of a new up-and-coming package (identified below)
* [`streamR`](https://cran.rstudio.com/web/packages/streamR/index.html) is more complicated, but allows you to query the Streaming API from R. It is ancient in computational terms (last updated in January 2014), but does what it needs to do.
* [`rtweet`](http://rtweet.info/) is a relatively recent addition to the R package universe that allows you to access both the REST and streaming APIs.

# Using `rtweet`

Here, we are going to practice using the `rtweet` package to search Twitter.

```r
install.packages("rtweet")
```

```{r twitter}
library(rtweet)
```

## OAuth authentication

OAuth is an open standard for authorization, commonly used as a way for Internet users to authorize websites or applications to access their information on other websites but without giving them the passwords. This still requires an API key, but is a bit more complicated to setup.

1. The first step is to create a Twitter application for yourself. In order to do this, you do need to have a registered Twitter account. Go to (https://apps.twitter.com/app/new) and log in. After filling in the basic info (make sure to set the "Callback URL" to "http://127.0.0.1:1410"), go to the **Permissions** tab and select "Read, Write and Access direct messages". Make sure to click on the save button after doing this. In the **Details** tab, take note of your consumer key and consumer secret.

1. Store your API key and secret using the `.Rprofile` method. Edit `.Rprofile` and add

    ```r
    options(twitter_app_name = "Your app name")
    options(twitter_api_key = "Your API key")
    options(twitter_api_secret = "Your API secret")
    ```

    Restart RStudio to get the keys loaded into your environment.
    
1. Run from the console:

    ```r
    twitter_token <- create_token(
      app = getOption("twitter_app_name"),
      consumer_key = getOption("twitter_api_key"),
      consumer_secret = getOption("twitter_api_secret"))
    ```
    
    This will open your browser and take you to a page to authorize your app to access your Twitter account. You need this in order to search any content on Twitter.
    
1. At this point you should get a message back in RStudio "Authentication complete." You're done setting up the authentication for `rtweet` - you'd just need to set the `token` argument equal to `twitter_token` (the token object we just created).

### Storing Twitter token

Rather than creating a token every time you open R, remembering where you saved your token(s) or passwords, and specifying a token every time you use a function, users are encouraged to save tokens as an environment variable. Since `twitter_token` is no longer a character string, we cannot use the `.Rprofile` method from before. Instead, we can store the token as an **environment variable**, or an object stored on your computer separately from your project directory.

To save a personal access token as an environment variable, complete the following steps:

1. Use `readr::write_rds()` to save `twitter_token` to your home directory. The code below should locate and construct the path to your home directory for you. Assuming you've saved your token as `twitter_token`, the final line in the code below will save your token for you as well.

    ```r
    # path of home directory
    home_directory <- path.expand("~/")

    # combine with name for token
    file_name <- file.path(home_directory, "twitter_token.rds")

    # save token to home directory
    readr::write_rds(twitter_token, path = file_name)
    ```

1. Create a plain text file containing the path to your token object and save it to your home directory as `.Renviron`.
    
    To create a plain text file in R, modify the code below. Change `TWITTER_PAT` location to match the path you used earlier (in the example below, you'd want to change "/Users/soltoffbc/twitter_token.rds"). You can also create a plain text document in any text editor like TextEdit or Notepad. If you're using Rstudio, select File > New File > Text File.

    **Important**: Like with `.Rprofile`, make sure the last line of `.Renviron` is blank. I achieved this in the code below by including `fill = TRUE` in the `cat()` function.

    ```r
    # On my mac, the .Renviron text looks like this:
    #     TWITTER_PAT=/Users/soltoffbc/twitter_token.rds

    # Assuming you followed the procedures to create "file_name"
    # from the previous code chunk, then the code below should
    # create and save your environment variable.

    cat(paste0("TWITTER_PAT=", file_name),
        file = file.path(home_directory, ".Renviron"),
        append = TRUE)
    ```

1. Restart R so it can process the environment variable.

1. If you completed the steps successfully, you should be able to use `rtweet` functions without specifying [or even *thinking* about] tokens [ever again].

## Searching tweets

To find 3000 recent tweets using the "rstats" hashtag:

```{r twitter-rstats}
rt <- search_tweets(
  q = "#rstats",
  n = 3000,
  include_rts = FALSE
)
rt
```

* `q` - the search query
* `n` - maximum number of tweets to be returned
* `include_rts = FALSE` - exclude retweets generated by Twitter's built-in "retweet" function. We only want original tweets.

The resulting object is a `tibble` data frame with one row for each tweet. The data frame contains the full text of the tweet (`text`), the username of the poster (`screen_name`), as well as a wealth of metadata.

Note that the Twitter REST API limits all searches to the past 6-9 days. **You will not retrieve any earlier results.**

## Searching users

Use `get_timeline()` or `get_timelines()` to retrieve tweets from one or more specified Twitter users. This only works for users with public profiles or those that have authorized your app.

```{r twitter-count}
countvoncount <- get_timeline(user = "countvoncount", n = 1000)
countvoncount
```

With `get_timelines()`, you are not limited to only the most recent 6-9 days of tweets.

## Visualizing tweets

Because the resulting objects are data frames, you can perform standard data transformation, summarization, and visualization on the underlying data.

`rtweet` includes the `ts_plot()` function which automates some common time series visualization methods. For example, we can quickly visualize the frequency of `#rstats` tweets:

```{r rstats-freq}
ts_plot(rt, by = "3 hours")
```

The `by` argument allows us to aggregate over different lengths of time.

```{r rstats-freq-day}
ts_plot(rt, by = "1 hours")
```

And because `ts_plot()` uses `ggplot2`, we can modify the graphs using familiar `ggplot2` functions:

```{r rstats-freq-clean}
ts_plot(rt, by = "3 hours") +
  theme(plot.title = element_text(face = "bold")) +
  labs(
    x = NULL, y = NULL,
    title = "Frequency of #rstats Twitter statuses from past 9 days",
    subtitle = "Twitter status (tweet) counts aggregated using three-hour intervals",
    caption = "\nSource: Data collected from Twitter's REST API via rtweet"
  )
```

# Exercise: Practice using `rtweet`

1. Create a new R project on your computer. You can use Git or not - it is just for practice in class today
1. Setup your API key with a Twitter app
1. Authenticate using the `rtweet` package in R
1. Find the 1000 most recent tweets by [Katy Perry](https://twitter.com/katyperry), [Kim Kardashian West](https://twitter.com/KimKardashian), and [Ariana Grande](https://twitter.com/ArianaGrande).
1. Visualize their tweet frequency by week. Who posts most often? Who posts least often?

<details> 
  <summary>Click for the solution</summary>
  <p>

```{r twitter-popstars}
popstars <- get_timelines(
  user = c("katyperry", "KimKardashian", "ArianaGrande"),
  n = 1000
)

popstars %>%
  group_by(screen_name) %>%
  ts_plot(by = "week")
```
    
  </p>
</details>

# Acknowledgments {.toc-ignore}

```{r child='_ack_stat545.Rmd'}
```
* OAuth token storage derived from ["Obtaining and using access tokens"](http://rtweet.info/articles/auth.html).

# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```
