---
title: "Practice getting data from the Twitter API"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
library(broom)

set.seed(1234)
theme_set(theme_minimal())
```

There are several packages for R for accessing and searching Twitter. Twitter actually has two separate APIs:

1. The **REST** API - this allows you programmatic access to read and write Twitter data. For research purposes, this allows you to search the recent history of tweets and look up specific users.
1. The **Streaming** API - this allows you to access the public data flowing through Twitter in real-time. It requires your R session to be running continuously, but allows you to capture a much larger sample of tweets while avoiding rate limits for the REST API.

* [`twitteR`](https://cran.rstudio.com/web/packages/twitteR/index.html) is the most popular package for R, but it only allows you to access the REST API.
* [`streamR`](https://cran.rstudio.com/web/packages/streamR/index.html) is more complicated, but allows you to query the Streaming API from R.

# Using `twitteR`

Here, we are going to practice using the `twitteR` package to search Twitter.

```r
install.packages("twitteR")
```

```{r twitter}
library(twitteR)
```

## OAuth authentication

OAuth is an open standard for authorization, commonly used as a way for Internet users to authorize websites or applications to access their information on other websites but without giving them the passwords. This still requires an API key, but is a bit more complicated to setup.

1. The first step is to create a Twitter application for yourself. In order to do this, you do need to have a registered Twitter account. Go to (https://apps.twitter.com/app/new) and log in. After filling in the basic info (make sure to set the "Callback URL" to "http://127.0.0.1:1410"), go to the **Permissions** tab and select "Read, Write and Access direct messages". Make sure to click on the save button after doing this. In the **Details** tab, take note of your consumer key and consumer secret.

1. Store your API key and token using the `.Rprofile` method. Edit `.Rprofile` and add

    ```r
    options(twitter_api_key = "Your API key")
    options(twitter_api_token = "Your API secret")
    ```

    Restart RStudio to get the keys loaded into your environment.
    
1. Run from the console:

    ```r
    setup_twitter_oauth(consumer_key = getOption("twitter_api_key"),
                        consumer_secret = getOption("twitter_api_token"))
    ```
    
    This will open your browser and take you to a page to authorize your app to access your Twitter account. You need this in order to search any content on Twitter.
    
1. At this point you should get a message back in RStudio "Authentication complete." You're done setting up the authentication for `twitteR`.

**You must do this in order to query Twitter from within a non-interactive session (i.e. an R Markdown document or using `source()` to run a script).** If you do not, you will get an error because R will not be able to finish the authentication process without your input.

## Searching tweets

```{r twitter-setup, echo = FALSE}
setup_twitter_oauth(consumer_key = getOption("twitter_api_key"),
                    consumer_secret = getOption("twitter_api_token"))
```

```{r twitter-rstats}
tweets <- searchTwitter('#rstats', n = 5)
tweets
```

## Searching users

Use `getUser()` which returns a `user` object that you can perform additional functions on. This only works for users with public profiles or those that have authorized your app.

```{r twitter-clinton}
clinton <- getUser("hillaryclinton")
clinton$getDescription()
clinton$getFriends(n = 5)
```

## Tidying tweets

By default `twitteR` returns tweets in complex lists.

```{r twitter-list}
str(tweets)
```

To get the data into a data frame, use

```{r twitter-df}
df <- twListToDF(tweets) %>%
  as_tibble()
df
```

# Exercise: Practice using `twitteR`

1. Create a new R project on your computer. You can use Git or not - it is just for practice in class today
1. Setup your API key with a Twitter app
1. Authenticate using the `twitteR` package in R
1. Find the 50 most recent tweets by [President Donald Trump](https://twitter.com/realDonaldTrump) and store them in a data frame
    * `userTimeline()` can be used to retrieve tweets from individual users
    * `searchTwitter()` finds tweets from any public account that references the username

<details> 
  <summary>Click for the solution</summary>
  <p>

```{r twitter-donald}
setup_twitter_oauth(consumer_key = getOption("twitter_api_key"),
                    consumer_secret = getOption("twitter_api_token"))

trump <- userTimeline("realDonaldTrump", n = 50)

trump_df <- twListToDF(trump) %>%
  as_tibble()
trump_df
```
    
  </p>
</details>

# Acknowledgments {.toc-ignore}

```{r child='_ack_stat545.Rmd'}
```

# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```
