---
title: "Practice drawing raster maps"
date: 2019-03-01

type: book
toc: true
draft: false
aliases: ["/notes/raster-maps-practice/"]
categories: ["dataviz", "geospatial"]

weight: 52
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
library(ggmap)
library(here)

options(digits = 3)
set.seed(123)
theme_set(theme_minimal())
```

{{% callout note %}}

Run the code below in your console to download this exercise as a set of R scripts.

```r
usethis::use_course("cis-ds/visualize-spatial-i")
```

{{% /callout %}}

## New York City 311 data

New York City has [an excellent data portal](https://opendata.cityofnewyork.us/) publishing a large volume of public records. Here we'll look at a subset of the [311 service requests](https://data.cityofnewyork.us/Social-Services/311-Service-Requests-from-2010-to-Present/erm2-nwe9). I used `RSocrata` and the data portal's [API](/notes/application-program-interface/) to retrieve a portion of the data set.

{{% callout note %}}

If you are copying-and-pasting code from this demonstration, use `nyc_311 <- read_csv("https://info5940.infosci.cornell.edu/data/nyc-311.csv")` to download the file from the course website.

{{% /callout %}}

```{r nyc-311, include = FALSE}
nyc_311 <- read_csv(file = here("static", "data", "nyc-311.csv"))
```

```{r nyc-311-glimpse, dependson = "nyc-311"}
glimpse(nyc_311)
```

## Exercise: Visualize the 311 data

1. Obtain map tiles using `ggmap` for New York City.

    {{< spoiler text="Click for the solution" >}}

```{r bb-nyc}
# store bounding box coordinates
nyc_bb <- c(
  left = -74.263045,
  bottom = 40.487652,
  right = -73.675963,
  top = 40.934743
)

nyc <- get_stamenmap(
  bbox = nyc_bb,
  zoom = 11
)

# plot the raster map
ggmap(nyc)
```

    {{< /spoiler >}}

1. Generate a scatterplot of complaints about sidewalk conditions.

    {{< spoiler text="Click for the solution" >}}

```{r potholes-point, dependson = c("nyc-311", "bb-chicago")}
# initialize map
ggmap(nyc) +
  # add layer with scatterplot
  # use alpha to show density of points
  geom_point(
    data = filter(nyc_311, complaint_type == "Sidewalk Condition"),
    mapping = aes(
      x = longitude,
      y = latitude
    ),
    size = .25,
    alpha = .05
  )
```

    {{< /spoiler >}}

1. Generate a heatmap of complaints about sidewalk conditions. Do you see any unusual patterns or clusterings?

    {{< spoiler text="Click for the solution" >}}

```{r potholes-heatmap, dependson = c("nyc-311", "bb-chicago")}
# initialize the map
ggmap(nyc) +
  # add the heatmap
  stat_density_2d(
    data = filter(nyc_311, complaint_type == "Sidewalk Condition"),
    mapping = aes(
      x = longitude,
      y = latitude,
      fill = stat(level)
    ),
    alpha = .1,
    bins = 50,
    geom = "polygon"
  )
```

    Seems to be clustered most dense in Manhattan and Brooklyn. Makes sense because they receive significant pedestrian traffic.

    {{< /spoiler >}}

1. Obtain map tiles for Roosevelt Island.

    {{< spoiler text="Click for the solution" >}}

```{r bb-roosevelt}
roosevelt_bb <- c(
  left = -73.967121,
  bottom = 40.748700,
  right = -73.937080,
  top = 40.774704
)
roosevelt <- get_stamenmap(
  bbox = roosevelt_bb,
  zoom = 14
)

# plot the raster map
ggmap(roosevelt)
```

    {{< /spoiler >}}

1. Generate a scatterplot of food poisoning complaints.

    {{< spoiler text="Click for the solution" >}}

```{r dead-animals-point, dependson = c("nyc-311", "bb-roosevelt")}
# initialize the map
ggmap(roosevelt) +
  # add a scatterplot layer
  geom_point(
    data = filter(nyc_311, complaint_type == "Food Poisoning"),
    mapping = aes(
      x = longitude,
      y = latitude
    ),
    alpha = 0.2
  )
```

    {{< /spoiler >}}

### Session Info

```{r child = here::here("R", "_session-info.Rmd")}
```
