---
title: "R startup procedures"
date: 2019-03-01

type: docs
toc: true
draft: false
categories: ["project-management"]

menu:
  notes:
    parent: Project management
    weight: 5
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)

set.seed(1234)
theme_set(theme_minimal())
```

R offers the ability to customize its startup procedures. That is, every time an R process is initiated, users can define custom R code to evaluate before the session begins. This custom code is generally stored in one of two locations:

1. [`.Renviron`](#renviron) - which contains environment variables to be set in R sessions.
2. [`.Rprofile`](#rprofile) - which contains R code to be run in each session.

## `.Renviron`

The `.Renviron` file is most useful for defining sensitive information such as [API keys](/notes/application-program-interface/#api-authentication) (such as GitHub or Twitter) as well as R specific environment variables like the history size (`R_HISTSIZE=100000`) and default library locations `R_LIBS_USER`.

The `.Renviron` file contains lists of environment variables to set. This is _not_ R code, it uses a format similar to that used on the command line shell.

The easiest way to edit `.Renviron` is by running `usethis::edit_r_environ()`.

A simple example of a `.Renviron` file is

```shell
R_HISTSIZE=100000
GITHUB_PAT=abc123
R_LIBS_USER=~/R/%p/%v
```

You will not typically need to edit this file manually. For example, `rtweet` [automatically stores your Twitter API authentication information in `.Renviron`](/notes/twitter-api-practice/#oauth-authentication).

## `.Rprofile`

The `.Rprofile` file contains R code to be run when R starts up. It is run after the `.Renviron` file is sourced. You generally have only a single `.Renviron` file on your computer. However you can have multiple `.Rprofile` files stored on a single computer.

* Typically `.Rprofile` is located in the users' home directory (`~/.Rprofile` on MacOS/Linux). This `.Rprofile` is run whenever you start up R.
* Each R Project folder can also contain a separate `.Rprofile`. This `.Rprofile` is specific to that project and will only be run if you open that R Project in RStudio.

The easiest way to edit `.Rprofile` is by running `usethis::edit_r_profile()`.

Some common things people often add to their `.RProfile`

1. Set a default CRAN mirror
1. Write a welcome message
1. Customize their R prompt
1. Change options, screen width, numeric display
1. Store API keys/tokens that are necessary for only a single project

### Git tracking of `.Rprofile`

You can commit `.Rprofile` inside of a Git repository, however you should only do this for a private repository or if `.Rprofile` does not contain API keys. If you commit `.Rprofile` to a public repo, anybody can access and view the file. If it contains an API key, anyone can use it to access that API (which is typically a violation of your user agreement, not to mention if it is a paid API you will be charged for these queries).

## Reproducibility

A good rule of thumb is you should only put things in your `.Rprofile` that you run interactively in the R terminal. If it ever appears in a R script or R Markdown file it should **not** be in your `.Rprofile`.

If you set these options in your `.Rprofile`, then try to run one of your scripts on another system without your `.Rprofile` it will no longer be reproducible. Some problematic examples are loading packages used in analysis (such as `dplyr` or `ggplot2`) or changing default options which change the _value_ of outputs, such as `options(stringsAsFactors = FALSE)`. 

In addition because the `.Rprofile` is run by **every** R process (including those started by R itself) it is important to guard most of the code with `interactive()`, so it is only run in interactive sessions (sessions you are controlling with a terminal).

A simple example of a `.Rprofile` is

```r
options(repos = c(CRAN = "https://cran.rstudio.org"))
if (interactive()) {
  options(width = 120)
}
```

## Acknowledgments

* Substantial material drawn from [What They Forgot To Teach You About R](https://whattheyforgot.org/) by Jenny Bryan and Jim Hester. Licensed under the licensed under the [CC BY-SA 4.0 Creative Commons License](https://creativecommons.org/licenses/by-sa/4.0/).

## Session Info

```{r child = here::here("R", "_session-info.Rmd")}
```
