---
title: "Practice transforming and visualizing factors"
date: 2019-03-01

type: docs
toc: true
draft: false
aliases: ["/datawrangle_factors_exercise.html"]
categories: ["datawrangle"]

menu:
  notes:
    parent: Data wrangling
    weight: 6
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
library(rcfss)
theme_set(theme_minimal())
```

{{% callout note %}}

Run the code below in your console to download this exercise as a set of R scripts.

```r
usethis::use_course("uc-cfss/data-wrangling-relational-data-and-factors")
```

{{% /callout %}}

```{r load-data}
# load the data
data("gun_deaths")
gun_deaths
```

## Convert `month` into a factor column

{{< spoiler text="Click for the solution" >}}

```{r month-factor}
# create a character vector with all month values
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)

# or use the built-in constant
month.abb

# use mutate() and factor() to convert the column and store the result
(gun_deaths <- gun_deaths %>%
  mutate(month = factor(month,
                        levels = month_levels)))
```

{{< /spoiler >}}

## Visualize the total gun deaths per month, in chronological order

{{< spoiler text="Click for the solution" >}}

```{r month-deaths}
ggplot(data = gun_deaths,
       mapping = aes(x = month)) +
  geom_bar() +
  labs(title = "Gun Deaths in the United States (2012-2014)",
       x = "Month",
       y = "Number of gun deaths")
```

{{< /spoiler >}}

## Visualize the total gun deaths per month, sorted from lowest to highest

{{< spoiler text="Click for the solution" >}}

```{r month-deaths-sort}
# with geom_col() and fct_reorder()
gun_deaths %>%
  count(month) %>%
  ggplot(mapping = aes(x = fct_reorder(.f = month, .x = n), y = n)) +
  geom_col() +
  labs(title = "Gun Deaths in the United States (2012-2014)",
       x = "Month",
       y = "Number of gun deaths")

# with geom_bar() and fct_infreq()
ggplot(data = gun_deaths,
       mapping = aes(x = month %>%
                       fct_infreq() %>%
                       fct_rev())) +
  geom_bar() +
  labs(title = "Gun Deaths in the United States (2012-2014)",
       x = "Month",
       y = "Number of gun deaths")
```

{{< /spoiler >}}

## Visualize the frequency of intent of gun deaths using a bar chart, sorted from most to least frequent

{{< spoiler text="Click for the solution" >}}

```{r intent}
# identify all possible types of intent
intent_levels <- c("Accidental", "Homicide", "Suicide", "Undetermined")

gun_deaths %>%
  # remove rows with missing intent values
  drop_na(intent) %>%
  # parse_factor() is a tidyverse friendly form of factor()
  mutate(intent = parse_factor(intent, levels = intent_levels)) %>%
  ggplot(mapping = aes(x = fct_infreq(intent))) +
  geom_bar() +
  labs(title = "Gun Deaths in the United States (2012-2014)",
       x = "Intent of death",
       y = "Number of gun deaths")
```

{{< /spoiler >}}

## Visualize total gun deaths by season of the year using a bar chart.

Hint: do not use `cut()` to create the `season` column.

{{< spoiler text="Click for the solution" >}}

```{r season}
gun_deaths %>%
  # use fct_collapse() to condense into 4 categories
  mutate(season = fct_collapse(month,
                               "Winter" = c("Jan", "Feb", "Mar"),
                               "Spring" = c("Apr", "May", "Jun"),
                               "Summer" = c("Jul", "Aug", "Sep"),
                               "Fall" = c("Oct", "Nov", "Dec"))) %>%
  ggplot(mapping = aes(x = season)) +
  geom_bar() +
  labs(title = "Gun Deaths in the United States (2012-2014)",
       x = "Season",
       y = "Number of gun deaths")
```

{{< /spoiler >}}

## Session Info

```{r child = here::here("R", "_session-info.Rmd")}
```
