---
title: "Tidy evaluation"
author: "INFO 5940 <br /> Cornell University"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: r
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
# generate CSS file
library(xaringanthemer)
style_duo_accent(
  primary_color = "#B31B1B",
  secondary_color = "#F8981D",
  inverse_header_color = "#222222",
  black_color = "#222222",
  header_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  text_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  code_font_google = xaringanthemer::google_font("Source Code Pro"),
  base_font_size = "24px",
  code_font_size = "20px",
  # title_slide_background_image = "https://github.com/uc-dataviz/course-notes/raw/main/images/hexsticker.svg",
  # title_slide_background_size = "contain",
  # title_slide_background_position = "top",
  header_h1_font_size = "2rem",
  header_h2_font_size = "1.75rem",
  header_h3_font_size = "1.5rem",
  extra_css = list(
    "h1" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h2" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h3" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    ".tiny" = list("font-size" = "70%"),
    ".small" = list("font-size" = "90%"),
    ".midi" = list("font-size" = "150%"),
    ".tiny .remark-code" = list("font-size" = "70%"),
    ".small .remark-code" = list("font-size" = "90%"),
    ".midi .remark-code" = list("font-size" = "150%"),
    ".large" = list("font-size" = "200%"),
    ".xlarge" = list("font-size" = "600%"),
    ".huge" = list(
      "font-size" = "400%",
      "font-family" = "'Montserrat', sans-serif",
      "font-weight" = "bold"
    ),
    ".hand" = list(
      "font-family" = "'Gochi Hand', cursive",
      "font-size" = "125%"
    ),
    ".task" = list(
      "padding-right" = "10px",
      "padding-left" = "10px",
      "padding-top" = "3px",
      "padding-bottom" = "3px",
      "margin-bottom" = "6px",
      "margin-top" = "6px",
      "border-left" = "solid 5px #F1DE67",
      "background-color" = "#F3D03E"
    ),
    ".pull-left" = list(
      "width" = "49%",
      "float" = "left"
    ),
    ".pull-right" = list(
      "width" = "49%",
      "float" = "right"
    ),
    ".pull-left-wide" = list(
      "width" = "70%",
      "float" = "left"
    ),
    ".pull-right-narrow" = list(
      "width" = "27%",
      "float" = "right"
    ),
    ".pull-left-narrow" = list(
      "width" = "27%",
      "float" = "left"
    ),
    ".pull-right-wide" = list(
      "width" = "70%",
      "float" = "right"
    ),
    ".blue" = list(color = "#2A9BB7"),
    ".purple" = list(color = "#a493ba"),
    ".yellow" = list(color = "#f1de67"),
    ".gray" = list(color = "#222222")
  )
)

source(here::here("R", "slide-opts.R"))
```

```{r pkgs, include = FALSE, cache = FALSE}
library(tidyverse)
library(palmerpenguins)
library(here)
library(countdown)

set.seed(1234)
theme_set(theme_minimal(base_size = rcis::base_size))
```

class: inverse, middle

# Motivation for tidy eval

---

## Motivation for tidy eval

Tidy eval (and non-standard evaluation generally in R) exists so that we can refer to data columns using bare names.

--

```r
# base R with standard evaluation
penguins[penguins[["species"]] == "Adelie", ]
```

```r
# dplyr with tidy evaluation
dplyr::filter(.data = penguins, species == "Adelie")
```

--

 - It makes things easier if you are working interactively.
 - It makes things more interesting if you are writing functions. 

---

## Families of tidy-eval functions

There are two families of tidy-eval functions:

- **data-masking**: evaluate variables in context of data frame

    ```r
    penguins %>%
      dplyr::mutate(body_mass_kg = body_mass_g / 1000)
    ```
    
--


- **tidy-select**: specify columns within a data frame
 
    ```r
    penguins %>%
     dplyr::select(starts_with("bill"))
    ```

---

## Ways to express tidy-eval 

- using bare names, what we do in interactive programming:

    ```r
    penguins %>%
      dplyr::mutate(body_mass_kg = body_mass_g / 1000)
    ```

--


- using strings, useful for Shiny :

    ```r
    penguins %>%
      dplyr::mutate("body_mass_kg" := .data[["body_mass_g"]] / 1000)
    ```

---

## Goal

Write functions that:

- Call `tidyverse` functions, e.g. `dplyr::filter()`
- "Look and feel" like `tidyverse` functions:
   - your users can call using bare names
- Use string-techniques when we know the column names

--

Out of scope:

- Underlying theory of [quasiquotation](https://adv-r.hadley.nz/metaprogramming.html), i.e. `rlang::quo()`, `rlang::enquo()`, `rlang::eval_tidy()` 

---

## Pass the dots

This is the simplest possible solution. 

- If the `tidyverse` function you're using takes `...` as an argument,
- and that's what you want to pass along, 

--

then you can **pass the dots**.

```r
my_select <- function(.data, ...) {
  dplyr::select(.data, ...)
}
```
