---
title: "Working with statistical models"
author: "[MACS 30500](https://info5940.infosci.cornell.edu) <br /> University of Chicago"
output: rcis::xaringan
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  cache = TRUE,
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  fig.retina = 2, fig.width = 12
)

library(tidyverse)
library(broom)
library(gapminder)
library(patchwork)
library(coefplot)
library(margins)
library(here)
library(socviz)
library(rcis)
library(titanic)
library(tidytext)
library(gganimate)

set.seed(1234)
theme_set(theme_minimal(base_size = rcis::base_size))
```

# Working with models

* Visualizing raw data
* Visualizing summarized data
* `stat_*()` functions
* Default `geom_*()` functions

---

# Visualizing models separately

```{r viz-gapminder-combo}
p <- ggplot(
  data = gapminder,
  mapping = aes(x = gdpPercap, y = lifeExp)
) +
  scale_x_log10(labels = scales::dollar) +
  labs(
    x = "GDP per capita",
    y = "Life expectancy"
  ) +
  theme_minimal(base_size = rcis::base_size * .7)

p_robust <- p +
  geom_point(alpha = 0.1) +
  geom_smooth(color = "tomato", fill = "tomato", method = MASS::rlm) +
  geom_smooth(color = "steelblue", fill = "steelblue", method = "lm") +
  labs(subtitle = "OLS vs. robust regression")

p_poly <- p +
  geom_point(alpha = 0.1) +
  geom_smooth(
    color = "tomato", method = "lm", size = 1.2,
    formula = y ~ splines::bs(x, 3), se = FALSE
  ) +
  labs(subtitle = "Polynomial fit")

p_quant <- p +
  geom_point(alpha = 0.1) +
  geom_quantile(
    color = "tomato", size = 1.2, method = "rqss",
    lambda = 1, quantiles = c(0.20, 0.5, 0.85)
  ) +
  labs(subtitle = "Quantile regression")

p_robust +
  p_poly +
  p_quant
```

---

# Combining `ggplot()`s

* How do we combine separate `ggplot()` objects into a single chart?
* `gridExtra`
* `cowplot`
* [`patchwork`](https://patchwork.data-imaginist.com/articles/patchwork.html)

---

# Showing several fits at once

* Mapping variables to the plot
* Instead of a column, use a character string
* [Example using SCOTUS](https://github.com/cfss-fa19/hw03/blob/master/demo/scotus_solution.md#with-two-data-frames)

---

# Showing several fits at once

<div style="width:100%;height:0;padding-bottom:56%;position:relative;"><iframe src="https://giphy.com/embed/8AfVHQbGG8dxGCC7ES" width="100%" height="100%" style="position:absolute" frameBorder="0" class="giphy-embed" allowFullScreen></iframe></div>

---

# Getting model-based graphics right

* Statistics is hard
* Communicating results is hard

--

## General guidelines

* Present your findings in substantive terms
* Show your degree of confidence
* Show your data (when you can)

---

# Coefficient plots

```{r complex-model}
# estimate a model using gun deaths data to compare a results table vs. coefficient plot
gun_deaths_mod <- lm(age ~ intent + police + sex + race + education,
  data = gun_deaths
)

# regression results table
tidy(gun_deaths_mod) %>%
  mutate(
    # remove prefixes from variable labels using socviz::prefix_strip
    term = prefix_strip(term, c("intent", "race", "sex", "education")),
    # fix variable labels
    term = recode(
      term,
      M = "Male"
    )
  ) %>%
  knitr::kable(
    col.names = c(
      "Term", "Coefficient", "Standard Error",
      "t-statistic", "p-value"
    ),
    digits = 2,
    format = "html"
  )
```

---

# Coefficient plots

```{r complex-model-plot, dependson = "complex-model", fig.height = 8}
gun_deaths_mod_vars <- c("intent", "race", "sex", "education")

tidy(gun_deaths_mod, conf.int = TRUE) %>%
  mutate(
    # remove prefixes from variable labels using socviz::prefix_strip
    term = prefix_replace(
      var_names = term,
      prefixes = gun_deaths_mod_vars,
      replacements = str_c(gun_deaths_mod_vars, "_")
    )
  ) %>%
  separate(col = term, into = c("variable", "value"), sep = "_") %>%
  mutate(
    variable = str_to_title(variable),
    value = recode(value, M = "Male")
  ) %>%
  filter(value != "Intercept") %>%
  mutate(value = reorder_within(value, estimate, variable)) %>%
  ggplot(mapping = aes(
    x = value,
    y = estimate,
    ymin = conf.low,
    ymax = conf.high
  )) +
  geom_pointrange() +
  scale_x_reordered() +
  facet_wrap(~variable, scales = "free_y", ncol = 1) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Value"
  )
```

---

# Marginal effects plots

* Expected change in $Y$ associated with a change in $X$
* Constant in OLS models
* Non-constant in logistic regression models

---

# Marginal effects plots

```{r titanic-pred, include = FALSE}
# estimate interactive model
titanic_glm <- glm(Survived ~ Age * Sex, data = titanic_train, family = "binomial")

# generate predicted values using margins::cplot()
age_range <- range(titanic_train$Age, na.rm = TRUE) %>%
  round()
age_range <- seq(from = age_range[[1]], to = age_range[[2]])

titanic_female <- cplot(titanic_glm,
  x = "Age",
  data = prediction::find_data(titanic_glm) %>%
    filter(Sex == "female"),
  xvals = age_range,
  what = "prediction",
  draw = FALSE
)
titanic_male <- cplot(titanic_glm,
  x = "Age",
  data = prediction::find_data(titanic_glm) %>%
    filter(Sex == "male"),
  xvals = age_range,
  what = "prediction",
  draw = FALSE
)

titanic_pred <- bind_rows(
  Female = titanic_female,
  Male = titanic_male,
  .id = "sex"
)
```

```{r titanic-marg-eff, dependson = "titanic-pred", fig.height = 8}
titanic_rolling <- map2_df(
  .x = seq(from = 0, to = 60, length.out = 50),
  .y = seq(from = 20, to = 80, length.out = 50),
  ~ filter(titanic_pred) %>%
    filter(xvals >= .x, xvals <= .y) %>%
    group_by(sex) %>%
    mutate(ymin = ifelse(sex == "Female", min(yvals), max(yvals))),
  .id = "start_val"
) %>%
  ungroup() %>%
  # add unique group for each start_val/sex combo
  # to avoid errors with transition_state
  mutate(
    group = group_indices(., start_val, sex),
    start_val = parse_number(start_val)
  )

ggplot(titanic_pred, aes(x = xvals, color = sex)) +
  geom_line(aes(y = yvals)) +
  geom_line(aes(y = upper), linetype = 2) +
  geom_line(aes(y = lower), linetype = 2) +
  geom_ribbon(
    data = titanic_rolling,
    aes(x = xvals, ymin = ymin, ymax = yvals, fill = sex, group = group),
    position = "identity"
  ) +
  scale_color_brewer(type = "qual", aesthetics = c("color", "fill")) +
  labs(
    title = "Probability of Surviving the Titanic",
    x = "Age",
    y = "Predicted Probability of Survival",
    color = "Sex",
    fill = "Sex"
  ) +
  theme(legend.position = "bottom") +
  transition_states(states = start_val, transition_length = .5, wrap = FALSE)
```

---

# Average marginal effects

```{r obama-mod}
gss_sm$polviews_m <- relevel(gss_sm$polviews, ref = "Moderate")

out_bo <- glm(obama ~ polviews_m + sex * race,
  family = "binomial", data = gss_sm
)
tidy(out_bo)
```

---

# Average marginal effects

```{r obama-margins, dependson = "obama-mod"}
bo_m <- margins(out_bo)

bo_gg <- as_tibble(summary(bo_m))
prefixes <- c("polviews_m", "sex")
bo_gg$factor <- prefix_strip(bo_gg$factor, prefixes)
bo_gg$factor <- prefix_replace(bo_gg$factor, "race", "Race: ")

p <- ggplot(data = bo_gg, aes(
  x = reorder(factor, AME),
  y = AME, ymin = lower, ymax = upper
))

p + geom_hline(yintercept = 0, color = "gray80") +
  geom_pointrange() + coord_flip() +
  labs(x = NULL, y = "Average Marginal Effect")
```

---

# Conditional effects

```{r obama-cond, dependson = "obama-mod", include = FALSE}
ce_sex <- cplot(out_bo, x = "sex", draw = FALSE) %>%
  mutate(variable = "Sex")
ce_race <- cplot(out_bo, x = "race", draw = FALSE) %>%
  mutate(variable = "Race")
ce_obama <- bind_rows(ce_sex, ce_race)
```

```{r obama-cond-plot, dependson = "obama-cond"}
ce_obama %>%
  mutate(xvals = reorder_within(xvals, yvals, variable)) %>%
  ggplot(mapping = aes(
    x = xvals,
    y = yvals, ymin = lower, ymax = upper
  )) +
  geom_hline(yintercept = 0, color = "gray80") +
  geom_pointrange() +
  scale_x_reordered() +
  coord_flip() +
  facet_wrap(~variable, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "Conditional Effect")
```

---

# Marginal effects plots

<div style="width:100%;height:0;padding-bottom:41%;position:relative;"><iframe src="https://giphy.com/embed/H7x1H0veAJlo4" width="100%" height="100%" style="position:absolute" frameBorder="0" class="giphy-embed" allowFullScreen></iframe></div>
