---
title: "Practice transforming and visualizing factors"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
library(knitr)
library(rcfss)
theme_set(theme_minimal())

# load the data
data("gun_deaths")
gun_deaths
```

# Convert `month` into a factor column

<details> 
  <summary>Click for the solution</summary>
  <p>
  
```{r month-factor}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)

# or use the built-in constant
month.abb

(gun_deaths <- gun_deaths %>%
  mutate(month = factor(month, levels = seq(from = 1, to = 12), labels = month_levels)))
```

  </p>
</details>

# Visualize the total gun deaths per month, in chronological order

<details> 
  <summary>Click for the solution</summary>
  <p>
  
```{r month-deaths}
ggplot(gun_deaths, aes(month)) +
  geom_bar() +
  labs(title = "Gun Deaths in the United States (2012-2014)",
       x = "Month",
       y = "Number of gun deaths")
```

  </p>
</details>

# Visualize the total gun deaths per month, sorted from lowest to highest

<details> 
  <summary>Click for the solution</summary>
  <p>
  
```{r month-deaths-sort}
# with geom_col() and fct_reorder()
gun_deaths %>%
  count(month) %>%
  ggplot(aes(fct_reorder(month, n), n)) +
  geom_col() +
  labs(title = "Gun Deaths in the United States (2012-2014)",
       x = "Month",
       y = "Number of gun deaths")

# with geom_bar() and fct_infreq()
ggplot(gun_deaths, aes(month %>%
                         fct_infreq() %>%
                         fct_rev())) +
  geom_bar() +
  labs(title = "Gun Deaths in the United States (2012-2014)",
       x = "Month",
       y = "Number of gun deaths")
```

  </p>
</details>

# Visualize the frequency of intent of gun deaths using a bar chart, sorted from most to least frequent

<details> 
  <summary>Click for the solution</summary>
  <p>
  
```{r intent}
intent_levels <- c("Accidental", "Homicide", "Suicide", "Undetermined")

gun_deaths %>%
  filter(!is.na(intent)) %>%
  mutate(intent = parse_factor(intent, levels = intent_levels)) %>%
  ggplot(aes(intent %>%
               fct_infreq() %>%
               fct_rev())) +
  geom_bar() +
  labs(title = "Gun Deaths in the United States (2012-2014)",
       x = "Intent of death",
       y = "Number of gun deaths") +
  coord_flip()
```

  </p>
</details>

# Visualize total gun deaths by season of the year using a bar chart.

Hint: do not use `cut()` to create the `season` column.

<details> 
  <summary>Click for the solution</summary>
  <p>
  
```{r season}
gun_deaths %>%
  mutate(season = fct_collapse(month,
                               "Winter" = c("Jan", "Feb", "Mar"),
                               "Spring" = c("Apr", "May", "Jun"),
                               "Summer" = c("Jul", "Aug", "Sep"),
                               "Fall" = c("Oct", "Nov", "Dec"))) %>%
  ggplot(aes(season)) +
  geom_bar() +
  labs(title = "Gun Deaths in the United States (2012-2014)",
       x = "Season",
       y = "Number of gun deaths")
```

  </p>
</details>

# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```




