---
title: "Presenting academic research"
author: "MACS 30200 - Perspectives on Computational Research"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Objectives

* Identify major components of a research poster
* Discuss differences between a research paper and a research poster
* Present examples of research posters
* Identify templates for creating research posters in Powerpoint, $\LaTeX$, and R Markdown
* Identify locations to print research posters at UChicago

```{r packages, cache = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal())
```

# Designing research posters

## Potential modes of presenting results at conferences

Conferences are a common location to present research and share results with the larger academic community. Depending on the conference, you may have opportunities to present work in the following ways:

### Oral presentation

Typically you are a part of a panel of related papers (between 4-6 papers per panel). You submit a proposal months in advance, typically just based on the abstract.^[Some highly selective conferences require you to submit the entire paper when you apply for the conference.] You write the paper and send it to the other members of the panel. The panel consists of the other presenters, as well as a **chair** and/or **discussants**. Discussants are assigned to read the papers in advance of the conference and prepare a critique to share during the panel session. The chair may also serve as a discussant, but more generally is responsible for cajoling the panel members into submitting their papers in a timely fashion, ensuring the discussants read the papers and show up for the panel, and keep track of time during the panel to make sure panelists stick to the time requirements.

During the panel session, each panelist is allotted between 8-15 minutes to present their paper. Panelists typically prepare a slideshow to support their oral presentation. There may be an additional 5-10 minutes of critique from the discussants, and then Q&A with the audience.

#### Lightning talk

**Lightning talks** are a specialized format that condenses oral presentations into rapid fire forms. Individuals typically get five minutes for their oral presentation, with talks arranged back-to-back. Many lightning talks also standardize the slides format, so that the speaker can use precisely 20 slides that auto-advance every 15 seconds. This format forces speakers to condense their points and be very clear and precise to convey only the critical information.

### Poster presentation

While graduate students can be accepted at conferences as part of panels, many graduate students participate in **poster sessions**. Again I am generalizing, but typically conferences group research projects by common themes or subdisciplines with all speakers scheduled for a 2 hour session. During this time the posters are on display and the authors stand with their posters as interested audience members walk around and view the posters. Unlike with oral presentations, all interaction is one-on-one with the audience. The poster serves as the visual medium for drawing in the audience and communicating the key findings of the project, and the audience can then ask follow-up questions of the author.

## Goals of a research poster

A research poster can be thought of as an extended abstract for your research paper or project. It is an opportunity to present research results without requiring the audience to read a full paper or even attend a guest lecture or oral presentation. It is a **snapshot** of your research - more in-depth than an abstract, but less in-depth than the full paper. Designing a good-quality research poster is tricky, as you want to provide enough information so that readers can engage in a detailed conversation about the work, but remain short enough so that a reader can consume it in just a few minutes and walk away understanding the importance of the research.

## Major components

Many of the components of a research paper overlap with the research paper on which it is based.

## Must haves

* Title - convey the issue or research question
* Introduction - quickly introduce the reader to the issue or question
    * Avoid extraneous background information and jargon
    * Quickly situate the issue in the literature and pitch an interesting, novel hypothesis
* Data and methods - identify the research design and procedure
    * Does not have to fully flesh out all aspects, but establish the major points
* Results - the outcome of the research
* Conclusions
    * Reminder of the major result and whether the hypothesis/hypotheses were supported by the data
    * Convince the reader why your project is interesting
* Contact information
    * How will the audience find you later?
    * Include institutional affiliation, email address, web site, etc
    * If you want to be really fancy, include a QR code with a link to the full page/GitHub repo online

## Optional

* Literature cited
    * If you include references, you can include the full citations on the poster
    * But in my opinion, this is a waste of space. I'd recommend simply referring readers to your full paper for the works cited page.^[You do have a copy available to hand out or show online, right?]
* Acknowledgements
    * Typically found in the research paper
    * Doesn't need to go into the poster
    * **Exception** - if you receive internal or external funding for the project, it is good practice to include an acknowledgment on the poster

### What not to include

* Abstract - the poster is an extended abstract, you don't need to duplicate this!
* Verbatim text - don't just copy and paste chunks of text from your paper into the poster! Summarize, reshape, condense, etc.

## Tips for designing poster content and layouts

* Don't reproduce the paper
* Keep the words to a minimum
* Choose your words carefully (informative phrases)
* Have a logical flow
* Make the poster legible
* Visualize your results
    * Functional
    * Pretty

## Examples of scientific posters

* [A bad poster](https://s-media-cache-ak0.pinimg.com/originals/18/1a/6c/181a6cd948f5f0ac02bdd0509db90779.jpg)
* [Do decisions follow dollars? Campaign contributions and state supreme court decisions](https://github.com/UC-MACSS/persp-research/raw/master/demos/poster-example-1.pdf)
* [A matched case analysis of Supreme Court behavior under mandatory and discretionary jurisdiction](https://github.com/UC-MACSS/persp-research/raw/master/demos/poster-example-2.pdf)

## Presenting the poster

Poster presentations are not just "setup the poster and leave". It is an interactive presentation.

* Prepare an oral pitch for your project. You should create a memorable and positive impression and open the door to further conversation.
    * What is the topic of your research?
    * What is the question you want to answer?
    * Why is this question interesting or important?
    * How does this work connect to broader topics or problems in your discipline?
    * This is not a duplication of your poster
* Don't just read from your poster - the poster is a background the audience can refer to while you talk
* Behave professionally
    * Dress professionally. For our session, I expect you to wear at least [business casual](https://www.thebalance.com/what-is-business-casual-attire-2061168)
    * Don't be on your phone the entire time. If you need to check it, be discreet then put it away
    * Talk to your audience, not your poster
    * Treat the poster session like a job interview. Your audience could include a future employer or PhD advisor. Make a good impression
* Have copies of your paper as handouts

# Templates for creating research posters

## Powerpoint

* [UChicago template](https://github.com/UC-MACSS/persp-research/blob/master/demos/E-Poster%20Template%20Academic%20conference.ppt)
* [Poster Presentations templates](https://www.posterpresentations.com/html/free_poster_templates.html) - [PosterPresentations.com](http://www.posterpresentations.com) provides poster printing services (shipped to you, so turnaround time is longer), but also offers free Powerpoint templates

## $\LaTeX$

* [$\LaTeX$ poster template](http://www.brian-amberg.de/uni/poster/) - older and clunkier, but what I used for the first example
* [`beamerposter`](https://github.com/deselaers/latex-beamerposter) - used for the second example poster

## R Markdown

* [`flexdashboard` poster](https://github.com/odeleongt/flexdashboard-poster) - uses `flexdashboard` to design the poster. Rendering as a PDF seems to be broken, but rendering as a `.png` file works fine and quality looks good. Need to adjust the parameters in the rendering script to get an appropriate size and aspect ratio
* [`drposter`](https://github.com/bbucior/drposter) - package for generating posters in HTML/CSS using `reveal.js` to generate the pdf document. Still in alpha testing, not sure how effective this will be

# Printing posters

* It's best to render the poster as a PDF document. This leads to the best print quality
* **ALWAYS ALWAYS ALWAYS PROOF-READ YOUR POSTER**. Don't waste 50+ dollars printing your poster only to find out you have a glaring typo in your title. Make sure everything is spelled correctly, all your figures and tables are based on the correct data or model specification, etc.
* Lamination - unless you plan to reuse a poster, don't bother spending the extra money to laminate the poster
* Foam core backing - you don't need it for this poster session. We will provide display boards and clips to secure your poster to the board

## Poster size

* 36 x 48 inches
* 48 x 60 inches
* The larger the poster, the more expensive the print job
* I recommend using a landscape orientation (longest dimension is width), rather than portrait (longest dimension is height). It's easier to view all the content at eye level

## Where to print the poster

* [Booth School print shop](https://facilities.chicagobooth.edu/old/poster&banner.htm)
    * \$28 for a 36 x 48 laminated poster
    * Generally requires 3 business days to print, so if you do this you probably need to place the order by the end of Friday this week (remember that Monday is a federal holiday)
* [John Crerar Library - PSD Graphic Arts](http://graphicarts.uchicago.edu/)
    * Price list online is not correct - I confirmed this via email
    * ??? for a 24 x 36 inch non-laminated poster
    * \$85 for a 36 x 48 inch non-laminated poster
    * Benefit - turnaround time is 24 hours. From their site:
    
        > We can generally have a poster ready within 24 hours. Files received before NOON (12:00pm) will be ready for pick up during the afternoon of the same day before the office closes (4:30pm). Files received after 12:00pm will be ready for pick up when the office opens (8:30am) on the following morning. Turnaround for poster printing is based on an order of one poster from a single individual requested during posted office hours (8:30am – 4:30pm / Monday – Friday).
        
        So theoretically you could wait until Tuesday evening the night before the poster session to submit your order and still pick it up before 11:30am. I don't think I'd wait that long.
    * They don't take credit cards. Bring cash or check
* [Fedex Office](http://www.fedex.com/us/office/) at 1315 E 57th St
    * Starting at \$43.50 for 24 x 36 inch non-laminated poster
    * Starting at \$87 for 36 x 48 inch non-laminated poster
    * Requires at least 24 hour to print and be ready for pickup

# Grading rubric for poster presentation

1. Overall design/layout (10 points)
    * Clarity
    * Organization
    * Structure/flow
    * Use of color
    * Parsimony and efficiency of language
1. Visualization (10 points)
    * Use of plots and tables to visualize and communicate results
1. Content (10 points)
    * What is the question?
    * What is the answer?
    * How did you answer the question?
    * How strong is the evidence?

# Acknowledgments {.toc-ignore}

* [Designing conference posters](http://colinpurrington.com/tips/poster-design)

# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```




