---
title: "Syllabus"
output:
  html_document:
    code_download: false
---

```{r include = FALSE}
library(readxl)
library(dplyr)
library(lubridate)
library(stringr)
library(here)
```

```{r include = FALSE}
# import excel table with topics
rawdat <- here("course_admin", "schedule.xlsx") %>%
  read_excel()

# clean up table and extract date info
cdat <- rawdat %>%
  mutate(wday = wday(date, label = TRUE),
         month = month(date, label = TRUE),
         day = day(date),
         date = paste(wday, month, day),
         date = ifelse(is.na(wday), NA, date),
         link_it = as.logical(link_it),
         notes = sprintf("%s: %s", cm, topic))

# identify links to course modules
cmdat <- data_frame(html = list.files(pattern = '^(cm|setup|git).*html$'))
cmdat <- cmdat %>% mutate(cm = str_extract(html, "(cm|setup|git)[0-9]+"))
cdat <- cdat %>% left_join(cmdat)

# add links to notes column
cdat$notes[cdat$link_it] <-
  str_replace(cdat$notes[cdat$link_it],
              "((cm[0-9]{3})|((setup|git)[0-9]{2}))", '<a href="LINK">\\1</a>')
cdat$notes[cdat$link_it] <-
  str_replace(cdat$notes[cdat$link_it], "LINK", cdat$html[cdat$link_it])
```

```{r results = 'asis', echo = FALSE}
cdat %>%
  select(date, notes) %>%
  knitr::kable(col.names = c("Date", "Topic"))
```

## Homework

* [Homework 01: Edit README.md](hw01-edit-README.html)
* [Homework 02: Exploring and visualizing data](hw02-explore-data.html)
* [Homework 03: Wrangling and visualizing data](hw03-wrangle-data.html)
* [Homework 04: Programming in R](hw04-programming.html)
* [Homework 05: Generating reproducible research](hw05-reproducible-research.html)
* [Homework 06: Statistical learning](hw06-stat-learn.html)
* [Homework 07: Geospatial visualization](hw07-geospatial.html)
* [Homework 08: Collecting and analyzing data from the web](hw08-webdata.html)
* [Homework 09: Analyzing text data](hw09-text_analysis.html)
* [Homework 10: Build a Shiny application](hw10-shiny.html)
