---
title: "Practicing sentiment analysis with Harry Potter"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r packages, cache = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(tidytext)
library(wordcloud)
library(harrypotter)

set.seed(1234)
theme_set(theme_minimal())
```

# Load Harry Potter text

```r
devtools::install_github("bradleyboehmke/harrypotter")
```

```{r hp}
library(harrypotter)

# names of each book
hp_books <- c("philosophers_stone", "chamber_of_secrets",
              "prisoner_of_azkaban", "goblet_of_fire",
              "order_of_the_phoenix", "half_blood_prince",
              "deathly_hallows")

# combine books into a list
hp_words <- list(
  philosophers_stone,
  chamber_of_secrets,
  prisoner_of_azkaban,
  goblet_of_fire,
  order_of_the_phoenix,
  half_blood_prince,
  deathly_hallows
) %>%
  # name each list element
  set_names(hp_books) %>%
  # convert each book to a data frame and merge into a single data frame
  map_df(as_tibble, .id = "book") %>%
  # convert book to a factor
  mutate(book = factor(book, levels = hp_books)) %>%
  # remove empty chapters
  filter(!is.na(value)) %>%
  # create a chapter id column
  group_by(book) %>%
  mutate(chapter = row_number(book)) %>%
  # tokenize the data frame
  unnest_tokens(word, value)

hp_words
```

# Most frequent words, by book

Remove stop words.

```{r word-freq, fig.asp = 2}
hp_words %>%
  # delete stopwords
  anti_join(stop_words) %>%
  # summarize count per word per book
  count(book, word) %>%
  # highest freq on top
  arrange(desc(n)) %>% 
  # identify rank within group
  group_by(book) %>% # 
  mutate(top = seq_along(word)) %>%
  # retain top 15 frequent words
  filter(top <= 15) %>%
  # create barplot
  ggplot(aes(x = -top, y = n, fill = book)) + 
  geom_col(color = "black") +
  # print words in plot instead of as axis labels
  geom_text(aes(label = word), hjust = "left", nudge_y = 100) +
  labs(title = "Most frequent words in Harry Potter",
       x = NULL,
       y = "Word count") +
  facet_wrap( ~ book) +
  coord_flip() +
  theme(legend.position = "none",
        # rotate x text
        axis.text.x = element_text(angle = 45, hjust = 1),
        # remove tick marks and text on y-axis
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank())
```

# Estimate sentiment

## Generate data frame with sentiment derived from the NRC

<details> 
  <summary>Click for the solution</summary>
  <p>
  
```{r sentiment-nrc}
(hp_nrc <- hp_words %>% 
  inner_join(get_sentiments("nrc")) %>%
  group_by(book, chapter, sentiment))
```

  </p>
</details>

## Visualize which words in the NRC sentiment dictionary appear most frequently

Sometimes words which are defined in a general sentiment dictionary can be outliers in specific contexts. That is, an author may use a word without intending to convey a specific sentiment but the dictionary defines it in a certain way.

We can use a [**wordcloud**](http://tidytextmining.com/sentiment.html#wordclouds) as a quick check to see if there are any outliers in the context of *Harry Potter*.

```{r sentiment-outliers}
library(wordcloud)

hp_nrc %>%
  group_by(word) %>%
  count() %>%
  with(wordcloud(word, n, max.words = 100))
```

As we can see, "harry" appears quite frequently in the books. In the vast majority of appearances, "harry" is used to refer to the main character and is not meant to convey a specific sentiment.

```{r sentiment-harry}
get_sentiments("nrc") %>%
  filter(word == "harry")
```

It would be best to remove this word from further sentiment analysis, treating it as if it were another stop word.

```{r sentiment-outlier-remove}
hp_nrc <- hp_nrc %>%
  filter(word != "harry")
```

## Visualize which words appear most frequently for each sentiment type

<details> 
  <summary>Click for a solution</summary>
  <p>
  
```{r nrc-freq, fig.asp = 1}
hp_nrc %>%
  # summarize count per word
  ungroup %>%
  count(word, sentiment) %>%
  # highest freq on top
  arrange(desc(n)) %>% 
  # identify rank within group
  group_by(sentiment) %>% # 
  mutate(top = seq_along(word)) %>%
  # retain top 15 frequent words
  filter(top <= 15) %>%
  # create barplot
  ggplot(aes(x = -top, y = n, fill = sentiment)) + 
  geom_col(color = "black") +
  # print words in plot instead of as axis labels
  geom_text(aes(label = word), hjust = "left", nudge_y = 100) +
  labs(title = "Most frequent words in Harry Potter",
       x = NULL,
       y = "Word count") +
  facet_wrap( ~ sentiment, ncol = 5) +
  coord_flip() +
  theme(legend.position = "none",
        # rotate x text
        axis.text.x = element_text(angle = 45, hjust = 1),
        # remove tick marks and text on y-axis
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank())
```

  </p>
</details>

## Visualize the positive/negative sentiment for each book over time using the AFINN dictionary

<details> 
  <summary>Click for a solution</summary>
  <p>
  
```{r affin-over-time}
hp_words %>% 
  inner_join(get_sentiments("afinn")) %>%
  group_by(book, chapter) %>%
  summarize(score = sum(score)) %>%
  ggplot(aes(chapter, score, fill = book)) +
  geom_col() +
  facet_wrap(~ book, scales = "free_x") +
  labs(title = "Emotional arc of Harry Potter books",
       subtitle = "AFINN sentiment dictionary",
       x = "Chapter",
       y = "Emotional score") +
  theme(legend.position = "none")

# cumulative score
hp_words %>% 
  inner_join(get_sentiments("afinn")) %>%
  group_by(book) %>%
  mutate(cumscore = cumsum(score)) %>%
  ggplot(aes(chapter, cumscore, fill = book)) +
  geom_step() +
  facet_wrap(~ book, scales = "free_x") +
  labs(title = "Emotional arc of Harry Potter books",
       subtitle = "AFINN sentiment dictionary",
       x = "Chapter",
       y = "Cumulative emotional score")
```

  </p>
</details>

## Visualize the sentimental content of each chapter in each book using the NRC dictionary

<details> 
  <summary>Click for a solution</summary>
  <p>
  
```{r sentiment-over-time, fig.asp = 1}
hp_nrc %>%
  count(sentiment, book, chapter) %>%
  filter(!(sentiment %in% c("positive", "negative"))) %>%
  # create area plot
  ggplot(aes(x = chapter, y = n)) +
  geom_col(aes(fill = sentiment)) + 
  # add black smoothing line without standard error
  geom_smooth(aes(fill = sentiment), method = "loess", se = F, col = 'black') + 
  theme(legend.position = 'none') +
  labs(x = "Chapter", y = "Emotion score", # add labels
       title = "Harry Plotter: Emotions during the saga",
       subtitle = "Using tidytext and the nrc sentiment dictionary") +
  # seperate plots per sentiment and book and free up x-axes
  facet_grid(sentiment ~ book, scale = "free")
```

  </p>
</details>

# Acknowledgments {.toc-ignore}

* This page is derived in part from [Harry Plotter: Celebrating the 20 year anniversary with `tidytext` and the `tidyverse` in R](https://paulvanderlaken.com/2017/08/03/harry-plotter-celebrating-the-20-year-anniversary-with-tidytext-the-tidyverse-and-r/) and licensed under a [Creative Commons Attribution-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-sa/4.0/).

# Session Info {.toc-ignore}

```{r child='_sessioninfo.Rmd'}
```
