---
title: "Interaction terms in regression models"
author: |
  | MACS 30200
  | University of Chicago
output: rcfss::cfss_slides
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      echo = FALSE)

library(tidyverse)
library(forcats)
library(broom)
library(modelr)
library(stringr)
library(titanic)
library(rcfss)
library(car)
library(plotly)
library(haven)
library(coefplot)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal(base_size = 22))
```

## Additive model

$$Y = \beta_0 + \beta_1 X + \beta_2 Z + e_i$$

## Additive model

```{r sim-linear}
n_sim <- 100

additive <- data_frame(x = runif(n_sim),
                       z = runif(n_sim)) %>%
  mutate(z = ifelse(z > .5, 1, 0),
         y = x + z + rnorm(n_sim))

additive %>%
  add_predictions(lm(y ~ x + z, data = .)) %>%
  ggplot(aes(x, y, color = factor(z))) +
  geom_point() +
  geom_line(aes(y = pred)) +
  theme(legend.position = "none")
```

## Additive model

$$E(Y) = \beta_0 + \beta_1 X + \beta_2 Z$$

$$\frac{\delta E(Y)}{\delta X} = \beta_1$$

$$\frac{\delta E(Y)}{\delta Z} = \beta_2$$

## Multiplicative interaction model

$$Y = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 XZ + e_i$$

* Direct effects
* Constitutive terms
* Interaction term

## Multiplicative interaction model

$$
\begin{split}
E(Y) & = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 XZ \\
     & = \beta_0 + \beta_2 Z + (\beta_1 + \beta_3 Z) X
\end{split}
$$

$$\frac{\delta E(Y)}{\delta X} = \beta_1 + \beta_3 Z$$

$$E(Y) = \beta_0 + \beta_2 Z + \psi_1 X$$

$$
\begin{split}
E(Y) & = \beta_0 + \beta_1 X + (\beta_2 + \beta_3 X) Z \\
     & = \beta_0 + \beta_1 X + \psi_2 Z
\end{split}
$$

## Multiplicative interaction model

* Conditional impact
* If $Z = 0$, then:
    
    $$
\begin{split}
E(Y) & = \beta_0 + \beta_1 X + \beta_2 (0) + \beta_3 X (0) \\
     & = \beta_0 + \beta_1 X
\end{split}
    $$
        
* If $X = 0$, then:
    
    $$
\begin{split}
E(Y) & = \beta_0 + \beta_1 (0) + \beta_2 Z + \beta_3 (0) Z \\
     & = \beta_0 + \beta_2 Z
\end{split}
    $$
* $\psi_1 = \beta_1$ and $\psi_2 = \beta_2$
* $+\beta_3$ and $-\beta_3$
* $\psi_1$ and $\psi_2$

## Conducting inference

* Obtaining estimates of parameters

    $$\hat{\psi}_1 = \hat{\beta}_1 + \hat{\beta}_3 Z$$
    $$\hat{\psi}_2 = \hat{\beta}_2 + \hat{\beta}_3 X$$
* Obtaining estimates of standard errors

## Conducting inference

1. $\text{Var}(aX) = a^2 \text{Var}(X)$
1. $\text{Var}(X+Y) = \text{Var}(X) + \text{Var}(Y) + 2 \text{Cov}(X,Y)$
1. $\text{Cov}(X, aY) = a \text{Cov}(X,Y)$

## Conducting inference

$$\widehat{\text{Var}(\hat{\psi}_1}) = \widehat{\text{Var} (\hat{\beta}_1)} +Z^2  \widehat{\text{Var} (\hat{\beta}_3)} + 2 Z \widehat{\text{Cov} (\hat{\beta}_1, \hat{\beta}_3)}$$

$$\widehat{\text{Var}(\hat{\psi}_2}) = \widehat{\text{Var} (\hat{\beta}_2)} + X^2  \widehat{\text{Var} (\hat{\beta}_3)} + 2 X \widehat{\text{Cov} (\hat{\beta}_2, \hat{\beta}_3)}$$

* Depend on $\beta_1$, $\beta_2$, and/or $\beta_3$
* Both also depend on the level/value of the interacted variable

## Two dichtomous covariates

$$Y = \beta_0 + \beta_1 D_1 + \beta_2 D_2 + \beta_3 D_1 D_2 + e_i$$

$$
\begin{split}
E(Y | D_1 = 0, D_2 = 0) & = \beta_0 \\
E(Y | D_1 = 1, D_2 = 0) & = \beta_0 + \beta_1 \\
E(Y | D_1 = 0, D_2 = 1) & = \beta_0 + \beta_2 \\
E(Y | D_1 = 1, D_2 = 1) & = \beta_0 + \beta_1 + \beta_2 + \beta_3 \\
\end{split}
$$

## Two dichtomous covariates

```{r sim-two-dich}
two_dich <- data_frame(x = runif(n_sim),
                       z = runif(n_sim)) %>%
  mutate_at(vars(x, z), funs(ifelse(. > .5, 1, 0))) %>%
  mutate(y = 10 + 20 * x - 20 * z + 40 * (x * z) + rnorm(n_sim, 0, 5))
```

```{r sim-two-dich-hist}
ggplot(two_dich, aes(y, color = interaction(x, z))) +
  geom_density() +
  scale_color_discrete(labels = c("D1 = D2 = 0",
                                  "D1 = 1, D2 = 0",
                                  "D1 = 0, D2 = 1",
                                  "D1 = D2 = 1"),
                       guide = guide_legend(nrow = 2)) +
  labs(title = expression(paste("Values of ", Y, " for various combinations of values of ", D[1], ",", D[2])),
       x = "Value of Y",
       color = NULL) +
  theme(legend.position = "bottom")
```

## Two dichtomous covariates

```{r sim-two-dich-box}
ggplot(two_dich, aes(interaction(z, x), y)) +
  geom_boxplot() +
  scale_x_discrete(labels = c(expression("D1 = D2 = 0"),
                                  expression("D1 = 0, D2 = 1"),
                                  expression("D1 = 1, D2 = 0"),
                                  expression("D1 = D2 = 1"))) +
  labs(title = expression(paste("Values of ", Y, " for various combinations of values of ", D[1], ",", D[2])),
       x = NULL,
       y = "Values of Y")
```

## One dichotomous and one continuous covariate

$$Y = \beta_0 + \beta_1 X + \beta_2 D + \beta_3 XD + e_i$$

$$
\begin{split}
E(Y | X, D = 0) & = \beta_0 + \beta_1 X \\
E(Y | X, D = 1) & = (\beta_0 + \beta_2) + (\beta_1 + \beta_3) X
\end{split}
$$

----

```{r sim-cont-dich-00}
    data_frame(x = runif(n_sim, -2, 2),
               z = runif(n_sim, -2, 2)) %>%
      mutate_at(vars(z), funs(ifelse(. > .5, 1, 0))) %>%
      mutate(y = 0 + 20 * x + 0 * z + 0 * (x * z) + rnorm(n_sim, 0, 5)) %>%
      ggplot(aes(x, y, color = factor(z), shape = factor(z))) +
      geom_vline(xintercept = 0, alpha = .3) +
      geom_hline(yintercept = 0, alpha = .3) +
      geom_point() +
      geom_abline(intercept = 0, slope = 20) +
      labs(title = expression({beta[2] == beta[3]} == 0),
           color = "Z",
           shape = "Z") +
      theme(legend.position = "bottom")
```
    
----

```{r sim-cont-dich-01}
    data_frame(x = runif(n_sim, -2, 2),
               z = runif(n_sim, -2, 2)) %>%
      mutate_at(vars(z), funs(ifelse(. > .5, 1, 0))) %>%
      mutate(y = 0 + 20 * x + 10 * z + 0 * (x * z) + rnorm(n_sim, 0, 5)) %>%
      ggplot(aes(x, y, color = factor(z), shape = factor(z))) +
      geom_vline(xintercept = 0, alpha = .3) +
      geom_hline(yintercept = 0, alpha = .3) +
      geom_point() +
      geom_abline(intercept = 0, slope = 20) +
      geom_abline(intercept = 10, slope = 20, linetype = 2) +
      labs(title = expression(paste({beta[2] != 0}, ",", beta[3] == 0)),
           color = "Z",
           shape = "Z") +
      theme(legend.position = "bottom")
```
    
----

```{r sim-cont-dich-10}
    data_frame(x = runif(n_sim, -2, 2),
               z = runif(n_sim, -2, 2)) %>%
      mutate_at(vars(z), funs(ifelse(. > .5, 1, 0))) %>%
      mutate(y = 0 + 20 * x + 0 * z + -40 * (x * z) + rnorm(n_sim, 0, 5)) %>%
      ggplot(aes(x, y, color = factor(z), shape = factor(z))) +
      geom_vline(xintercept = 0, alpha = .3) +
      geom_hline(yintercept = 0, alpha = .3) +
      geom_point() +
      geom_abline(intercept = 0, slope = 20) +
      geom_abline(intercept = 0, slope = -20, linetype = 2) +
      labs(title = expression(paste({beta[2] == 0}, ",", beta[3] != 0)),
           color = "Z",
           shape = "Z") +
      theme(legend.position = "bottom")
```
    
----

```{r sim-cont-dich-11}
    data_frame(x = runif(n_sim, -2, 2),
               z = runif(n_sim, -2, 2)) %>%
      mutate_at(vars(z), funs(ifelse(. > .5, 1, 0))) %>%
      mutate(y = 0 + 20 * x + 20 * z + -40 * (x * z) + rnorm(n_sim, 0, 5)) %>%
      ggplot(aes(x, y, color = factor(z), shape = factor(z))) +
      geom_vline(xintercept = 0, alpha = .3) +
      geom_hline(yintercept = 0, alpha = .3) +
      geom_point() +
      geom_abline(intercept = 0, slope = 20) +
      geom_abline(intercept = 20, slope = -20, linetype = 2) +
      labs(title = expression(paste({beta[2] == 0}, ",", beta[3] != 0)),
           color = "Z",
           shape = "Z") +
      theme(legend.position = "bottom")
```

## Two continuous covariates

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2 + e_i$$

```{r two-cont}
# no interactive effects
data_frame(x = runif(n_sim, 0, 10),
           z = runif(n_sim, 0, 10),
           y = 10 + 10 * x - 10 * z + 0 * x * z) %>%
  plot_ly(x = ~x, y = ~ z, z = ~ y, type = "mesh3d") %>%
  layout(title = "No interactive effects")
```

## Two continuous covariates

```{r two-cont-x}
# inteactive effects
data_frame(x = runif(n_sim, 0, 10),
           z = runif(n_sim, 0, 10),
           y = 10 + 10 * x - 10 * z + 10 * x * z) %>%
  plot_ly(x = ~x, y = ~ z, z = ~ y, type = "mesh3d") %>%
  layout(title = "Interactive effects")
```

## Quadratic, cubic, and other polynomial effects

$$Y = \beta_0 + \beta_1 X + \beta_2 X^2 + e$$

$$\frac{\delta E(Y)}{\delta X} = \beta_1 + 2 \beta_2 X$$

## Quadratic, cubic, and other polynomial effects {.scrollable}

```{r poly-interact}
data_frame(x = runif(n_sim, -2, 2)) %>%
  mutate(y = 10 + 10 * x - (50 * x^2) + rnorm(n_sim, 0, 5)) %>%
  add_predictions(lm(y ~ poly(x, 2, raw = TRUE), data = .)) %>%
  ggplot(aes(x, y)) +
  geom_point() +
  geom_line(aes(y = pred)) +
  labs(title = "Example of a quadratic relationship")

data_frame(x = runif(n_sim, -2, 2)) %>%
  mutate(y = -500 - 20 * x + 300 * x^2 + rnorm(n_sim, 0, 50)) %>%
  add_predictions(lm(y ~ poly(x, 2, raw = TRUE), data = .)) %>%
  ggplot(aes(x, y)) +
  geom_point() +
  geom_line(aes(y = pred)) +
  labs(title = "Example of a quadratic relationship")

data_frame(x = runif(n_sim, -2, 2)) %>%
  mutate(y = 10 + 10 * x - 50 * x^2 + 300 * x^3 + rnorm(n_sim, 0, 300)) %>%
  add_predictions(lm(y ~ poly(x, 3, raw = TRUE), data = .)) %>%
  ggplot(aes(x, y)) +
  geom_point() +
  geom_line(aes(y = pred)) +
  labs(title = "Example of a cubic relationship")
```

## Higher-order interaction terms

$$
\begin{align}
Y = \beta_0 &+ \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + \beta_4 X_1 X_2 \\
& + \beta_5 X_1 X_3 + \beta_6 X_2 X_3 + \beta_7 X_1 X_2 X_3 + e
\end{align}
$$

## Higher-order interaction terms

$$
\begin{align}
Y = \beta_0 &+ \beta_1 X + \beta_2 D_1 + \beta_3 D_2 + \beta_4 X D_1 \\
& + \beta_5 X D_2 + \beta_6 D_1 D_2 + \beta_7 X D_1 D_2 + e
\end{align}
$$

## Higher-order interaction terms

```{r three-way}
data_frame(x = runif(n_sim, -2, 2),
           d1 = runif(n_sim),
           d2 = runif(n_sim)) %>%
  mutate_at(vars(d1, d2), funs(ifelse(. > .5, 1, 0))) %>%
  mutate(y = 10 + 10 * x + 10 * d1 - 20 * d2 - 20 * x * d1 + 20 * x *d2 - 50 * x * d1 * d2 + rnorm(n_sim, 0, 10)) %>%
  add_predictions(lm(y ~ x * d1 * d2, data = .)) %>%
  ggplot(aes(x, y)) +
  facet_grid(d1 ~ d2, labeller = label_both) +
  geom_point() +
  geom_line(aes(y = pred))
```

## Key rules

* Don't omit the "direct effects"
* Zero should be meaningful
* Rescaling the variables doesn't guarantee statistical significance
* Flexible alternatives
* Interpreting three(+)-way interactions

## Estimating models with multiplicative interactions

````{r nes-data}
# get nes data
nes <- read_dta("../data/nes2008.dta") %>%
  select(obama_therm_post, partyid3, libcon7, libcon7_obama) %>%
  mutate_each(funs(ifelse(is.nan(.), NA, .))) %>%
  rename(ObamaTherm = obama_therm_post,
         RConserv = libcon7,
         ObamaConserv = libcon7_obama) %>%
  mutate(GOP = ifelse(partyid3 == 3, 1, 0)) %>%
  select(-partyid3) %>%
  na.omit
```

* Obama feeling thermometer (`ObamaTherm`)
* `RConserv`
* `ObamaConserv`
* `GOP`

## Obama data

```{r nes-sum}
summary(nes)
```

## Basic linear model

```{r obama-linear}
obama_base <- lm(ObamaTherm ~ RConserv + GOP, data = nes)
tidy(obama_base)
glance(obama_base)
```

## Dichotomous interaction

$$
\begin{align}
\text{Obama thermometer} = \beta_0 &+ \beta_1 (\text{Respondent conservatism}) \\
& + \beta_2 (\text{GOP respondent})\\
& + \beta_3 (\text{Respondent conservatism}) (\text{GOP respondent}) \\
& + e
\end{align}
$$

## Dichotomous interaction

```{r obama-ideo-x-gop}
obama_ideo_gop <- lm(ObamaTherm ~ RConserv * GOP, data = nes)
tidy(obama_ideo_gop)
glance(obama_ideo_gop)
```

## Dichotomous interaction

* `GOP = 0`

    $$
\begin{align}
E(\text{Obama thermometer}) = `r coef(obama_ideo_gop)[["(Intercept)"]]` & `r coef(obama_ideo_gop)[["RConserv"]]` (\text{Respondent conservatism})  `r coef(obama_ideo_gop)[["GOP"]]` (0)\\
&  `r coef(obama_ideo_gop)[["RConserv:GOP"]]` (\text{Respondent conservatism} \times 0) \\
= `r coef(obama_ideo_gop)[["(Intercept)"]]` & `r coef(obama_ideo_gop)[["RConserv"]]` (\text{Respondent conservatism})
\end{align}
    $$

* `GOP = 1`

    $$
\begin{align}
E(\text{Obama thermometer}) & = (`r coef(obama_ideo_gop)[["(Intercept)"]]` `r coef(obama_ideo_gop)[["GOP"]]` (1)) + (`r coef(obama_ideo_gop)[["RConserv"]]` `r coef(obama_ideo_gop)[["RConserv:GOP"]]` (\text{Respondent conservatism} \times 1)) \\
& = `r coef(obama_ideo_gop)[["(Intercept)"]] + coef(obama_ideo_gop)[["GOP"]]` `r coef(obama_ideo_gop)[["RConserv"]] + coef(obama_ideo_gop)[["RConserv:GOP"]]` (\text{Respondent conservatism})
\end{align}
    $$

## Dichotomous interaction

```{r obama-ideo-x-gop-plot}
nes %>%
  add_predictions(obama_ideo_gop) %>%
  ggplot(aes(RConserv, ObamaTherm, color = factor(GOP))) +
  geom_jitter(alpha = .5) +
  geom_line(aes(y = pred)) +
  scale_color_manual(values = c("blue", "red")) +
  labs(x = "Respondent conservatism",
       y = "Obama thermometer score") +
  theme(legend.position = "none")
```

## Separate models

```{r obama-split}
tidy(lm(ObamaTherm ~ RConserv, data = filter(nes, GOP == 0)))
tidy(lm(ObamaTherm ~ RConserv, data = filter(nes, GOP == 1)))
```

## Causal direction

```{r obama-other-direction}
nes %>%
  data_grid(RConserv, GOP) %>%
  add_predictions(obama_ideo_gop) %>%
  spread(GOP, pred) %>%
  mutate(diff = `1` - `0`) %>%
  ggplot(aes(RConserv, diff)) +
  geom_point() +
  labs(title = "Expected Obama thermometer score",
       x = "Respondent conservatism",
       y = "First difference between Republicans\nand non-Republicans")
```

## Calculating standard errors

$$
\begin{align}
\text{Obama thermometer} = \beta_0 &+ (\beta_1 + \beta_3 \text{GOP}) (\text{Respondent conservatism}) \\
& + \beta_2 (\text{GOP respondent}) + e \\
 = &\beta_0 + \psi_1 (\text{Respondent conservatism}) + \beta_2 (\text{GOP respondent}) + e
\end{align}
$$

* Point estimate

    ```{r point}
coef(obama_ideo_gop)[["GOP"]] + coef(obama_ideo_gop)[["RConserv:GOP"]]
    ```

* Standard error

    $$\hat{\sigma}_{\hat{\psi}_1} = \sqrt{\widehat{\text{Var}(\hat{\beta}_1)} + (\text{GOP})^2 \widehat{\text{Var}(\hat{\beta_3})} + 2 (\text{GOP}) \widehat{\text{Cov}(\hat{\beta}_1 \hat{\beta}_3)}}$$

    ```{r std-err}
vcov(obama_ideo_gop)

sqrt(vcov(obama_ideo_gop)["RConserv", "RConserv"] +
       (1)^2 * vcov(obama_ideo_gop)["RConserv:GOP", "RConserv:GOP"] +
       2 * 1 * vcov(obama_ideo_gop)["RConserv", "RConserv:GOP"])
    ```

## Conducting inference {.scrollable}

```{r obama-other-dir-std-err}
# function to get point estimates and standard errors
# model - lm object
# mod_var - name of moderating variable in the interaction
instant_effect <- function(model, mod_var){
  # get interaction term name
  int.name <- names(model$coefficients)[[which(str_detect(names(model$coefficients), ":"))]]
  
  marg_var <- str_split(int.name, ":")[[1]][[which(str_split(int.name, ":")[[1]] != mod_var)]]
  
  # store coefficients and covariance matrix
  beta.hat <- coef(model)
  cov <- vcov(model)
  
  # possible set of values for mod_var
  if(class(model)[[1]] == "lm"){
    z <- seq(min(model$model[[mod_var]]), max(model$model[[mod_var]]))
  } else {
    z <- seq(min(model$data[[mod_var]]), max(model$data[[mod_var]]))
  }
  
  # calculate instantaneous effect
  dy.dx <- beta.hat[[marg_var]] + beta.hat[[int.name]] * z
  
  # calculate standard errors for instantaeous effect
  se.dy.dx <- sqrt(cov[marg_var, marg_var] +
                     z^2 * cov[int.name, int.name] +
                     2 * z * cov[marg_var, int.name])
  
  # combine into data frame
  data_frame(z = z,
             dy.dx = dy.dx,
             se = se.dy.dx)
}

# point range plot
instant_effect(obama_ideo_gop, "RConserv") %>%
  ggplot(aes(z, dy.dx,
             ymin = dy.dx - 1.96 * se,
             ymax = dy.dx + 1.96 * se)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Marginal effect of GOP",
       subtitle = "By respondent conservatism",
       x = "Respondent conservatism",
       y = "Estimated marginal effect")

# line plot
instant_effect(obama_ideo_gop, "RConserv") %>%
  ggplot(aes(z, dy.dx)) +
  geom_line() +
  geom_line(aes(y = dy.dx - 1.96 * se), linetype = 2) +
  geom_line(aes(y = dy.dx + 1.96 * se), linetype = 2) +
  geom_hline(yintercept = 0) +
  labs(title = "Marginal effect of GOP",
       subtitle = "By respondent conservatism",
       x = "Respondent conservatism",
       y = "Estimated marginal effect")
```

## Hypothesis testing {.scrollable}

```{r hypo-test, echo = TRUE}
linearHypothesis(obama_ideo_gop, "RConserv + RConserv:GOP")
```

```{r hypo-test-rconserv7, echo = TRUE}
linearHypothesis(obama_ideo_gop, "GOP + 7 * RConserv:GOP")
```

## Continuous interaction

$$
\begin{align}
\text{Obama thermometer} = \beta_0 &+ \beta_1 (\text{Respondent conservatism}) \\
& + \beta_2 (\text{Obama conservatism})\\
& + \beta_3 (\text{Respondent conservatism}) (\text{Obama conservatism}) \\
& + e
\end{align}
$$

## Continuous interaction

```{r obama-two-cont}
obama_ideo2 <- lm(ObamaTherm ~ RConserv * ObamaConserv, data = nes)
tidy(obama_ideo2)
glance(obama_ideo2)
```

## Continuous interaction {.scrollable}

```{r obama-two-marg-effect}
instant_effect(obama_ideo2, "RConserv") %>%
  ggplot(aes(z, dy.dx,
             ymin = dy.dx - 1.96 * se,
             ymax = dy.dx + 1.96 * se)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Marginal effect of Obama's conservatism",
       subtitle = "By respondent conservatism",
       x = "Respondent conservatism",
       y = "Estimated marginal effect")

instant_effect(obama_ideo2, "ObamaConserv") %>%
  ggplot(aes(z, dy.dx,
             ymin = dy.dx - 1.96 * se,
             ymax = dy.dx + 1.96 * se)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Marginal effect of respondent's conservatism",
       subtitle = "By Obama conservatism",
       x = "Obama conservatism",
       y = "Estimated marginal effect")
```

## Predicted values plots

```{r obama-heatmap}
nes %>%
  data_grid(RConserv, ObamaConserv) %>%
  add_predictions(obama_ideo2) %>%
  ggplot(aes(RConserv, ObamaConserv, z = pred, fill = pred)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_gradient2(midpoint = 50) +
  geom_contour() +
  labs(title = "Expected Obama thermometer score",
       x = "Respondent conservatism",
       y = "Obama conservatism",
       fill = "Prediction")
```

## Predicted values plots

```{r obama-3d}
nes %>%
  data_grid(RConserv, ObamaConserv) %>%
  add_predictions(obama_ideo2)  %>%
  spread(ObamaConserv, pred) %>%
  select(-RConserv) %>%
  as.matrix %>%
  plot_ly(x = ~ 1:7, y = ~ 1:7, z = .) %>%
  add_surface() %>%
  layout(title = "Predicted values for Obama feeling thermometer",
         scene = list(
           xaxis = list(
             title = "Respondent conservatism"
           ),
           yaxis = list(
             title = "Obama conservatism"
           ),
           zaxis = list(
             title = "Predictions"
           )
         )
  )
```

## Interaction terms in GLMs {.scrollable}

$$p(\text{Survival}) = \frac{e^{\beta_0 + \beta_{1}\text{Age} + \beta_{2}\text{Sex}}}{1 + e^{\beta_0 + \beta_{1}\text{Age} + \beta_{2}\text{Sex}}}$$

```{r titanic_data, message = FALSE}
library(titanic)
titanic <- titanic_train %>%
  as_tibble() %>%
  # remove missing values
  na.omit() %>%
  # convert sex to a binary 0/1 variable for the instant_effects function
  mutate(Male = ifelse(Sex == "male", 1, 0))
```

```{r survive_age_woman}
survive_age_woman <- glm(Survived ~ Age + Male, data = titanic,
                         family = binomial)
tidy(survive_age_woman)
glance(survive_age_woman)

titanic %>%
  data_grid(Age, Male) %>%
  augment(survive_age_woman, newdata = .) %>%
  # convert to probabilities
  mutate(.fitted.lwr = .fitted - 1.96 * .se.fit,
         .fitted.upr = .fitted + 1.96 * .se.fit) %>%
  mutate_at(vars(starts_with(".fitted")), logit2prob) %>%
  mutate(Sex = ifelse(Male == 1, "Male", "Female")) %>%
  ggplot(aes(Age, .fitted, ymin = .fitted.lwr, ymax = .fitted.upr,
             color = Sex, fill = Sex)) +
  geom_ribbon(alpha = .2, color = NA) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Probability of surviving the Titanic",
       subtitle = "Age + Sex",
       y = "Predicted probability of survival",
       color = "Sex",
       fill = "Sex")
```

## Interaction terms in GLMs {.scrollable}

$$p(\text{Survival}) = \frac{e^{\beta_0 + \beta_{1}\text{Age} + \beta_{2}\text{Sex} + \beta_3 (\text{Age} \times \text{Sex})}}{1 + e^{\beta_0 + \beta_{1}\text{Age} + \beta_{2}\text{Sex} + \beta_3 (\text{Age} \times \text{Sex})}}$$

```{r survive_age_woman_x}
survive_age_woman_x <- glm(Survived ~ Age * Male, data = titanic,
                         family = binomial)
tidy(survive_age_woman_x)
glance(survive_age_woman_x)

titanic %>%
  data_grid(Age, Male) %>%
  augment(survive_age_woman_x, newdata = .) %>%
  # convert to probabilities
  mutate(.fitted.lwr = .fitted - 1.96 * .se.fit,
         .fitted.upr = .fitted + 1.96 * .se.fit) %>%
  mutate_at(vars(starts_with(".fitted")), logit2prob) %>%
  mutate(Sex = ifelse(Male == 1, "Male", "Female")) %>%
  ggplot(aes(Age, .fitted, ymin = .fitted.lwr, ymax = .fitted.upr,
             color = Sex, fill = Sex)) +
  geom_ribbon(alpha = .2, color = NA) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Probability of surviving the Titanic",
       subtitle = "Age x Sex",
       y = "Predicted probability of survival",
       color = "Sex",
       fill = "Sex")

instant_effect(survive_age_woman_x, "Age") %>%
  ggplot(aes(z, dy.dx)) +
  geom_line() +
  geom_line(aes(y = dy.dx - 1.96 * se), linetype = 2) +
  geom_line(aes(y = dy.dx + 1.96 * se), linetype = 2) +
  geom_hline(yintercept = 0) +
  labs(title = "Marginal effect of sex",
       subtitle = "By age",
       x = "Age",
       y = "Estimated marginal effect")

instant_effect(survive_age_woman_x, "Male") %>%
  mutate(z = factor(z, levels = 0:1, labels = c("Female", "Male"))) %>%
  ggplot(aes(z, dy.dx,
             ymin = dy.dx - 1.96 * se,
             ymax = dy.dx + 1.96 * se)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Marginal effect of age",
       subtitle = "By sex",
       x = NULL,
       y = "Estimated marginal effect")
```

## Nonparametric regression {.scrollable}

$$
\begin{align}
\text{Obama thermometer} = \beta_0 &+ f_1 (\text{Respondent conservatism}) \\
& + f_2 (\text{Obama conservatism}) + e
\end{align}
$$

```{r gam}
library(mgcv)

obama_gam <- gam(ObamaTherm ~ s(RConserv, k = 5) + s(ObamaConserv, k = 5),
                 data = nes)

# 3d plot
nes %>%
  data_grid(RConserv, ObamaConserv) %>%
  add_predictions(obama_gam)  %>%
  spread(ObamaConserv, pred) %>%
  select(-RConserv) %>%
  as.matrix %>%
  plot_ly(x = ~ 1:7, y = ~ 1:7, z = .) %>%
  add_surface() %>%
  layout(title = "Predicted values for Obama feeling thermometer",
         scene = list(
           xaxis = list(
             title = "Respondent conservatism"
           ),
           yaxis = list(
             title = "Obama conservatism"
           ),
           zaxis = list(
             title = "Predictions"
           )
         )
  )
```

## Nonparametric regression {.scrollable}

$$
\begin{align}
\text{Obama thermometer} = \beta_0 &+ f_1 (\text{Respondent conservatism}) \\
& + f_2 (\text{Obama conservatism})\\
& + f_3 (\text{Respondent conservatism} \times \text{Obama conservatism}) + e
\end{align}
$$

```{r gam-interact}
obama_gam_x <- gam(ObamaTherm ~ s(RConserv, ObamaConserv, k = 5),
                 data = nes)
summary(obama_gam_x)

# 3d plot
nes %>%
  data_grid(RConserv, ObamaConserv) %>%
  add_predictions(obama_gam_x)  %>%
  spread(ObamaConserv, pred) %>%
  select(-RConserv) %>%
  as.matrix %>%
  plot_ly(x = ~ 1:7, y = ~ 1:7, z = .) %>%
  add_surface() %>%
  layout(title = "Predicted values for Obama feeling thermometer",
         scene = list(
           xaxis = list(
             title = "Respondent conservatism"
           ),
           yaxis = list(
             title = "Obama conservatism"
           ),
           zaxis = list(
             title = "Predictions"
           )
         )
  )
```

## Decision trees

```{r tree-interact}
library(tree)
library(ggdendro)

# hackish function to get line segment coordinates for ggplot
partition.tree.data <- function (tree, label = "yval", add = FALSE, ordvars, ...) 
{
  ptXlines <- function(x, v, xrange, xcoord = NULL, ycoord = NULL, 
                       tvar, i = 1L) {
    if (v[i] == "<leaf>") {
      y1 <- (xrange[1L] + xrange[3L])/2
      y2 <- (xrange[2L] + xrange[4L])/2
      return(list(xcoord = xcoord, ycoord = c(ycoord, y1, 
                                              y2), i = i))
    }
    if (v[i] == tvar[1L]) {
      xcoord <- c(xcoord, x[i], xrange[2L], x[i], xrange[4L])
      xr <- xrange
      xr[3L] <- x[i]
      ll2 <- Recall(x, v, xr, xcoord, ycoord, tvar, i + 
                      1L)
      xr <- xrange
      xr[1L] <- x[i]
      return(Recall(x, v, xr, ll2$xcoord, ll2$ycoord, tvar, 
                    ll2$i + 1L))
    }
    else if (v[i] == tvar[2L]) {
      xcoord <- c(xcoord, xrange[1L], x[i], xrange[3L], 
                  x[i])
      xr <- xrange
      xr[4L] <- x[i]
      ll2 <- Recall(x, v, xr, xcoord, ycoord, tvar, i + 
                      1L)
      xr <- xrange
      xr[2L] <- x[i]
      return(Recall(x, v, xr, ll2$xcoord, ll2$ycoord, tvar, 
                    ll2$i + 1L))
    }
    else stop("wrong variable numbers in tree.")
  }
  if (inherits(tree, "singlenode")) 
    stop("cannot plot singlenode tree")
  if (!inherits(tree, "tree")) 
    stop("not legitimate tree")
  frame <- tree$frame
  leaves <- frame$var == "<leaf>"
  var <- unique(as.character(frame$var[!leaves]))
  if (length(var) > 2L || length(var) < 1L) 
    stop("tree can only have one or two predictors")
  nlevels <- sapply(attr(tree, "xlevels"), length)
  if (any(nlevels[var] > 0L)) 
    stop("tree can only have continuous predictors")
  x <- rep(NA, length(leaves))
  x[!leaves] <- as.double(substring(frame$splits[!leaves, "cutleft"], 
                                    2L, 100L))
  m <- model.frame(tree)
  if (length(var) == 1L) {
    x <- sort(c(range(m[[var]]), x[!leaves]))
    if (is.null(attr(tree, "ylevels"))) 
      y <- frame$yval[leaves]
    else y <- frame$yprob[, 1L]
    y <- c(y, y[length(y)])
    if (add) {
      # lines(x, y, type = "s", ...)
    }
    else {
      a <- attributes(attr(m, "terms"))
      yvar <- as.character(a$variables[1 + a$response])
      xo <- m[[yvar]]
      if (is.factor(xo)) 
        ylim <- c(0, 1)
      else ylim <- range(xo)
      # plot(x, y, ylab = yvar, xlab = var, type = "s", ylim = ylim,
      #      xaxs = "i", ...)
    }
    data_frame(x = x, y = y)
  }
  else {
    if (!missing(ordvars)) {
      ind <- match(var, ordvars)
      if (any(is.na(ind))) 
        stop("unmatched names in vars")
      var <- ordvars[sort(ind)]
    }
    lab <- frame$yval[leaves]
    if (is.null(frame$yprob)) 
      lab <- format(signif(lab, 3L))
    else if (match(label, attr(tree, "ylevels"), nomatch = 0L)) 
      lab <- format(signif(frame$yprob[leaves, label], 
                           3L))
    rx <- range(m[[var[1L]]])
    rx <- rx + c(-0.025, 0.025) * diff(rx)
    rz <- range(m[[var[2L]]])
    rz <- rz + c(-0.025, 0.025) * diff(rz)
    xrange <- c(rx, rz)[c(1, 3, 2, 4)]
    xcoord <- NULL
    ycoord <- NULL
    xy <- ptXlines(x, frame$var, xrange, xcoord, ycoord, 
                   var)
    xx <- matrix(xy$xcoord, nrow = 4L)
    yy <- matrix(xy$ycoord, nrow = 2L)

    return(list(data_frame(xmin = xx[1L,],
                           ymin = xx[2L,],
                           xmax = xx[3L,],
                           ymax = xx[4L,]),
                data_frame(x = yy[1L,],
                           y = yy[2L,],
                           label = lab)))
    # if (!add) 
    #   plot(rx, rz, xlab = var[1L], ylab = var[2L], type = "n", 
    #        xaxs = "i", yaxs = "i", ...)
    # segments(xx[1L, ], xx[2L, ], xx[3L, ], xx[4L, ])
    # text(yy[1L, ], yy[2L, ], as.character(lab), ...)
  }
}

obama_tree <- tree(ObamaTherm ~ RConserv + GOP, data = nes)

tree_data <- dendro_data(obama_tree)
ptree <- ggplot(segment(tree_data)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), 
               alpha = 0.5) +
  geom_text(data = label(tree_data), 
            aes(x = x, y = y, label = label_full), vjust = -0.5, size = 3) +
  geom_text(data = leaf_label(tree_data), 
            aes(x = x, y = y, label = label), vjust = 0.5, size = 3) +
  theme_dendro()
ptree
```

## Support vector machines {.scrollable}

```{r svm-linear}
library(e1071)

# convert titanic$Survived to factor
titanic <- titanic %>%
  mutate(Survived = as.factor(Survived))

# estimate and plot the linear model
titanic_svm <- svm(Survived ~ Age + Male, data = titanic, kernel = "linear")
plot(titanic_svm, titanic, Age ~ Male)

# estimate and plot the interactive model
titanic_svm_x <- svm(Survived ~ Age + Male, data = titanic, kernel = "radial")
plot(titanic_svm_x, titanic, Age ~ Male)
```

## Neural networks

* Highly non-linear and interactive

