---
title: "Practice tidying data"
date: 2019-03-01

type: docs
toc: true
draft: false
aliases: ["/datawrangle_tidy_exercise.html"]
categories: ["datawrangle"]

menu:
  notes:
    parent: Data wrangling
    weight: 8
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
```

{{% callout note %}}

Run the code below in your console to download this exercise as a set of R scripts.

```r
usethis::use_course("uc-cfss/data-wrangling-tidy-data")
```

{{% /callout %}}

For each exercise, tidy the data frame. Before you write any code examine the structure of the data frame and mentally (or with pen-and-paper) sketch out what you think the tidy data structure should be.

## Race data

```{r race}
library(rcfss)
race
```

Important info:

* `Name` - pretty obvious
* `50`:`350` - column names define different lengths of time
* Cell values are scores associated with each name and length of time

{{< spoiler text="Click for a hint" >}}

**Tidy data structure**

```{r race-hint, echo = FALSE}
pivot_longer(
  data = race,
  cols = -Name,
  names_to = "Time",
  values_to = "Score",
  # ensure the Time column is stored as a numeric column
  names_transform = list(Time = readr::parse_number)
)
```

{{< /spoiler >}}

{{< spoiler text="Click for the solution" >}}

```{r race-hint, echo = TRUE}
```

Except for the `Name` column, the remaining columns are actually one variable spread across multiple columns. The column names are a distinct variable, and the columns' values are another variable. `pivot_longer()` is the appropriate function.

{{% callout note %}}

Because the column names are actually numeric values, we use `names_transform = list(Time = readr::parse_number)` to coerce the new `Time` column into a numeric column. `names_transform` allows us to manually specify the column type for the name column. Because (rarely) you may have more than one name column, we store each type separately inside of `list()`. `readr::parse_number()` is a function for converting a character vector to a numeric vector, so `names_transform = list(Time = readr::parse_number)` ensures the `Time` column is stored as a numeric column.

{{% /callout %}}

{{< /spoiler >}}

## Clinical trials

```{r results}
results
```

Important info:

* `Ind` - individual participating in the experiment
* `Treatment` - trial type (`Treat` or `Cont`)
* `value` - result of experiment

{{< spoiler text="Click for a hint" >}}

**Tidy data structure**

```{r results-hint, echo = FALSE}
pivot_wider(data = results, names_from = Treatment, values_from = value)
```

{{< /spoiler >}}

{{< spoiler text="Click for the solution" >}}

```{r results-hint, echo = TRUE}
```

This dataset is not tidy because observations are spread across multiple rows. There only needs to be one row for each individual. Then `Treat` and `Cont` can be stored in separate columns.

{{< /spoiler >}}

## Grades

```{r grades}
grades
```

This one is a bit tougher. Important info:

* **The unit of analysis is ID-Year-Quarter.** That is, in the tidy formulation each observation represents one individual during one quarter in a given year.
* **Each test is unique.** As in they should be treated as two separate variables.

{{< spoiler text="Click for a hint" >}}

**Tidy data structure**

```{r grades-hint, echo = FALSE}
pivot_longer(
  data = grades,
  cols = c(Fall:Winter),
  names_to = "Quarter",
  values_to = "Score"
) %>%
  pivot_wider(
    names_from = Test,
    values_from = Score
  )
```

{{< /spoiler >}}

{{< spoiler text="Click for the solution" >}}

```{r grades-hint, echo = TRUE}
```

In this example, the basic unit of observation is the test. Each individual takes two separate tests (`Math` or `Writing`) at multiple points in time: during each quarter (`Fall`, `Winter`, `Spring`) as well as in multiple years (`2008` and `2009`). So our final data frame should contain five columns: `ID` (identifying the student), `Year` (year the test was taken), `Quarter` (quarter in which the test was taken), `Math` (score on the math test), and `Writing` (score on the writing test).

Where can we begin? Initially we can make the data frame longer by making `Fall`, `Winter`, and `Spring` into a single column (we can use the inclusive select function `:` to gather these three columns):

```{r grades-longer}
pivot_longer(
  data = grades,
  cols = c(Fall:Winter),
  names_to = "Quarter",
  values_to = "Score"
)
```

Good, but now we have observations spread across multiple rows. Remember that we want each test to be a separate variable. To do that, we can `pivot_wider()` those values across two columns.

```{r grades-wider}
pivot_longer(
  data = grades,
  cols = c(Fall:Winter),
  names_to = "Quarter",
  values_to = "Score"
) %>%
  pivot_wider(
    names_from = Test,
    values_from = Score
  )
```

{{< /spoiler >}}

## Activities

```{r activities}
activities
```

This one is also pretty difficult, but if you think it through conceptually it is doable. The unit of analysis is a single individual (identified by `id`) observed at two different times (`T1` and `T2`) performing different actions (`work`, `play`, `talk`, and `total` - note that `total` is not merely the sum of the first three values). Individuals in this experiment were assigned to either treatment or control (`trt`) and this information should be preserved in the final data frame.

{{< spoiler text="Click for a hint" >}}

**Tidy data structure**

```{r activities-hint, echo = FALSE}
pivot_longer(
  data = activities,
  cols = c(work.T1:talk.T2),
  names_to = "variable",
  values_to = "value"
) %>%
  separate(variable, into = c("location", "time")) %>%
  pivot_wider(names_from = location, values_from = value)
```

{{< /spoiler >}}

{{< spoiler text="Click for the solution" >}}

This is a more complex operation. The basic problem is that we have variables stored in multiple columns (location, with possible values of `work`, `play`, and `talk`). We need to combine these columns into a single column for each variable. But what happens if we just make the data frame longer in this way?

```{r activities-longer}
pivot_longer(
  data = activities,
  cols = c(work.T1:talk.T2),
  names_to = "variable",
  values_to = "value"
)
```

We've created a new problem! Actually, two problems:

1. We have a single observation stored across multiple rows: we want a single row for each `id` x `trt` pairing
2. We have two variables stored in a single column: `variable` contains the information on both location (`work`, `play`, and `talk`) as well as when the measurement was taken (`T1` or `T2`)

The best approach is to fix the second problem by separating the columns, then spreading the different types of measurements back into their own columns.

```{r activities-separate-longer}
pivot_longer(
  data = activities,
  cols = c(work.T1:talk.T2),
  names_to = "variable",
  values_to = "value"
) %>%
  separate(variable, into = c("location", "time"))

pivot_longer(
  data = activities,
  cols = c(work.T1:talk.T2),
  names_to = "variable",
  values_to = "value"
) %>%
  separate(variable, into = c("location", "time")) %>%
  pivot_wider(names_from = location, values_from = value)
```

{{< /spoiler >}}

## Session Info

```{r child = here::here("R", "_session-info.Rmd")}
```
