---
title: "Practice generating layered graphics using ggplot2"
date: 2019-03-01

type: book
toc: true
draft: false
aliases: ["/dataviz_gapminder.html"]
categories: ["dataviz"]

weight: 24
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
options(digits = 3)
```

```{r packages, cache = FALSE, message = FALSE}
library(tidyverse)
```

{{% callout note %}}

Run the code below in your console to download this exercise as a set of R scripts.

```r
usethis::use_course("cis-ds/grammar-of-graphics")
```

{{% /callout %}}

Given your preparation for today's class, now let's practice generating layered graphics in R using data from [Gapminder World](https://www.gapminder.org/data/), which compiles country-level data on quality-of-life measures.

## Load the `gapminder` dataset

If you have not already installed the `gapminder` package and you try to load it using the following code, you will get an error:

```{r gapminder-install, eval = FALSE}
library(gapminder)
```

```
Error in library(gapminder) : there is no package called ‘gapminder’
```

If this happens, install the gapminder package by running `install.packages("gapminder")` in your console.

Once you've done this, run the following code to load the gapminder dataset, the `ggplot2` library, and a helper library for printing the contents of `gapminder`:

```{r gapminder}
library(gapminder)
library(ggplot2)
library(tibble)

glimpse(gapminder)
```

{{% callout note %}}

Run `?gapminder` in the console to open the help file for the data and definitions for each of the columns.

{{% /callout %}}

Using the grammar of graphics and your knowledge of the `ggplot2` library, generate a series of graphs that explore the relationships between specific variables.

## Generate a histogram of life expectancy

{{< spoiler text="Click for the solution" >}}

```{r histo}
ggplot(data = gapminder, mapping = aes(x = lifeExp)) +
  geom_histogram()
```

{{< /spoiler >}}

### Generate separate histograms of life expectancy for each continent

**Hint: think about how to [split your plots to show different subsets of data.](http://r4ds.had.co.nz/data-visualisation.html#facets)**

{{< spoiler text="Click for the solution" >}}

```{r histo-facet}
ggplot(data = gapminder, mapping = aes(x = lifeExp)) +
  geom_histogram() +
  facet_wrap(facets = vars(continent))
```

{{< /spoiler >}}

## Compare the distribution of life expectancy, by continent by generating a boxplot

{{< spoiler text="Click for the solution" >}}

```{r boxplot}
ggplot(data = gapminder, mapping = aes(x = continent, y = lifeExp)) +
  geom_boxplot()
```

{{< /spoiler >}}

### Redraw the plot, but this time use a violin plot

{{< spoiler text="Click for the solution" >}}

```{r violin-plot}
ggplot(data = gapminder, mapping = aes(x = continent, y = lifeExp)) +
  geom_violin()
```

{{< /spoiler >}}

## Generate a scatterplot of the relationship between per capita GDP and life expectancy

{{< spoiler text="Click for the solution" >}}

```{r scatter}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

{{< /spoiler >}}

### Add a smoothing line to the scatterplot

{{< spoiler text="Click for the solution" >}}

```{r scatter-smooth}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth()
```

{{< /spoiler >}}

## Identify whether this relationship differs by continent

### Use the color aesthetic to identify differences

{{< spoiler text="Click for the solution" >}}

```{r scatter-color}
ggplot(
  data = gapminder,
  mapping = aes(x = gdpPercap, y = lifeExp, color = continent)
) +
  geom_point() +
  geom_smooth()
```

{{< /spoiler >}}

### Use faceting to identify differences

{{< spoiler text="Click for the solution" >}}

```{r scatter-facet}
# using facet_wrap()
ggplot(
  data = gapminder,
  mapping = aes(x = gdpPercap, y = lifeExp, color = continent)
) +
  geom_point() +
  geom_smooth() +
  facet_wrap(facets = vars(continent))

# using facet_grid()
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  geom_smooth() +
  facet_grid(cols = vars(continent))
```

Why use `facet_grid()` here instead of `facet_wrap()`? Good question! Let's reframe it and instead ask, what is the difference between `facet_grid()` and `facet_wrap()`?^[Example drawn from [this StackOverflow thread](https://stackoverflow.com/questions/20457905/whats-the-difference-between-facet-wrap-and-facet-grid-in-ggplot2).]

The answer below refers to the case when you have 2 arguments in `facet_grid()` or `facet_wrap()`. `facet_grid(rows = vars(x), cols = vars(y))` will display $y \times x$ plots even if some plots are empty. For example:

```{r facet-grid}
library(palmerpenguins)

ggplot(data = penguins, aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point() +
  facet_grid(rows = vars(species), cols = vars(island))
```

There are 3 distinct `species` and `island` values. This plot displays $3 \times 3 = 9$ plots, even if some are empty (for example, Chinstrap penguins were not observed on Biscoe Island).

`facet_wrap(facets = vars(species, island))` displays only the plots having actual values.

```{r facet-wrap}
ggplot(data = penguins, aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point() +
  facet_wrap(facets = vars(species, island))
```

There are 5 plots displayed now, one for every combination of `species` and `island`. So for this exercise, I would use `facet_wrap()` because we are faceting on a single variable. If we faceted on multiple variables, `facet_grid()` may be more appropriate.

{{< /spoiler >}}

## Bonus: Identify the outlying countries on the right-side of the graph by labeling each observation with the country name

{{< spoiler text="Click for the solution" >}}

```{r text}
ggplot(
  data = gapminder,
  mapping = aes(x = gdpPercap, y = lifeExp, label = country)
) +
  geom_smooth() +
  geom_text()
```

{{< /spoiler >}}

## Session Info

```{r child = here::here("R", "_session-info.Rmd")}
```
