---
date: "2018-09-09T00:00:00-05:00"
draft: false
weight: 10
title: "Why Git?"
toc: false
type: book
aliases: ["/git00.html", "/setup/what-is-git/"]
---

```{r}
#| label = "setup",
#| include = FALSE
knitr::opts_chunk$set(cache = TRUE)
```

In this course (and in your own work), you will be writing lots of programs. Generally the first draft is not the final draft, be it a research paper or a computer script. We want a way to track our changes over time. Perhaps this is to make sure we have a record of what we've already done that doesn't work, so we can avoid doing it again. Or maybe we want to share our code with collaborators who are working on a project with us. How can we do this?

One potential solution is to email copies of files back and forth as we make changes. But if we do this, we risk having lots of versions of files floating around. How do we know which is the most recent? What happens if someone edits a file and forgets to email it to us? How will we make sure all the changes are merged into the final version?

Or perhaps instead we can do all of our work on a cloud-based storage solution such as [Dropbox](https://www.dropbox.com) or [Google Drive](https://drive.google.com). This ensures changes are synchronized between computers. But we won't know who made what changes to a file. And what happens if two people want to work on the same file at the same time? One person will have to wait for the other to finish before they can edit that file. Plus how will we track previous versions of the file? Dropbox and other cloud storage services have some [version control solutions](https://www.dropbox.com/en/help/113), but these are not comprehensive or only store versions for a limited time. Plus every time we save a new version of the program, the entire file has to be retained. On large projects, this will eat up storage space quickly.

We want a solution that:

* Keeps old versions of files indefinitely. Since all these versions are stored, we can always go back and see who modified the file and what changes they made. Or if we make a mistake in the future that breaks the program, we can revert back to an older version to fix it.
* Since we know who modified each file, if we have questions in the future we can go to that person with our questions.
* When collaborating with multiple people on the same project, we want any changes made by project members to be integrated to the most recent version. If I try to modify a file and don't incorporate another member's revisions, I need to be told there is a conflict so I can merge all the changes.

# Git

Git is a *version control system* originally created for developers to collaborate on large software projects. Git tracks changes in the project over time so that there is always a comprehensive, structured record of the project. Each project is stored in a *repository* that includes all files that are part of the project. As social scientists, this is more than just computer scripts - this can include data files and reports, as well as our source code.

Git can be used locally by you on a single computer to track changes in a project. You do not need to be connected to the internet to use Git. However if you want to share your work with a larger audience, the easiest solution is to host the repository on a web site for others to download and inspect. You can host a public (open to the world) or private (open to just you or a few individuals) repository. [GitHub](https://www.github.com) has become the largest hoster of Git repositories and includes many useful features beyond the standard Git functions.

Chances are that by now you've seen or even used GitHub. Professionally, you should know how to use Git and GitHub to manage projects and share code. In this class, we will use Git and GitHub to host our [course site](https://github.com/cis-ds/course-site), share code, and distribute/collect assignments.

## Acknowledgments

```{r}
#| child = here::here("R", "_ack_stat545.Rmd")
```
```{r}
#| child = here::here("R", "_ack_swcgit.Rmd")
```
