---
title: "Exploratory data analysis"
author: "MACS 30500 <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: r
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
# generate CSS file
library(xaringanthemer)
rcfss::xaringan_theme()

source(here::here("R", "slide-opts.R"))

# enable panelsets and default theme
xaringanExtra::use_panelset()
```

```{r pkgs, include = FALSE, cache = FALSE}
library(tidyverse)
library(palmerpenguins)
library(rcfss)
library(knitr)
library(patchwork)
library(flipbookr)

set.seed(1234)
theme_set(theme_gray(base_size = 16))
```

# Exploratory data analysis

1. Generate questions about your data
1. Search for answers by visualizing, transforming, and modeling your data
1. Use what you learn to refine your questions and or generate new questions
1. Rinse and repeat until you publish a paper

--

* Variation
* Covariation

---

```{r penguins-eda, warning = FALSE, include = FALSE}
ggplot(
  data = penguins,
  mapping = aes(
    x = body_mass_g,
    y = flipper_length_mm
  )
) +
  geom_point() +
  geom_smooth()
```

`r chunk_reveal(chunk_name = "penguins-eda", break_type = "auto", title = "## Exploratory of Confirmatory Data Analysis")`

---

```{r penguins-final, include = FALSE}
ggplot(
  data = penguins,
  mapping = aes(
    x = body_mass_g,
    y = flipper_length_mm
  )
) +
  geom_point(alpha = .1) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Relationship between body mass and\nflipper length of a penguin",
    subtitle = "Sample of 344 penguins",
    x = "Body mass(g)",
    y = "Flipper length(mm)"
  ) +
  theme_xaringan(
    title_font_size = 18,
    text_font_size = 16
  )
```

`r chunk_reveal(chunk_name = "penguins-final", break_type = "auto", title = "## Characteristics of Confirmatory Data Analysis")`

---

class: inverse, middle

# `scorecard`

---

## `scorecard`

```{r epa-data}
glimpse(scorecard)
```

---

class: inverse, middle

# Univariate graphs

---

```{r histogram, include = FALSE}
ggplot(
  data = scorecard,
  mapping = aes(x = cost)
) +
  geom_histogram()
```

`r chunk_reveal(chunk_name = "histogram", title = "## Histogram")`

---

```{r histogram-bins, include = FALSE}
ggplot(
  data = scorecard,
  mapping = aes(x = cost)
) +
  geom_histogram(bins = 50) + #ROTATE
  geom_histogram(bins = 30) + #ROTATE
  geom_histogram(bins = 10) #ROTATE
```

`r chunk_reveal(chunk_name = "histogram-bins", break_type = "rotate", title = "## Histogram")`

---

```{r barplot, include = FALSE}
ggplot(
  data = scorecard,
  mapping = aes(x = type)
) +
  geom_bar()
```

`r chunk_reveal(chunk_name = "barplot", title = "## Bar chart")`

---

## Covariation

1. Two-dimensional graphs
1. Multiple window plots
1. Utilizing additional channels

---

```{r boxplot, include = FALSE}
ggplot(
  data = scorecard,
  mapping = aes(
    x = type,
    y = cost
  )
) +
  geom_boxplot()
```

`r chunk_reveal(chunk_name = "boxplot", title = "## Box plot")`

---

```{r scatterplot, include = FALSE}
ggplot(
  data = scorecard,
  mapping = aes(
    x = cost,
    y = netcost
  )
) +
  geom_point()
```

`r chunk_reveal(chunk_name = "scatterplot", title = "## Scatterplot")`

---

```{r histogram-facet, include = FALSE}
ggplot(
  data = scorecard,
  mapping = aes(x = cost)
) +
  geom_histogram() + #BREAK
  facet_wrap(facets = vars(type)) #BREAK
```

`r chunk_reveal(chunk_name = "histogram-facet", break_type = "user", title = "## Multiple windows plot")`

---

```{r scatterplot-facet, include = FALSE}
ggplot(
  data = scorecard,
  mapping = aes(
    x = cost,
    y = netcost
  )
) +
  geom_point() + #BREAK
  facet_wrap(facets = vars(type)) #BREAK
```

`r chunk_reveal(chunk_name = "scatterplot-facet", break_type = "user", title = "## Multiple windows plot")`

---

```{r scatterplot-mult-channels, include = FALSE}
ggplot(
  data = scorecard,
  mapping = aes(
    x = cost,
    y = netcost,
    color = type, #ROTATE
    color = type, size = debt #ROTATE
  )
) +
  geom_point()
```

`r chunk_reveal(chunk_name = "scatterplot-mult-channels", break_type = "rotate", font_size_code = "60%", title = "## Utilizing additional channels")`
