---
title: "Exploratory data analysis"
author: "[MACS 30500](https://cfss.uchicago.edu) <br /> University of Chicago"
output: rcfss::xaringan
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE,
                      fig.retina = 2, fig.width = 12)

library(tidyverse)
library(palmerpenguins)
library(rcfss)
library(knitr)
library(patchwork)

set.seed(1234)
theme_set(theme_gray(base_size = rcfss::base_size))
```

# Exploratory data analysis

1. Generate questions about your data
1. Search for answers by visualizing, transforming, and modeling your data
1. Use what you learn to refine your questions and or generate new questions
1. Rinse and repeat until you publish a paper

--

* Variation
* Covariation

---

# Characteristics of EDA

```{r penguins-eda, warning = FALSE, fig.height = 6}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point() +
  geom_smooth()
```

---

# Characteristics of CDA

```{r penguins-final, fig.height = 4, warning = FALSE}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(alpha = .1) +
  geom_smooth(se = FALSE) +
  labs(title = "Relationship between body mass and flipper length of a penguin",
       subtitle = "Sample of 344 penguins",
       x = "Body mass(g)",
       y = "Flipper length(mm)") +
  theme_minimal(base_size = 18)
```

---

# `scorecard`

```{r epa-data}
glimpse(scorecard)
```

---

# Histogram

```{r histogram, fig.height = 6}
ggplot(data = scorecard,
       mapping = aes(x = cost)) +
  geom_histogram()
```

---

# Binning histograms

.pull-left[

```{r histogram-bins-50, fig.width = 6, fig.height = 6}
ggplot(data = scorecard,
       mapping = aes(x = cost)) +
  geom_histogram(bins = 50)

```

]

.pull-right[

```{r histogram-bins-10, fig.width = 6, fig.height = 6}
ggplot(data = scorecard,
       mapping = aes(x = cost)) +
  geom_histogram(bins = 10)
```

]

---

# Bar chart

```{r barplot, fig.height = 6}
ggplot(data = scorecard,
       mapping = aes(x = type)) +
  geom_bar()
```

---

# Covariation

1. Two-dimensional graphs
1. Multiple window plots
1. Utilizing additional channels

---

# Box plot

```{r boxplot, fig.height = 6}
ggplot(data = scorecard,
       mapping = aes(x = type, y = cost)) +
  geom_boxplot()
```

---

# Scatterplot

```{r scatterplot, fig.height = 6}
ggplot(data = scorecard,
       mapping = aes(x = cost, y = netcost)) +
  geom_point()
```

---

# Multiple window plots

```{r histogram-facet, fig.height = 6}
ggplot(data = scorecard,
       mapping = aes(x = cost)) +
  geom_histogram() +
  facet_wrap(~ type)
```

---

# Multiple window plots

```{r scatterplot-facet, fig.height = 5.5}
ggplot(data = scorecard,
       mapping = aes(x = cost, y = netcost)) +
  geom_point() +
  facet_wrap(~ type)
```

---

# Utilizing additional channels

```{r scatterplot-color, fig.height = 6}
ggplot(data = scorecard,
       mapping = aes(x = cost,
                     y = netcost,
                     color = type)) +
  geom_point()
```

---

# Utilizing additional channels

```{r scatterplot-color-size, fig.height = 5.5}
ggplot(data = scorecard,
       mapping = aes(x = cost,
                     y = netcost,
                     color = type,
                     size = debt)) +
  geom_point()
```
