---
title: "Exploratory data analysis"
author: "[MACS 30500](https://cfss.uchicago.edu) <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      beforeInit: "../macros.js"
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, message = FALSE,
                      fig.retina = 2, fig.width = 12)

library(tidyverse)
library(nycflights13)
library(rcfss)
library(knitr)
library(patchwork)

set.seed(1234)
theme_set(theme_gray(base_size = 20))
```

# Exploratory data analysis

1. Generate questions about your data
1. Search for answers by visualising, transforming, and modeling your data
1. Use what you learn to refine your questions and or generate new questions
1. Rinse and repeat until you publish a paper

--

* Variation
* Covariation

---

# Characteristics of EDA

```{r diamonds-eda, fig.height = 6}
ggplot(diamonds, aes(carat, price)) +
  geom_point() +
  geom_smooth()
```

---

# Characteristics of CDA

```{r diamonds-final, fig.height = 4}
ggplot(diamonds, aes(carat, price)) +
  geom_point(alpha = .01) +
  geom_smooth(se = FALSE) +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "Exponential relationship between carat size and price",
       subtitle = "Sample of 54,000 diamonds",
       x = "Carat size",
       y = "Price") +
  theme_minimal()
```

---

# `mpg`

```{r epa-data}
mpg
```

---

# Histogram

```{r histogram}
ggplot(mpg, aes(hwy)) +
  geom_histogram()
```

---

# `geom_rug()`

```{r rug}
ggplot(mpg, aes(hwy)) +
  geom_histogram() +
  geom_rug()
```

---

# Binning histograms

.pull-left[

```{r histogram-bins-50, fig.width = 6}
ggplot(mpg, aes(hwy)) +
  geom_histogram(bins = 50) +
  geom_rug()
```

]

.pull-right[

```{r histogram-bins-10, fig.width = 6}
ggplot(mpg, aes(hwy)) +
  geom_histogram(bins = 10) +
  geom_rug()
```

]



---

# Bar chart

```{r barplot}
ggplot(mpg, aes(class)) +
  geom_bar()
```

---

# Covariation

1. Two-dimensional graphs
1. Multiple window plots
1. Utilizing additional channels

---

# Box plot

```{r boxplot}
ggplot(mpg, aes(class, hwy)) +
  geom_boxplot()
```

---

# Scatterplot

```{r scatterplot}
ggplot(mpg, aes(displ, hwy)) +
  geom_point()
```

---

# Multiple window plots

```{r histogram-facet}
ggplot(mpg, aes(hwy)) +
  geom_histogram() +
  facet_wrap(~ drv)
```

---

# Multiple window plots

```{r scatterplot-facet}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~ drv)
```

---

# Utilizing additional channels

```{r scatterplot-color}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point()
```

---

# Utilizing additional channels

```{r scatterplot-color-size}
ggplot(mpg, aes(displ, hwy, color = class, size = cyl)) +
  geom_point()
```

---

# Utilizing additional channels

```{r scatterplot-shape, fig.height = 4}
ggplot(mpg, aes(displ, hwy, shape = class)) +
  geom_point()
```
