---
title: "Data wrangling"
author: |
  | MACS 30500
  | University of Chicago
date: "January 18, 2017"
output: rcfss::cfss_slides
---

---
title: "Data wrangling"
author: |
  | MACS 30500
  | University of Chicago
date: "October 5, 2016"
output: rcfss::cfss_slides
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE,
                      warning = FALSE)

set.seed(11091987)

library(tidyverse)
library(rcfss)
library(knitr)
theme_set(theme_gray(base_size = 24))
```

## Vectors

* Basic unit of data storage in R
* Types of vectors
    * Logical
    * Numeric
    * Character
    * Factors
* Certain operations can be performed only on certain types of vectors

## Vectors can only be of one type {.scrollable}

```{r}
x <- c(1, 2, "abc", TRUE)
x
parse_logical(x)
parse_integer(x)
parse_character(x)
```

## Parsing vectors

```{r}
challenge <- read_csv(readr_example("challenge.csv"))
```

## Parsing vectors {.scrollable}

```{r}
challenge_parsed <- read_csv(
  readr_example("challenge.csv"), 
  col_types = cols(
    x = col_double(),
    y = col_date()
  )
)
tail(challenge_parsed)
```

## Tidy data

![Figure 12.1 from [@hadley2016]](http://r4ds.had.co.nz/images/tidy-1.png)

## Common tidying tasks

* Gathering
* Spreading
* Separating
* Uniting

## Gathering

```{r}
table4a
```

## Gathering

```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```

## Spreading

```{r}
table2
```

## Spreading

```{r}
table2 %>%
  spread(key = type, value = count)
```

## Separating

```{r}
table3
```

## Separating

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```

## Uniting

```{r}
table5
```

## Uniting {.scrollable}

```{r}
table5 %>% 
  unite(new, century, year)

# remove underscore
table5 %>% 
  unite(new, century, year, sep = "")
```

## Practice tidying

Perform exercises in [notes](block006_tidy_data.html#practice_tidying_data)

## Introduction to relational data

* Multiple tables of data that when combined together answer research questions
* Relations define the important element, not just the individual tables
* Relations are defined between a pair of tables
* Relational verbs
    * Mutating joins
    * Filtering joins

## Comics {.scrollable}

```{r, echo = FALSE}
library(rcfss)
superheroes_kable <- knitr::kable(superheroes)
publishers_kable <- knitr::kable(publishers)
```

```{r}
superheroes_kable
publishers_kable
```

## Inner join

```{r}
(ijsp <- inner_join(superheroes, publishers))
```
  
## Semi join

```{r}
(sjsp <- semi_join(superheroes, publishers))
```

## Left join

```{r}
(ljsp <- left_join(superheroes, publishers))
```

## Anti join

```{r}
(ajsp <- anti_join(superheroes, publishers))
```

## Full join

```{r}
(fjsp <- full_join(superheroes, publishers))
```




