---
title: "Data wrangling: relational data and factors"
author: |
  | MACS 30500
  | University of Chicago
date: "October 11, 2017"
output: rcfss::cfss_slides
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE,
                      warning = FALSE)

set.seed(1234)

library(tidyverse)
theme_set(theme_minimal(base_size = 24))
```

## Introduction to relational data

* Multiple tables of data that when combined together answer research questions
* Relations define the important element, not just the individual tables
* Relations are defined between a pair of tables
* Relational verbs
    * Mutating joins
    * Filtering joins

----

```{r kable, include = FALSE}
library(rcfss)
superheroes_kable <- knitr::kable(superheroes)
publishers_kable <- knitr::kable(publishers)
```

### Superheroes

```{r superheroes, echo = FALSE}
superheroes_kable
```

### Publishers

```{r publishers, echo = FALSE}
publishers_kable
```

## Inner join

```{r ijsp}
(ijsp <- inner_join(superheroes, publishers))
```
  
## Semi join

```{r sjsp}
(sjsp <- semi_join(superheroes, publishers))
```

## Left join

```{r ljsp}
(ljsp <- left_join(superheroes, publishers))
```

## Right join

```{r rjsp}
(rjsp <- right_join(superheroes, publishers))
```

## Anti join

```{r ajsp}
(ajsp <- anti_join(superheroes, publishers))
```

## Full join

```{r fjsp}
(fjsp <- full_join(superheroes, publishers))
```
