---
title: "Pipes and functions in R"
author: |
  | MACS 30500
  | University of Chicago
date: "January 23, 2017"
output: rcfss::cfss_slides
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE,
                      warning = FALSE)

set.seed(1234)

library(tidyverse)
```

----

> Using the `diamonds` dataset, calculate the average price for each cut of "I" colored diamonds.

1. Filter `diamonds` to only keep observations where the color is rated as "I"
1. Group the filtered `diamonds` data frame by cut
1. Summarize the grouped and filtered `diamonds` data frame by calculating the average carat size and price

## Intermediate steps

```{r intermediate}
diamonds_1 <- filter(diamonds, color == "I")
diamonds_2 <- group_by(diamonds_1, cut)
(diamonds_3 <- summarize(diamonds_2, price = mean(price)))
```

## Overwrite the original

```{r overwrite}
diamonds_t <- diamonds    # copy diamonds to diamonds_t just for demonstration purposes
diamonds_t <- filter(diamonds_t, color == "I")
diamonds_t <- group_by(diamonds_t, cut)
(diamonds_t <- summarize(diamonds_t, price = mean(price)))
```

## Function composition

```{r function_comp}
summarize(
  group_by(
    filter(diamonds, color == "I"),
    cut
  ),
  price = mean(price)
)
```

## Piping

```{r pipe}
diamonds %>%
  filter(color == "I") %>%
  group_by(cut) %>%
  summarize(price = mean(price))
```

## Exercise on piping

Complete [this exercise](http://cfss.uchicago.edu/block008_pipes.html#practice_piping_alternatives)

## Functions

* Easy to reuse
* Self-documenting
* Easy-ier to debug
* [If you have copied and pasted a block of code more than twice, convert it to a function.](http://r4ds.had.co.nz/functions.html#when-should-you-write-a-function)

## Function components

* Name
* Arguments
* Body

## Rescale function

```{r, collapse = TRUE}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

rescale01(c(0, 5, 10))
rescale01(c(-10, 0, 10))
rescale01(c(1, 2, 3, NA, 5))
```

* Name
* Arguments
* Body

## What is that?

```{r}
pythagorean <- function(a, b){
  hypotenuse <- sqrt(a^2 + b^2)
  return(hypotenuse)
}
```

* Name
* Arguments
* Body

## How to use a function {.scrollable}

```{r, error = TRUE}
# print the output of the function
pythagorean(a = 3, b = 4)

# save the output as a new object
(tri_c <- pythagorean(a = 3, b = 4))

# what happens to the hypotenuse from inside the function?
pythagorean(a = 3, b = 4)
hypotenuse
```

## Exercise

Stop and go [write a function](http://cfss.uchicago.edu/block009_functions.html#exercise:_calculate_the_sum_of_squares_of_two_variables)

## Conditional execution

```{r eval = FALSE}
if (condition) {
  # code executed when condition is TRUE
} else {
  # code executed when condition is FALSE
}
```

## Conditional execution

```{r eval = FALSE}
if (this) {
  # do that
} else if (that) {
  # do something else
} else {
  # do something completely different
}
```

## Conditional execution and `cut()`

```{r}
diamonds %>%
  select(carat) %>%
  mutate(carat_autobin = cut(carat, breaks = 5),
         carat_manbin = cut(carat,
                            breaks = c(0, 1, 2, 3, 6),
                            labels = c("Small", "Medium",
                                       "Large", "Huge")))
```

## `if()` versus `ifelse()`

```{r}
library(rcfss)
data("gun_deaths")

(educ <- select(gun_deaths, education))
```

## `if()` versus `ifelse()`

```{r, error = TRUE, collapse = TRUE}
(educ_if <- educ %>%
  mutate(hsPlus = if(education == "Less than HS"){
    "Less than HS"
  } else{
    "HS+"
  }))

count(educ_if, hsPlus)
```

## `if()` versus `ifelse()`

```{r, collapse = TRUE}
(educ_ifelse <- educ %>%
  mutate(hsPlus = ifelse(education == "Less than HS", "Less than HS", "HS+")))

count(educ_ifelse, hsPlus)
```

## Exercise

Go write a [fizzbuzz function](http://cfss.uchicago.edu/block009_functions.html#exercise:_write_a_fizzbuzz_function)




